### 2024-11-20 paper/paper.qmd
update paper with data section

---
title: "Examining the Overturn of Roe v. Wade: The Impact of Abortion Bans on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.268 deaths per 1,000 people in states that became abortion-banned, compared to states where abortion remained legal based on Difference in Difference linear modeling."
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(knitr)
```

```{r}
analysis_data<-read_parquet(here::here("data", "02-analysis_data", "analysis_data.parquet"))
```

# Introduction

The overturning of Roe v. Wade on June 24, 2022, in the Supreme Court’s Dobbs v. Jackson Women’s Health Organization decision ended nearly five decades of constitutional protection for abortion rights. This ruling returned authority over abortion regulations to individual states, resulting in a patchwork of laws across the country. The shift sparked debates over reproductive rights and public health, underscoring disparities in access to abortion services and their impact on health outcomes in states with restrictive policies.

The banning of abortion not only raises maternal mortality rates—evidenced by a 56% increase in Texas from 2019 to 2022 [@nbcnews]—but also influences infant mortality rates. Restrictive abortion laws can lead to more unintended or high-risk pregnancies, particularly among disadvantaged individuals with limited healthcare access, resulting in higher rates of preterm births, low birth weights, and delivery complications that elevate infant mortality.

Before

The estimand of our analysis is the change in the difference in infant death rates between states that banned abortion and states where abortion remained legal after June 24, 2022. The object of the estimation is to estimate the change in the difference in the infant death rate between the treatment and control groups based on observational data in the United States using the Difference-in-Difference approach. Our aim is to identify changes in infant mortality trends over time, providing a clearer understanding of the potential impacts of abortion bans on public health outcomes.

The structure of this paper is organized as follows: @sec-data 
provides details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions
and specifications of our linear regression framework. In
@sec-result, we present our findings, emphasizing the key predictors
of Harris's support. Finally, @sec-discussion explores the
implications of our results and suggests potential directions for future
research. @sec-a provides external data detail, @sec-b provides
model detail, @sec-c provides an investigation for pollster Siena College's
methodology, and  @sec-d provides an idealized methodology for polling the poll.

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We conduct our infant death rate observational data analysis using the R programming language
[@citeR]. Our dataset, obtained from CDC Wonder
[@cdc_wonder_2024], based on data as of December 2022, provides
a detailed overview of public opinion in the lead-up to the election.
Adhering to the guidelines presented in @tellingstories, we explore
various factors that influence voter support percentages, including the
timing of the polls, the traits of polling organizations, and regional
differences.

Overview text

## Measurement

## Outcome variables

### Death Rate

The `death rate` refers to the number of deaths of infants under one year of age per 1,000 live births in a specified population during a given time period. It serves as a key indicator of a population’s overall health and access to healthcare. Rates labeled as "Unreliable" occur when there are fewer than 20 deaths in the numerator, as these figures do not meet the National Center for Health Statistics (NCHS) standards for reliability or precision. After cleaning the dataset, the "Unreliable" label is converted into a separate variable named `reliable`.

```{r}
#| label: fig-deathrate
#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants.The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
#| echo: false

ggplot(analysis_data, aes(x = death_rate)) +
  geom_histogram(binwidth = 1, fill = "brown", color = "black") +
  labs(
    x = "Infant Death Rate (per 1,000 live births)", y = "Count",
    caption = "Data source: CDC Wonder."
  ) +theme_classic()
```
@fig-deathrate illustrates the distribution of infant death rates (per 1,000 live births) based on the available data. The majority of infant death rates cluster around lower values, with a peak observed between 5 and 10 deaths per 1,000 infants. As the death rate increases, the frequency of occurrences decreases, indicating that higher infant death rates are relatively rare. There are few instances where the rate exceeds 20 deaths per 1,000 infants, suggesting these are outliers. Overall, the distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.

## Predictor variables

### Abortion Legality & After Injunction

Abortion legality (named `abortion_illegal` in the dataset) is a newly created variable in our dataset, which refers to the legal status of abortion within a given state. It is represented as a binary variable, where 1 indicates that abortion is illegal in the state, and 0 indicates that abortion is legal. This classification reflects the policies and laws governing abortion access in each state.

Another created variable `after_injunction` indicates whether the date of infants' death falls after the implementation of the abortion restrictions following the overturn of Roe v. Wade in June 2022. It is coded as a binary variable, where 1 represents dates after June 2022, signifying the post-injunction period, and 0 represents dates before June 2022, indicating the pre-injunction period.

```{r}
#|include: false
library(usmap)
library(ggplot2)
library(dplyr)

# Ensure the number of states and `abortion_illegal` values match
illegal_states <- tolower(c("Alabama", "Arkansas", "Idaho", "Indiana", "Kentucky", 
                    "Louisiana", "Mississippi", "Oklahoma", "South Dakota", 
                    "Tennessee", "Texas", "West Virginia"))
abortion_status <- data.frame(
  state = tolower(c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", 
    "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", 
    "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", 
    "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", 
    "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", 
    "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", 
    "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", 
    "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"
  )),
  short = c(
    "AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", 
    "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", 
    "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
    "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"
  )
)|>mutate(abortion_illegal = as.factor(case_when(state %in% illegal_states ~ 1, TRUE ~ 0)))
```

```{r}
#|include: false
us_map <- map_data("state")
map_data <- us_map|>
   left_join(abortion_status, c("region"="state"))|>filter(!is.na(abortion_illegal))
```

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State. As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group)) +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  ) +
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) 
```
@fig-map illustrates the legality of abortion across U.S. states. Based on @reproductiveright, as of November 2024, abortion has been illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin following the June 2022 decision.
```{r}
#| label: fig-rateandstate
#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
#| echo: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() 
```
@fig-rateandstate illustrates trends in infant death rates (per 1,000 live births) over time in states where abortion is illegal versus legal. In states with abortion bans, infant mortality shows a slight increase after June 2022, marked by the dashed line representing the overturn of Roe v. Wade. In contrast, states where abortion remains legal exhibit relatively stable trends. This suggests a potential link between abortion bans and rising infant mortality rates, warranting further investigation.

### Year of Death

### State

The state of the mother’s legal residence at the time of birth. This data is categorized by the mother’s state of residence and excludes territories.

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>
  group_by(state)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`State`=state)

# Display the result as a table
kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
```

### Age of Mother

The Age of Mother (`age_of_mother`) refers to the age group of the mother at the time of childbirth.In our observational data, age is categorized into the brackets 15–19 years, 20–24 years, 25–29 years, and 30–34 years in the raw data, in compliance with privacy requirements.

@tbl-motherage displays the average infant death rate (per 1,000 live births) by the age group of mothers. It shows a clear trend: younger mothers, particularly those aged 15–19 and 20–24, have higher infant death rates (9.0 and 10.0, respectively), while older mothers in the 25–29 and 30–34 age groups have lower rates (7.4 and 6.6, respectively). This suggests that maternal age may be a factor influencing infant mortality, with younger mothers experiencing higher risks.
```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

# Display the result as a table
kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
```

### Mother's Single Race
The self-identified race of the mother, recorded as a single race category. In this dataset, the available categories are: Asian, Black or African American, and White. This classification facilitates the analysis of maternal and infant health outcomes across these racial groups.
```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

# Display the result as a table
kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
```

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
second_model <-
  readRDS(file = here::here("models/second_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::msummary(
  list(first_model,second_model),
  fmt = 3,coef_omit="state",notes = "Note. All models include a control for state and pollsters. The reference level of the state is Alaska, the reference level of population is likely voter, and the reference level of pollsters is Angus Reid.")
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

```{r}
analysis_data|>
  ggplot(aes(x = after_injunction, y = death_rate, color = as.factor(abortion_illegal))) +
  geom_point(alpha = 0.2) +
  geom_line(aes(group = state)) +
  theme_minimal() +
  labs(x = "Time period", y = "Death Rate", color = "Person got a new racket")  +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

\newpage

# References

### 2024-11-20 scripts/03-clean_data.R
update data clean

@@ -9,36 +9,46 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
+library(lubridate)
+library(arrow)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+# Specify the input file path
+input_file <- "data/01-raw_data/Linked_Birth_Infant_Death_Records_2017-2022_Expanded.txt"
+
+# Specify the output CSV file path
+output_file <- "data/02-analysis_data/cleaned_infant_death_record.csv"
+
+# Read the tab-separated file, allowing for irregular lines
+data <- read.delim(input_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE, fill = TRUE)
+
+# Write the data to a CSV file
+write.csv(data, output_file, row.names = TRUE)
+
+raw_data <- data
 
 cleaned_data <-
   raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
+  janitor::clean_names() 
+cleaned_data <-cleaned_data|> select(state,age_of_mother_9_code,mother_s_single_race_6,year_of_death,month_code,month,death_rate)|>
   tidyr::drop_na()
 
+# List of states where abortion is illegal after June 2022
+illegal_states <- c("Alabama", "Arkansas", "Idaho", "Indiana", "Kentucky", 
[…]

### 2024-11-20 paper/paper.qmd
update reference and introduction of paper

@@
 ---
-title: "Examining the Overturn of Roe v. Wade: The Impact of Abortion Bans on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
-subtitle: "The infant death rate increased by 0.268 deaths per 1,000 people in states that became abortion-banned, compared to states where abortion remained legal based on Difference in Difference linear modeling."
+title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
+subtitle: "The infant death rate increased by 0.31 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
 author: 
-  - First author
-  - Another author
+  - Diana Shen
 thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
@@
 ```{r}
+#| include: false
+#| warning: false
+#| message: false
 analysis_data<-read_parquet(here::here("data", "02-analysis_data", "analysis_data.parquet"))
@@
 The overturning of Roe v.
-Wade on June 24, 2022, in the Supreme Court’s Dobbs v.
-Jackson Women’s Health Organization decision ended nearly five decades of constitutional protection for abortion rights.
-This ruling returned authority over abortion regulations to individual states, resulting in a patchwork of laws across the country.
-The shift sparked debates over reproductive rights and public health, underscoring disparities in access to abortion services and their impact on health outcomes in states with restrictive policies.
+Wade on June 24, 2022, through the Supreme Court’s decision in Dobbs v.
+Jackson Women’s Health Organization, ended nearly five decades of constitutional protection for abortion rights.
+This landmark ruling shifted authority over abortion regulations to individual states, creating a patchwork of laws nationwide.
+The decision reignited debates about reproductive rights and public health, highlighting disparities in access to abortion services and their effects on health outcomes, especially in states with restrictive policies.
 
-The banning of abortion not only raises maternal mortality rates—evidenced by a 56% increase in Texas from 2019 to 2022 [@nbcnews]—but also influences infant mortality rates.
-Restrictive abortion laws can lead to more unintended or high-risk pregnancies, particularly among disadvantaged individuals with limited healthcare access, resulting in higher rates of preterm births, low birth weights, and delivery complications that elevate infant mortality.
-
-Before
+Banning abortion usually has been linked to a rise in maternal mortality rates, with Texas experiencing a 56% increase from 2019 to 2022 [@nbcnews].
+However, restrictive abortion laws can also impact infant mortality, as they often lead to more unintended or high-risk pregnancies among disadvantaged populations with limited access to healthcare.
+These pregnancies are associated with higher rates of preterm births, low birth weights, and delivery complications, all of which elevate infant mortality rates.
+Even before Roe v.
+Wade was overturned, researchers highlighted the impacts of restrictive abortion laws on mortality rates.
+@Burdick2024 found that states with 11–12 restrictive laws during 2014–2018 had a 16% higher infant mortality rate, particularly in counties with higher proportions of Black populations, inadequate prenatal care, and maternal smoking.
+Similarly, @Harper2023 showed that restrictive laws are linked to higher maternal, fetal, and infant mortality, states with moderate abortion legislation had significantly lower maternal mortality (25.79 per 100,000 live births) and infant mortality (20.56 per 1,000 live births) compared to restrictive states, underscoring the health benefits of less restrictive policies.
 
 The estimand of our analysis is the change in the difference in infant death rates between states that banned abortion and states where abortion remained legal after June 24, 2022.
-The object of the estimation is to estimate the change in the difference in the infant death rate between the treatment and control groups based on observational data in the United States using the Difference-in-Difference approach.
-Our aim is to identify changes in infant mortality trends over time, providing a clearer understanding of the potential impacts of abortion bans on public health outcomes.
+The object of the estimation is to estimate the change in the difference in the infant death rate between the treatment and control groups based on observational data in the United States.
 
-The structure of this paper is organized as follows: @sec-data 
-provides details on the data sources and variables used in our analysis.
-@sec-model explains the modeling approach, including the assumptions
-and specifications of our linear regression framework. In
-@sec-result, we present our findings, emphasizing the key predictors
-of Harris's support. Finally, @sec-discussion explores the
-implications of our results and suggests potential directions for future
-research. @sec-a provides external data detail, @sec-b provides
-model detail, @sec-c provides an investigation for pollster Siena College's
-methodology, and  @sec-d provides an idealized methodology for polling the poll.
+Instead of examining the impact of abortion laws over broader time periods like prior research, our study focuses on the effect of abortion bans on infant mortality rates in the United States specifically from 2021 to 2022, with an emphasis on changes following the 2022 Dobbs decision.
+Using a Difference-in-Differences (DID) approach, we assess the interaction between abortion legality in states and the timing of the injunction implemented to understand its influence on infant health outcomes.
+The study utilizes observational data from the Centers for Disease Control and Prevention’s WONDER database, capturing infant mortality trends across states with varying levels of abortion restrictiveness.
+Key predictors include the mother’s state of residence, race, age, and the year of infant death, enabling a nuanced examination of the effects of restrictive abortion laws on public health.
 
-Overview paragraph
+Our Results show that the overturning of Roe v.
+Wade and the implementation of abortion bans are linked to a rise in infant mortality rates in states where abortion became illegal.
+Through a Difference-in-Differences (DID) approach, we estimate that states enforcing abortion bans after the injunction saw an increase of 0.31 deaths per 1,000 live births compared to states where abortion remained legal.
+Furthermore, we find that maternal age and race are key determinants of infant mortality, with older maternal age groups associated with lower mortality rates and Black mothers experiencing significantly higher rates.
+These results emphasize the public health consequences of restrictive abortion laws and the urgent need for policies that mitigate these disparities.
 
-Estimand paragraph
+This study is important as we highlight the public health implications of abortion bans, particularly their role in increasing infant mortality rates.
+By focusing on the 2022 Dobbs decision, we provide timely evidence on how restrictive abortion policies disproportionately affect vulnerable populations, emphasizing the need for data-driven policies to mitigate these adverse outcomes.
 
-Results paragraph
-
-Why it matters paragraph
-
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+The structure of this paper is organized as follows: @sec-data  provides details on the data sources and variables used in our analysis.
+@sec-model explains the modeling approach, including the assumptions and specifications of our linear regression framework.
+In @sec-result, we present our findings, emphasizing the key predictors of Harris's support.
+Finally, @sec-discussion explores the implications of our results and suggests potential directions for future research.
+@sec-a provides external data detail, @sec-b provides model detail, @sec-c provides an datasheet, and  @sec-d provides an detailed introduction of surveys, sampling and observational data in our paper.
 
@@
 We conduct our infant death rate observational data analysis using the R programming language
-[@citeR]. Our dataset, obtained from CDC Wonder
-[@cdc_wonder_2024], based on data as of December 2022, provides
-a detailed overview of public opinion in the lead-up to the election.
-Adhering to the guidelines presented in @tellingstories, we explore
-various factors that influence voter support percentages, including the
-timing of the polls, the traits of polling organizations, and regional
-differences.
-
-Overview text
+[@citeR].
+Our dataset, obtained from CDC Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a detailed overview of the Infant Death Rate across the USA.
+CDC WONDER (Wide-ranging ONline Data for Epidemiologic Research) is an online system providing public health professionals and the public with access to a wide range of the Centers for Disease Control and Prevention (CDC) epidemiologic data and resources.
+Following the guidelines outlined in @tellingstories, we analyze the impact of abortion injunctions on infant death rates using the Difference-in-Difference method.
+This approach utilizes the interaction between the legality of abortion in the mother’s state of residence and the timing of the infant’s death as the predictor variable.
+Additionally, we include control variables such as infant year of death, the mother’s state of residence, race, and age to account for potential confounding factors.
 
@@
 
-### Abortion Legality & After Injunction
+### Interaction Term: Abortion Legality and After Injunction
 
@@
 
-Another created variable `after_injunction` indicates whether the date of infants' death falls after the implementation of the abortion restrictions following the overturn of Roe v.
+Another created variable `after_injunction` indicates whether the date of infants' death falls after the implementation of the abortion injunctions following the overturn of Roe v.
 Wade in June 2022.
 It is coded as a binary variable, where 1 represents dates after June 2022, signifying the post-injunction period, and 0 represents dates before June 2022, indicating the pre-injunction period.
+
+To evaluate the impact of abortion injunctions on infant death rates, we utilize the Difference-in-Difference (DID) method.
+As part of this approach, we construct an interaction term defined as the product of `abortion_illegal` and `after_injunction`.
+The resulting interaction term equals 1 when abortion is illegal and the time is after the injunction, and 0 otherwise.
+This term is included in the analysis to evaluate the joint impact of these factors on infant death rates.
 
@@
 ggplot() +
-  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group)) +
+  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
+    color = "black") +
   scale_fill_manual(
@@
     name = "Abortion Legal Or Not"
-  ) +
+  )+
   theme_void() +
@@
 
+```{r}
+#| label: tbl-interact
+#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing.
+States with “Illegal Abortion and After Injunction” (denoted as ‘1’) have a higher average infant death rate compared to states with “Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction” (denoted as ‘0’).
+#| echo: false
+#| warning: false
+#| message: false
+library(kableExtra)
+analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
+  group_by(illegal_after_injunction)|>
+  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`'0': Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction, '1': Illegal Abortion and After Injunction`=illegal_after_injunction)
+
+# Display the result as a table
+kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
+```
+@tbl-interact highlights average infant death rates (per 1,000 live births) based on abortion legality and injunction timing.
+States categorized as ‘1’, where abortion became illegal after the injunction, show a higher average infant death rate of 8.3 compared to 7.9 in ‘0’ states, where abortion remained legal or abortion used to be legal before the injunction.
+This suggests that restrictive abortion policies and their enforcement may contribute to higher infant mortality, potentially due to increased unintended or high-risk pregnancies, reduced access to maternal healthcare, and exacerbated social or economic disparities.
+
+## Control Variables
+
 ### Year of Death
@@
 ```{r}
-#| label: tbl-motherage
-#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother.
-It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
+#| label: fig-choromap
+#| fig-cap: Infant Death Rates Across U.S.
+States (per 1,000 Live Births), December 2022.
+Infant Mortality Rate is much higher iin Southeastern.
+Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
 #| echo: false
-#| warning: false
-#| message: false
-library(kableExtra)
-analysis_data_state <- analysis_data|>
-  group_by(state)|>
-  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`State`=state)
 
-# Display the result as a table
-kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
+library(ggnewscale)
+death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
+death_full <- us_map|>
+   full_join(death, c("region"="state"))
+
+ggplot() +
+  # First layer: Fill with death_rate
+  geom_polygon(
+    data = death_full,
+    aes(x = long, y = lat, group = group, fill = death_rate),
+    color = "black"
+  ) +
+  scale_fill_viridis_c(
+    name = "Infant Death Rate\n(per 1,000 live births)",
+    option = "C",
+    na.value = "grey50",
+    breaks = c(10, 15, 20,25),  # Custom breaks for the color legend
+  ) +
+  theme_void() +
+  labs(
+    caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
+  ) 
 ```
+@fig-choromap is a choropleth map visualizes the average infant death rate per 1,000 live births across U.S.
+states, with data categorized by color intensity.
+States with higher infant death rates are shown in lighter shades (yellow), while states with lower rates are represented in darker shades (purple).
+States with missing data are indicated in grey.
+The map highlights disparities in infant mortality rates, with some states, particularly in the southeastern U.S., exhibiting notably higher rates compared to others.
+States like Alabama and Mississippi show the highest rates, while states such as California and New York exhibit lower rates.
 

### 2024-11-21 paper/paper.qmd
update paper with model and data part

@@
 title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
-subtitle: "The infant death rate increased by 0.31 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
+subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
 author: 
@@
 bibliography: references.bib
+editor: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-The overturning of Roe v.
-Wade on June 24, 2022, through the Supreme Court’s decision in Dobbs v.
-Jackson Women’s Health Organization, ended nearly five decades of constitutional protection for abortion rights.
-This landmark ruling shifted authority over abortion regulations to individual states, creating a patchwork of laws nationwide.
-The decision reignited debates about reproductive rights and public health, highlighting disparities in access to abortion services and their effects on health outcomes, especially in states with restrictive policies.
+The overturning of Roe v. Wade on June 24, 2022, through the Supreme
+Court’s decision in Dobbs v. Jackson Women’s Health Organization, ended
+nearly five decades of constitutional protection for abortion rights.
+This landmark ruling shifted authority over abortion regulations to
+individual states, creating a patchwork of laws nationwide. The decision
+reignited debates about reproductive rights and public health,
+highlighting disparities in access to abortion services and their
+effects on health outcomes, especially in states with restrictive
+policies.
 
-Banning abortion usually has been linked to a rise in maternal mortality rates, with Texas experiencing a 56% increase from 2019 to 2022 [@nbcnews].
-However, restrictive abortion laws can also impact infant mortality, as they often lead to more unintended or high-risk pregnancies among disadvantaged populations with limited access to healthcare.
-These pregnancies are associated with higher rates of preterm births, low birth weights, and delivery complications, all of which elevate infant mortality rates.
-Even before Roe v.
-Wade was overturned, researchers highlighted the impacts of restrictive abortion laws on mortality rates.
-@Burdick2024 found that states with 11–12 restrictive laws during 2014–2018 had a 16% higher infant mortality rate, particularly in counties with higher proportions of Black populations, inadequate prenatal care, and maternal smoking.
-Similarly, @Harper2023 showed that restrictive laws are linked to higher maternal, fetal, and infant mortality, states with moderate abortion legislation had significantly lower maternal mortality (25.79 per 100,000 live births) and infant mortality (20.56 per 1,000 live births) compared to restrictive states, underscoring the health benefits of less restrictive policies.
+Banning abortion usually has been linked to a rise in maternal mortality
+rates, with Texas experiencing a 56% increase from 2019 to 2022
+[@nbcnews]. However, restrictive abortion laws can also impact infant
+mortality, as they often lead to more unintended or high-risk
+pregnancies among disadvantaged populations with limited access to
+healthcare. These pregnancies are associated with higher rates of
+preterm births, low birth weights, and delivery complications, all of
+which elevate infant mortality rates. Even before Roe v. Wade was
+overturned, researchers highlighted the impacts of restrictive abortion
+laws on mortality rates. @Burdick2024 found that states with 11–12
+restrictive laws during 2014–2018 had a 16% higher infant mortality
+rate, particularly in counties with higher proportions of Black
+populations, inadequate prenatal care, and maternal smoking. Similarly,
+@Harper2023 showed that restrictive laws are linked to higher maternal,
+fetal, and infant mortality, states with moderate abortion legislation
+had significantly lower maternal mortality (25.79 per 100,000 live
+births) and infant mortality (20.56 per 1,000 live births) compared to
+restrictive states, underscoring the health benefits of less restrictive
+policies.
 
-The estimand of our analysis is the change in the difference in infant death rates between states that banned abortion and states where abortion remained legal after June 24, 2022.
-The object of the estimation is to estimate the change in the difference in the infant death rate between the treatment and control groups based on observational data in the United States.
+The estimand of our analysis is the change in the difference in infant
+death rates between states that banned abortion and states where
+abortion remained legal after June 24, 2022. The object of the
+estimation is to estimate the change in the difference in the infant
+death rate between the treatment and control groups based on
+observational data in the United States.
 
-Instead of examining the impact of abortion laws over broader time periods like prior research, our study focuses on the effect of abortion bans on infant mortality rates in the United States specifically from 2021 to 2022, with an emphasis on changes following the 2022 Dobbs decision.
-Using a Difference-in-Differences (DID) approach, we assess the interaction between abortion legality in states and the timing of the injunction implemented to understand its influence on infant health outcomes.
-The study utilizes observational data from the Centers for Disease Control and Prevention’s WONDER database, capturing infant mortality trends across states with varying levels of abortion restrictiveness.
-Key predictors include the mother’s state of residence, race, age, and the year of infant death, enabling a nuanced examination of the effects of restrictive abortion laws on public health.
+Instead of examining the impact of abortion laws over broader time
+periods like prior research, our study focuses on the effect of abortion
+bans on infant mortality rates in the United States specifically from
+2021 to 2022, with an emphasis on changes following the 2022 Dobbs
+decision. Using a Difference-in-Differences (DID) approach, we assess
+the interaction between abortion legality in states and the timing of
+the injunction implemented to understand its influence on infant health
+outcomes. The study utilizes observational data from the Centers for
+Disease Control and Prevention’s WONDER database, capturing infant
+mortality trends across states with varying levels of abortion
+restrictiveness. Key predictors include the mother’s state of residence,
+race, age, and the year of infant death, enabling a nuanced examination
+of the effects of restrictive abortion laws on public health.
 
-Our Results show that the overturning of Roe v.
-Wade and the implementation of abortion bans are linked to a rise in infant mortality rates in states where abortion became illegal.
-Through a Difference-in-Differences (DID) approach, we estimate that states enforcing abortion bans after the injunction saw an increase of 0.31 deaths per 1,000 live births compared to states where abortion remained legal.
-Furthermore, we find that maternal age and race are key determinants of infant mortality, with older maternal age groups associated with lower mortality rates and Black mothers experiencing significantly higher rates.
-These results emphasize the public health consequences of restrictive abortion laws and the urgent need for policies that mitigate these disparities.
+Our Results show that the overturning of Roe v. Wade and the
+implementation of abortion bans are linked to a rise in infant mortality
+rates in states where abortion became illegal. Through a
+Difference-in-Differences (DID) approach, we estimate that states
+enforcing abortion bans after the injunction saw an increase of 0.285
+deaths per 1,000 live births compared to states where abortion remained
+legal. Furthermore, we find that maternal age and race are key
+determinants of infant mortality, with older maternal age groups
+associated with lower mortality rates and Black mothers experiencing
+significantly higher rates. These results emphasize the public health
+consequences of restrictive abortion laws and the urgent need for
+policies that mitigate these disparities.
 
-This study is important as we highlight the public health implications of abortion bans, particularly their role in increasing infant mortality rates.
-By focusing on the 2022 Dobbs decision, we provide timely evidence on how restrictive abortion policies disproportionately affect vulnerable populations, emphasizing the need for data-driven policies to mitigate these adverse outcomes.
+This study is important as we highlight the public health implications
+of abortion bans, particularly their role in increasing infant mortality
+rates. By focusing on the 2022 Dobbs decision, we provide timely
+evidence on how restrictive abortion policies disproportionately affect
+vulnerable populations, emphasizing the need for data-driven policies to
+mitigate these adverse outcomes.
 
-The structure of this paper is organized as follows: @sec-data  provides details on the data sources and variables used in our analysis.
-@sec-model explains the modeling approach, including the assumptions and specifications of our linear regression framework.
-In @sec-result, we present our findings, emphasizing the key predictors of Harris's support.
-Finally, @sec-discussion explores the implications of our results and suggests potential directions for future research.
-@sec-a provides external data detail, @sec-b provides model detail, @sec-c provides an datasheet, and  @sec-d provides an detailed introduction of surveys, sampling and observational data in our paper.
+The structure of this paper is organized as follows: @sec-data provides
+details on the data sources and variables used in our analysis.
+@sec-model explains the modeling approach, including the assumptions and
+specifications of our linear regression framework. In @sec-result, we
+present our findings, emphasizing the key predictors of Harris's
+support. Finally, @sec-discussion explores the implications of our
+results and suggests potential directions for future research. @sec-a
+provides external data detail, @sec-b provides model detail, @sec-c
+provides an datasheet, and @sec-d provides an detailed introduction of
+surveys, sampling and observational data in our paper.
 
@@
 
-We conduct our infant death rate observational data analysis using the R programming language
-[@citeR].
-Our dataset, obtained from CDC Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a detailed overview of the Infant Death Rate across the USA.
-CDC WONDER (Wide-ranging ONline Data for Epidemiologic Research) is an online system providing public health professionals and the public with access to a wide range of the Centers for Disease Control and Prevention (CDC) epidemiologic data and resources.
-Following the guidelines outlined in @tellingstories, we analyze the impact of abortion injunctions on infant death rates using the Difference-in-Difference method.
-This approach utilizes the interaction between the legality of abortion in the mother’s state of residence and the timing of the infant’s death as the predictor variable.
-Additionally, we include control variables such as infant year of death, the mother’s state of residence, race, and age to account for potential confounding factors.
+We conduct our infant death rate observational data analysis using the R
+programming language [@citeR]. Our dataset, obtained from CDC
+Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a
+detailed overview of the Infant Death Rate across the USA. CDC WONDER
+(Wide-ranging ONline Data for Epidemiologic Research) is an online
+system providing public health professionals and the public with access
+to a wide range of the Centers for Disease Control and Prevention (CDC)
+epidemiologic data and resources. Following the guidelines outlined in
+@tellingstories, we analyze the impact of abortion injunctions on infant
+death rates using the Difference-in-Difference method. This approach
+utilizes the interaction between the legality of abortion in the
+mother’s state of residence and the timing of the infant’s death as the
+predictor variable. Additionally, we include control variables such as
+infant year of death, the mother’s state of residence, race, and age to
+account for potential confounding factors.
 
 ## Measurement
+
+The process of translating real-world events into our dataset requires a
+systematic approach to measurement and data gathering. In this research,
+we analyze infant mortality rates in the United States to understand the
+impact of abortion policies and demographic factors. The CDC’s Linked
+Birth/Infant Death Records database provides comprehensive data,
+including maternal and infant characteristics, to support this study.
+Variables such as maternal age, race, education, birth weight,
+gestational age, and prenatal care are included to ensure a detailed
+examination of factors influencing infant health outcomes.
+
+The data, sourced from birth and death certificates, undergoes a
+thorough cleaning and validation process to address inconsistencies and
+missing information, ensuring accuracy. Infant mortality rates are
+calculated per 1,000 live births, with additional indicators such as
+reliability labels for rates based on fewer than 20 deaths. To further
+enhance data accuracy, weights are applied to adjust for unlinked infant
+death records, ensuring that all deaths—linked and unlinked—are
+proportionally represented. Weights are calculated for each state and
+infant age-at-death cohort (\<7 days, 7-27 days, and 28 days to 1 year)
+using the formula:
+$\text{Weight} = \frac{\text{Linked deaths} + \text{Unlinked deaths}}{\text{Linked deaths}}$
+This adjustment means areas with complete linkages have a weight of 1.0,
+while areas with incomplete linkages are adjusted proportionally.
+Additionally, imputation techniques are employed to handle missing
+birthweight data, assigning values based on gestational period, race,
+sex, and plurality, further reducing potential bias in the analysis.
+
+Once cleaned, the data is stratified by maternal and infant attributes,
+providing insights into trends and disparities across demographic
+groups. Statistical models are employed to estimate associations between
+state-level abortion policies and infant mortality, controlling for
+confounders such as maternal characteristics. This structured approach
+transforms complex datasets into actionable insights, shedding light on
+the health implications of varying abortion laws across the U.S. By
+leveraging comprehensive datasets and robust statistical techniques,
+this study offers a nuanced understanding of the intersection between
+policy, demographics, and infant health outcomes, informing public
+health strategies and legislative decisions.
 
@@
 
-The `death rate` refers to the number of deaths of infants under one year of age per 1,000 live births in a specified population during a given time period.
-It serves as a key indicator of a population’s overall health and access to healthcare.
-Rates labeled as "Unreliable" occur when there are fewer than 20 deaths in the numerator, as these figures do not meet the National Center for Health Statistics (NCHS) standards for reliability or precision.
-After cleaning the dataset, the "Unreliable" label is converted into a separate variable named `reliable`.
+The `death rate` refers to the number of deaths of infants under one
+year of age per 1,000 live births in a specified population during a
+given time period. It serves as a key indicator of a population’s
+overall health and access to healthcare. Rates labeled as "Unreliable"
+occur when there are fewer than 20 deaths in the numerator, as these
+figures do not meet the National Center for Health Statistics (NCHS)
+standards for reliability or precision. After cleaning the dataset, the
+"Unreliable" label is converted into a separate variable named
+`reliable`.
 
@@
 ```
-@fig-deathrate illustrates the distribution of infant death rates (per 1,000 live births) based on the available data.
-The majority of infant death rates cluster around lower values, with a peak observed between 5 and 10 deaths per 1,000 infants.
-As the death rate increases, the frequency of occurrences decreases, indicating that higher infant death rates are relatively rare.
-There are few instances where the rate exceeds 20 deaths per 1,000 infants, suggesting these are outliers.
-Overall, the distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
+
+@fig-deathrate illustrates the distribution of infant death rates (per
+1,000 live births) based on the available data. The majority of infant
+death rates cluster around lower values, with a peak observed between 5
+and 10 deaths per 1,000 infants. As the death rate increases, the
+frequency of occurrences decreases, indicating that higher infant death
+rates are relatively rare. There are few instances where the rate
+exceeds 20 deaths per 1,000 infants, suggesting these are outliers.
+Overall, the distribution is right-skewed, highlighting the predominance
+of lower infant death rates in the dataset.
 
@@
 
-Abortion legality (named `abortion_illegal` in the dataset) is a newly created variable in our dataset, which refers to the legal status of abortion within a given state.
-It is represented as a binary variable, where 1 indicates that abortion is illegal in the state, and 0 indicates that abortion is legal.
-This classification reflects the policies and laws governing abortion access in each state.
+Abortion legality (named `abortion_illegal` in the dataset) is a newly
+created variable in our dataset, which refers to the legal status of
+abortion within a given state. It is represented as a binary variable,
+where 1 indicates that abortion is illegal in the state, and 0 indicates
+that abortion is legal. This classification reflects the policies and
+laws governing abortion access in each state.
 
-Another created variable `after_injunction` indicates whether the date of infants' death falls after the implementation of the abortion injunctions following the overturn of Roe v.
-Wade in June 2022.
-It is coded as a binary variable, where 1 represents dates after June 2022, signifying the post-injunction period, and 0 represents dates before June 2022, indicating the pre-injunction period.
+Another created variable `after_injunction` indicates whether the date
+of infants' death falls after the implementation of the abortion
+injunctions following the overturn of Roe v. Wade in June 2022. It is
+coded as a binary variable, where 1 represents dates after June 2022,
+signifying the post-injunction period, and 0 represents dates before
+June 2022, indicating the pre-injunction period.
 
-To evaluate the impact of abortion injunctions on infant death rates, we utilize the Difference-in-Difference (DID) method.
-As part of this approach, we construct an interaction term defined as the product of `abortion_illegal` and `after_injunction`.
-The resulting interaction term equals 1 when abortion is illegal and the time is after the injunction, and 0 otherwise.
-This term is included in the analysis to evaluate the joint impact of these factors on infant death rates.
+To evaluate the impact of abortion injunctions on infant death rates, we
+utilize the Difference-in-Difference (DID) method. As part of this
+approach, we construct an interaction term defined as the product of
+`abortion_illegal` and `after_injunction`. The resulting interaction
+term equals 1 when abortion is illegal and the time is after the
+injunction, and 0 otherwise. This term is included in the analysis to
+evaluate the joint impact of these factors on infant death rates.
 
 ```{r}
-#|include: false
+#| include: false
+#| warning: false
+#| message: false
+
 library(usmap)
@@
 ```{r}
-#|include: false
+#| include: false
+#| warning: false
+#| message: false
 us_map <- map_data("state")
@@
     caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
-  ) 
+  ) +
+  theme(legend.position = "bottom")
 ```
-@fig-map illustrates the legality of abortion across U.S.
-states.
-Based on @reproductiveright, as of November 2024, abortion has been illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin following the June 2022 decision.
+
+@fig-map illustrates the legality of abortion across U.S. states. Based
+on @reproductiveright, as of November 2024, abortion has been illegal in
+Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi,
+Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West
+Virginia, and Wisconsin following the June 2022 decision.
+
 ```{r}
@@
 #| echo: false
+#| message: false
 analysis_data|>
@@
   scale_color_brewer(palette = "Set1")+
-  theme_classic() 
+  theme_classic() +
+  theme(legend.position = "bottom")
 ```
-@fig-rateandstate illustrates trends in infant death rates (per 1,000 live births) over time in states where abortion is illegal versus legal.
-In states with abortion bans, infant mortality shows a slight increase after June 2022, marked by the dashed line representing the overturn of Roe v.
-Wade.
-In contrast, states where abortion remains legal exhibit relatively stable trends.
-This suggests a potential link between abortion bans and rising infant mortality rates, warranting further investigation.
+
+@fig-rateandstate illustrates trends in infant death rates (per 1,000
+live births) over time in states where abortion is illegal versus legal.
+In states with abortion bans, infant mortality shows a slight increase
+after June 2022, marked by the dashed line representing the overturn of
+Roe v. Wade. In contrast, states where abortion remains legal exhibit
+relatively stable trends. This suggests a potential link between
+abortion bans and rising infant mortality rates, warranting further
+investigation.
 
@@
 #| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing.
-States with “Illegal Abortion and After Injunction” (denoted as ‘1’) have a higher average infant death rate compared to states with “Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction” (denoted as ‘0’).
+States with "Illegal Abortion and After Injunction" (denoted as '1') have a higher average infant death rate compared to states with "Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction" (denoted as '0').
 #| echo: false
@@
   group_by(illegal_after_injunction)|>
-  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`'0': Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction, '1': Illegal Abortion and After Injunction`=illegal_after_injunction)
+  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))
 
-# Display the result as a table
-kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
+kable(analysis_data_state, align = "c") |> 
+  footnote(
+    "Data Source: CDC Wonder"
+  )
 ```
-@tbl-interact highlights average infant death rates (per 1,000 live births) based on abortion legality and injunction timing.
-States categorized as ‘1’, where abortion became illegal after the injunction, show a higher average infant death rate of 8.3 compared to 7.9 in ‘0’ states, where abortion remained legal or abortion used to be legal before the injunction.
-This suggests that restrictive abortion policies and their enforcement may contribute to higher infant mortality, potentially due to increased unintended or high-risk pregnancies, reduced access to maternal healthcare, and exacerbated social or economic disparities.
+
+@tbl-interact highlights average infant death rates (per 1,000 live
+births) based on abortion legality and injunction timing. States
+categorized as ‘1’, where abortion became illegal after the injunction,
+show a higher average infant death rate of 8.3 compared to 7.9 in ‘0’
+states, where abortion remained legal or abortion used to be legal
+before the injunction. This suggests that restrictive abortion policies
+and their enforcement may contribute to higher infant mortality,
+potentially due to increased unintended or high-risk pregnancies,
+reduced access to maternal healthcare, and exacerbated social or
+economic disparities.
 
@@
 
+The Year of Death variable represents the calendar year in which the
+infant death occurred, specifically focusing on the years 2021 and 2022
+in this study. This variable is essential for analyzing temporal trends
+and identifying patterns or shifts in infant mortality rates over this
+critical period. By controlling the year of death, we can closely
+examine the impact of significant policy changes, such as the
+overturning of Roe v. Wade in 2022, as well as other temporal factors on
+infant health outcomes.
+
 ### State
 
-The state of the mother’s legal residence at the time of birth. This data is categorized by the mother’s state of residence and excludes territories.
+The state of the mother’s legal residence at the time of birth. This
+data is categorized by the mother’s state of residence and excludes
+territories.
 
@@
     caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
-  ) 
+  ) +
+  theme(legend.position = "bottom")
 ```
-@fig-choromap is a choropleth map visualizes the average infant death rate per 1,000 live births across U.S.
-states, with data categorized by color intensity.
-States with higher infant death rates are shown in lighter shades (yellow), while states with lower rates are represented in darker shades (purple).
-States with missing data are indicated in grey.
-The map highlights disparities in infant mortality rates, with some states, particularly in the southeastern U.S., exhibiting notably higher rates compared to others.
-States like Alabama and Mississippi show the highest rates, while states such as California and New York exhibit lower rates.
+
+@fig-choromap is a choropleth map visualizes the average infant death
+rate per 1,000 live births across U.S. states, with data categorized by
+color intensity. States with higher infant death rates are shown in
+lighter shades (yellow), while states with lower rates are represented
+in darker shades (purple). States with missing data are indicated in
+grey. The map highlights disparities in infant mortality rates, with
+some states, particularly in the southeastern U.S., exhibiting notably
+higher rates compared to others. States like Alabama and Mississippi
+show the highest rates, while states such as California and New York
+exhibit lower rates.
 
@@
 
-The Age of Mother (`age_of_mother`) refers to the age group of the mother at the time of childbirth.In our observational data, age is categorized into the brackets 15–19 years, 20–24 years, 25–29 years, and 30–34 years in the raw data, in compliance with privacy requirements.
+The Age of Mother (`age_of_mother`) refers to the age group of the
+mother at the time of childbirth.In our observational data, age is
+categorized into the brackets 15–19 years, 20–24 years, 25–29 years, and
+30–34 years in the raw data, in compliance with privacy requirements.
 
-@tbl-motherage displays the average infant death rate (per 1,000 live births) by the age group of mothers.
-It shows a clear trend: younger mothers, particularly those aged 15–19 and 20–24, have higher infant death rates (9.0 and 10.0, respectively), while older mothers in the 25–29 and 30–34 age groups have lower rates (7.4 and 6.6, respectively).
-This suggests that maternal age may be a factor influencing infant mortality, with younger mothers experiencing higher risks.
+@tbl-motherage displays the average infant death rate (per 1,000 live
+births) by the age group of mothers. It shows a clear trend: younger
+mothers, particularly those aged 15–19 and 20–24, have higher infant
+death rates (9.0 and 10.0, respectively), while older mothers in the
+25–29 and 30–34 age groups have lower rates (7.4 and 6.6, respectively).
+This suggests that maternal age may be a factor influencing infant
+mortality, with younger mothers experiencing higher risks.
+
 ```{r}
@@
 ### Mother's Single Race
-The self-identified race of the mother, recorded as a single race category.
-In this dataset, the available categories are: Asian, Black or African American, and White.
-This classification facilitates the analysis of maternal and infant health outcomes across these racial groups.
+
+The self-identified race of the mother, recorded as a single race
+category. In this dataset, the available categories are: Asian, Black or
+African American, and White. This classification facilitates the
+analysis of maternal and infant health outcomes across these racial
+groups.
+
 ```{r}
-#| label: tbl-motherage
+#| label: tbl-motherrace
 #| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother.
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+For our analysis, we use a Bayesian Difference-in-Differences (DID) model to quantify the impact of abortion injunctions on the infant death rate across states in the USA.
+This method compares changes in infant death rates over time between a treatment group (states where abortion injunctions were implemented) and a control group (states where abortion remains legal), providing a robust framework for estimating causal effects.
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+The model focuses on the infant death rate per 1,000 live births as the dependent variable and includes the following independent variables:\
+	-	Treatment Group Indicator `abortion_illegal`: A binary variable indicating the group: 1 if abortion is illegal in the state. 0 if abortion remains legal.\
+	-	Time Indicator `after_injunction`: A binary variable indicating the time period: 0 for the period before the injunction implemented. 1 for the period after the injunction implemented.
+	-	Interaction Term`after_injunction` × `abortion_illegal`: The interaction term captures the additional change in the outcome (e.g., percentage of support, behavior, or health-related metrics) for regions where abortion is illegal after the injunction implemented, beyond any changes observed in regions where abortion remains legal.
+
+To account for potential regional variations in infant death rates, we include state as a random effect in the model.
+This allows us to control for unobserved heterogeneity across states that might introduce bias.
+Additionally, we include age_of_mother, mothers_single_race, and year_of_death as control variables to adjust for demographic and temporal factors that could influence infant mortality rates, ensuring that the estimated effects are not confounded by these characteristics.
+
+By isolating the impact of the abortion injunction through this interaction term, the model identifies changes in infant death rates that can be attributed to the policy intervention.
+Additional details and diagnostics supporting the model’s implementation are provided in [Appendix-@sec-model-details].
+
+## Difference-In-Difference Approach (DID)
+
+DID is a causal inference method that estimates the impact of an intervention by comparing changes in outcomes over time between a treatment group and a control group.
+By assuming parallel trends in the absence of the intervention, the DID framework isolates the intervention’s effect through the interaction between time periods (pre- and post-intervention) and group membership.
+This makes it a robust method for evaluating policy changes and other significant events.
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+In our analysis, we employ two models.
+The first model includes only the predictor variables and the random effect for state.
+The second model builds on the first by incorporating additional control variables to account for potential confounding factors.
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+Define $y_i$ as as the infant death rate per 1,000 live births.
+$$y_i|\mu_i,\sigma \sim \mbox{Normal}(\mu_i, \sigma)\\
+\text{First Model}: \mu_i = \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}+ \beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
+\text{Second Model}:\mu_i = \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
+\beta_0 \sim \mbox{Normal}(0, 2.5) \\
+\beta_1 \sim \mbox{Normal}(0, 2.5) \\
+\beta_2 \sim \mbox{Normal}(0, 2.5) \\
+\beta_3 \sim \mbox{Normal}(0, 2.5) \\
+\beta_4 \sim \mbox{Normal}(0, 2.5) \\
+\beta_5 \sim \mbox{Normal}(0, 2.5) \\
+\beta_6 \sim \mbox{Normal}(0, 2.5) \\
+\gamma_j \sim \text{Normal}(0, \sigma_j^2)\\
+\sigma \sim \mbox{Exponential}(1)$$
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+The models are run in R (R Core Team 2023) using the rstanarm package of @rstanarm.
+The default priors from `rstanarm` are used for both GLM Bayesian models.
+We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.
+We use the default priors from `rstanarm` used for both Gaussian Bayesian models.
 
@@
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
+The Bayesian framework enhances the DID approach by incorporating prior information and quantifying uncertainty in a probabilistic manner.
+Using Markov Chain Monte Carlo (MCMC) sampling, the model estimates the posterior distributions of the parameters, offering flexibility in handling complex relationships and uncertainties in the data.
 
@@
 #| message: false
+#| include: false
 
@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+If my paper were 10 pages, then should be be at least 2.5 pages. The
+discussion is a chance to show off what you know and what you learnt
+from all this.
 
@@
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+Please don't use these as sub-heading labels - change them to be what
+your point actually is.
 
@@
 
-# Appendix {-}
+# Appendix {.unnumbered}
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
+check. This shows...
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
+prior. This shows...
 
@@
   coord_flip()
+```
+
+```{r}
+plot(first_model, pars = "(Intercept)", prob = 0.95)
+```
+
+```{r}
+prior_summary(second_model)
 ```
@@
 ```{r}
-analysis_data|>
-  ggplot(aes(x = after_injunction, y = death_rate, color = as.factor(abortion_illegal))) +
-  geom_point(alpha = 0.2) +
-  geom_line(aes(group = state)) +
-  theme_minimal() +
-  labs(x = "Time period", y = "Death Rate", color = "Person got a new racket")  +
-  scale_color_brewer(palette = "Set1") +
+plot(
+  first_model,
+  "areas"
+)
+plot(
+  second_model,
+  "areas"
+)
+```
+
+```{r}
+pp_check(second_model) +
+  theme_classic() +
   theme(legend.position = "bottom")

### 2024-11-24 paper/paper.qmd
update paper

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This study examines the impact of the Dobbs decision and subsequent abortion bans on infant mortality rates in the United States, focusing on data from 2021 to 2022.
+Using a Difference-in-Differences (DID) approach, we analyze trends in states where abortion remained legal compared to those where it became illegal after June 2022.
+Our results indicate that states enforcing abortion bans experienced an increase of 0.285 infant deaths per 1,000 live births, with compounded effects observed during the post-injunction period.
+Maternal age and race also emerged as significant predictors, with Black mothers facing disproportionately higher infant mortality rates.
+These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities.
+By utilizing recent, high-quality data and rigorous modeling, this study provides critical insights into the intersection of reproductive rights and infant health outcomes."
 format: pdf
+toc: true
 number-sections: true
@@
 support. Finally, @sec-discussion explores the implications of our
-results and suggests potential directions for future research. @sec-a
-provides external data detail, @sec-b provides model detail, @sec-c
-provides an datasheet, and @sec-d provides an detailed introduction of
-surveys, sampling and observational data in our paper.
+results and suggests potential directions for future research.
+[Appendix- @sec-data-details] provides external data detail, [Appendix-
+@sec-model-details] provides model detail, @sec-c provides an datasheet,
+and @sec-d provides an detailed introduction of surveys, sampling and
+observational data in our paper.
 
@@
 
-### State
-
-The state of the mother’s legal residence at the time of birth. This
-data is categorized by the mother’s state of residence and excludes
-territories.
-
-```{r}
-#| label: fig-choromap
-#| fig-cap: Infant Death Rates Across U.S.
-States (per 1,000 Live Births), December 2022.
-Infant Mortality Rate is much higher iin Southeastern.
-Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
-#| echo: false
-
-library(ggnewscale)
-death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
-death_full <- us_map|>
-   full_join(death, c("region"="state"))
-
-ggplot() +
-  # First layer: Fill with death_rate
-  geom_polygon(
-    data = death_full,
-    aes(x = long, y = lat, group = group, fill = death_rate),
-    color = "black"
-  ) +
-  scale_fill_viridis_c(
-    name = "Infant Death Rate\n(per 1,000 live births)",
-    option = "C",
-    na.value = "grey50",
-    breaks = c(10, 15, 20,25),  # Custom breaks for the color legend
-  ) +
-  theme_void() +
-  labs(
-    caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
-  ) +
-  theme(legend.position = "bottom")
-```
-
-@fig-choromap is a choropleth map visualizes the average infant death
-rate per 1,000 live births across U.S. states, with data categorized by
-color intensity. States with higher infant death rates are shown in
-lighter shades (yellow), while states with lower rates are represented
-in darker shades (purple). States with missing data are indicated in
-grey. The map highlights disparities in infant mortality rates, with
-some states, particularly in the southeastern U.S., exhibiting notably
-higher rates compared to others. States like Alabama and Mississippi
-show the highest rates, while states such as California and New York
-exhibit lower rates.
-
 ### Age of Mother
@@
 
-# Model
+## Random Variable
 
-For our analysis, we use a Bayesian Difference-in-Differences (DID) model to quantify the impact of abortion injunctions on the infant death rate across states in the USA.
-This method compares changes in infant death rates over time between a treatment group (states where abortion injunctions were implemented) and a control group (states where abortion remains legal), providing a robust framework for estimating causal effects.
+### State
 
-The model focuses on the infant death rate per 1,000 live births as the dependent variable and includes the following independent variables:\
-	-	Treatment Group Indicator `abortion_illegal`: A binary variable indicating the group: 1 if abortion is illegal in the state. 0 if abortion remains legal.\
-	-	Time Indicator `after_injunction`: A binary variable indicating the time period: 0 for the period before the injunction implemented. 1 for the period after the injunction implemented.
-	-	Interaction Term`after_injunction` × `abortion_illegal`: The interaction term captures the additional change in the outcome (e.g., percentage of support, behavior, or health-related metrics) for regions where abortion is illegal after the injunction implemented, beyond any changes observed in regions where abortion remains legal.
+The state of the mother’s legal residence at the time of birth. This
+data is categorized by the mother’s state of residence and excludes
+territories.
 
-To account for potential regional variations in infant death rates, we include state as a random effect in the model.
-This allows us to control for unobserved heterogeneity across states that might introduce bias.
-Additionally, we include age_of_mother, mothers_single_race, and year_of_death as control variables to adjust for demographic and temporal factors that could influence infant mortality rates, ensuring that the estimated effects are not confounded by these characteristics.
+```{r}
+#| label: fig-choromap
+#| fig-cap: Infant Death Rates Across U.S.
+States (per 1,000 Live Births), December 2022.
+Infant Mortality Rate is much higher iin Southeastern.
+Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
+#| echo: false
 
-By isolating the impact of the abortion injunction through this interaction term, the model identifies changes in infant death rates that can be attributed to the policy intervention.
-Additional details and diagnostics supporting the model’s implementation are provided in [Appendix-@sec-model-details].
+library(ggnewscale)
+death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
+death_full <- us_map|>
+   full_join(death, c("region"="state"))
+
+ggplot() +
+  # First layer: Fill with death_rate
+  geom_polygon(
+    data = death_full,
+    aes(x = long, y = lat, group = group, fill = death_rate),
+    color = "black"
+  ) +
+  scale_fill_viridis_c(
+    name = "Infant Death Rate\n(per 1,000 live births)",
+    option = "C",
+    na.value = "grey50",
+    breaks = c(10, 15, 20,25),  # Custom breaks for the color legend
+  ) +
+  theme_void() +
+  labs(
+    caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
+  ) +
+  theme(legend.position = "bottom")
+```
+
+@fig-choromap is a choropleth map visualizes the average infant death
+rate per 1,000 live births across U.S. states, with data categorized by
+color intensity. States with higher infant death rates are shown in
+lighter shades (yellow), while states with lower rates are represented
+in darker shades (purple). States with missing data are indicated in
+grey. The map highlights disparities in infant mortality rates, with
+some states, particularly in the southeastern U.S., exhibiting notably
+higher rates compared to others. States like Alabama and Mississippi
+show the highest rates, while states such as California and New York
+exhibit lower rates. \# Model
+
+For our analysis, we use a Bayesian Difference-in-Differences (DID)
+model to quantify the impact of abortion injunctions on the infant death
+rate across states in the USA. This method compares changes in infant
+death rates over time between a treatment group (states where abortion
+injunctions were implemented) and a control group (states where abortion
+remains legal), providing a robust framework for estimating causal
+effects.
 
@@
 
-DID is a causal inference method that estimates the impact of an intervention by comparing changes in outcomes over time between a treatment group and a control group.
-By assuming parallel trends in the absence of the intervention, the DID framework isolates the intervention’s effect through the interaction between time periods (pre- and post-intervention) and group membership.
-This makes it a robust method for evaluating policy changes and other significant events.
+Difference-in-Differences (DID) is a causal inference method that
+estimates the impact of an intervention by comparing changes in outcomes
+over time between a treatment group and a control group. By assuming
+parallel trends in the absence of the intervention, the DID framework
+isolates the intervention’s effect through **the interaction between
+time periods (pre- and post-intervention) and group membership**. This
+makes it a robust method for evaluating policy changes and other
+significant events.
+
+### Assumption
+
+The validity of the Difference-in-Differences (DID) approach in our
+study relies on several key assumptions, which, if violated, could
+affect the robustness and reliability of our results [@cunningham2021]:
+
+#### Parallel Trends Assumption
+
+The model relies on the assumption that, without the abortion
+injunction, the treatment group (states where abortion became illegal)
+and the control group (states where abortion remained legal) would have
+followed similar trends in infant mortality rates. However, this
+assumption can be challenged by differences in healthcare systems,
+socioeconomic conditions, or pre-existing policy environments, which may
+cause deviations from parallel trends.
+
+#### Compositional Differences
+
+In addition to parallel trends, compositional changes in repeated
+cross-sectional data over time can influence results. For instance,
+shifts in demographic factors such as maternal age or racial makeup
+between pre- and post-intervention periods may confound the observed
+changes in infant mortality rates. Including control variables helps
+adjust for these confounders, ensuring that the estimated effect
+reflects the policy intervention rather than changes in population
+composition.
+
+#### Long-term Effects vs. Reliability
+
+The length of the analysis period further complicates the reliability of
+results. While longer periods allow for capturing long-term effects,
+they also increase the likelihood of unrelated factors, such as new
+policies or healthcare advancements, influencing the outcome. Over time,
+some states in the control group may also implement similar abortion
+policies, diluting the treatment effect. Conversely, shorter analysis
+periods
+
+#### Functional Form Dependence
+
+Finally, the model assumes that the functional relationship between
+predictors and outcomes is correctly specified. If this assumption is
+violated, differences in outcomes between treatment and control groups
+may stem from misspecified relationships rather than the intervention
+itself. This highlights the importance of carefully assessing the
+model’s structure and assumptions to ensure valid results.
+
+## Model Parameters
+
+The model focuses on the infant death rate per 1,000 live births as the
+dependent variable and includes the following independent variables:
+
+-   **Treatment Group Indicator `abortion_illegal`**: A binary variable
+    indicating the group: 1 if abortion is illegal in the state. 0 if
+    abortion remains legal.
+
+-   **Time Indicator `after_injunction`**: A binary variable indicating
+    the time period: 0 for the period before the injunction implemented.
+    1 for the period after the injunction implemented.
+
+-   **Interaction Term`after_injunction` × `abortion_illegal`**: The
+    interaction term captures the additional change in the outcome
+    (e.g., percentage of support, behavior, or health-related metrics)
+    for regions where abortion is illegal after the injunction
+    implemented, beyond any changes observed in regions where abortion
+    remains legal.
+
+Additionally, we include these control variables to adjust for
+demographic and temporal factors that could influence infant mortality
+rates, ensuring that the estimated effects are not confounded by these
+characteristics:
+
+-   **`age_of_mother`**: The age of the mother at the time of birth,
+    represented as a continuous variable.
+
+-   **`mothers_single_race`**: Self-identified race of the mother,
+    represented as a categorical variable.
+
+-   **`year_of_death`**: The year in which the infant death occurred,
+    represented as a continuous variable.
+
+Finally, we include a random variable:
+
+-   **`state`**: Mother's residence state when the infant born as a
+    random effect in the model.
+
+By isolating the impact of the abortion injunction through this
+interaction term `after_injunction`×`abortion_illegal`, our model
+identifies changes in infant death rates that can be attributed to the
+policy intervention. Additional details and diagnostics supporting the
+model’s implementation are provided in [Appendix- @sec-model-details].
 
@@
 
-In our analysis, we employ two models.
-The first model includes only the predictor variables and the random effect for state.
-The second model builds on the first by incorporating additional control variables to account for potential confounding factors.
+In our analysis, we employ two models. The first model includes only the
+predictor variables and the random effect for state. The second model
+builds on the first by incorporating additional control variables to
+account for potential confounding factors.
 
-Define $y_i$ as as the infant death rate per 1,000 live births.
-$$y_i|\mu_i,\sigma \sim \mbox{Normal}(\mu_i, \sigma)\\
-\text{First Model}: \mu_i = \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}+ \beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
-\text{Second Model}:\mu_i = \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
-\beta_0 \sim \mbox{Normal}(0, 2.5) \\
-\beta_1 \sim \mbox{Normal}(0, 2.5) \\
-\beta_2 \sim \mbox{Normal}(0, 2.5) \\
-\beta_3 \sim \mbox{Normal}(0, 2.5) \\
-\beta_4 \sim \mbox{Normal}(0, 2.5) \\
-\beta_5 \sim \mbox{Normal}(0, 2.5) \\
-\beta_6 \sim \mbox{Normal}(0, 2.5) \\
-\gamma_j \sim \text{Normal}(0, \sigma_j^2)\\
-\sigma \sim \mbox{Exponential}(1)$$
+Define $y_i$ as as the infant death rate per 1,000 live births, and
+$\gamma_j$ as the random state effect. $$
+\begin{aligned}
+y_i|\mu_i,\sigma &\sim \mbox{Normal}(\mu_i, \sigma)\\
+\text{First Model}: \mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}\\&+\beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
+\text{Second Model}:\mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i\\&+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\&+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
+\beta_0 &\sim \mbox{Normal}(0, 2.5)\\
+\beta_1 &\sim \mbox{Normal}(0, 2.5)\\
+\beta_2 &\sim \mbox{Normal}(0, 2.5)\\
+\beta_3 &\sim \mbox{Normal}(0, 2.5)\\
+\beta_4 &\sim \mbox{Normal}(0, 2.5)\\
+\beta_5 &\sim \mbox{Normal}(0, 2.5)\\
+\beta_6 &\sim \mbox{Normal}(0, 2.5)\\
+\gamma_j&\sim \text{Normal}(0, \sigma_j^2)\\
+\sigma &\sim \mbox{Exponential}(1)
+\end{aligned}$$
 
-The models are run in R (R Core Team 2023) using the rstanarm package of @rstanarm.
-The default priors from `rstanarm` are used for both GLM Bayesian models.
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.
-We use the default priors from `rstanarm` used for both Gaussian Bayesian models.
+The models are run in R (R Core Team 2023) using the rstanarm package of
+@rstanarm. The default priors from `rstanarm` are used for both GLM
+Bayesian models. We run the model in R [@citeR] using the `rstanarm`
+package of @rstanarm. We use the default priors from `rstanarm` used for
+both Gaussian Bayesian models.
 
@@
 
-The Bayesian framework enhances the DID approach by incorporating prior information and quantifying uncertainty in a probabilistic manner.
-Using Markov Chain Monte Carlo (MCMC) sampling, the model estimates the posterior distributions of the parameters, offering flexibility in handling complex relationships and uncertainties in the data.
+For our analysis, we use separate Bayesian regression models to estimate
+the infant death rates in states where abortion injunctions were
+implemented and in states where abortion remains legal. We use the
+interaction term between after_injunction and abortion_illegal to
+measure the causal effect of the abortion injunction. Both models share
+the same predictors to ensure consistency and comparability in capturing
+the determinants of infant mortality.
+
+For one of our models, we include control variables (`age_of_mother`,
+`mothers_single_race`, and `year_of_death`) to adjust for confounders
+that could influence infant death rates, addressing the assumption of
+compositional differences in the DID approach and ensuring unbiased
+estimates of the intervention’s effect. In the other model, we exclude
+these controls to focus on the overall intervention impact, avoid
+overfitting, and prioritize simplicity when variability is adequately
+accounted for by the design or random effects. Thus, our decision to fit
+two models is to compare the outcomes: one adjusts for confounders to
+refine the estimates, while the other focuses on the direct effect of
+the intervention.
+
+Our model includes `state` as a random effect to account for regional
+variations in infant mortality that could introduce bias. We use a
+random effect rather than a fixed effect to capture variability across
+states without estimating specific effects for each one. This approach
+models unobserved heterogeneity, such as differences in healthcare
+systems and socioeconomic conditions, using a single variance term,
+which avoids overfitting and ensures efficiency and generalizability. By
+adjusting for unmeasured state-specific factors, the model provides
+robust and reliable findings while balancing parsimony and
+interpretability, making it well-suited for our analysis.
+
+We choose the Bayesian framework since it enhances the DID approach by
+incorporating prior information and quantifying uncertainty in a
+probabilistic manner. In our Bayesian models, we use weakly informative
+priors from the `rstanarm` package, including a Normal prior ($\mu=0$
+and $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1)
+for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
+priors based on data scale. This enhances convergence and ensures priors
+are calibrated to predictor ranges without dominating the data. The
+model assumes plausible parameter centering, random effects capture
+unobserved heterogeneity, and the DID framework relies on parallel
+trends between groups. However, limitations include potential bias from
+misspecified priors, computational challenges for hierarchical models,
+and sensitivity to data sparsity or imbalances. This model may not be
+suitable if group-level effects are negligible, parallel trends are
+violated, or data is insufficient, but when applied appropriately, it
+provides robust and interpretable causal estimates.
 
@@
 
-Our results are summarized in @tbl-modelresults.
-
 ```{r}
 #| echo: false
-#| eval: true
 #| warning: false
@@
   list(first_model,second_model),
-  fmt = 3,coef_omit="state",notes = "Note.
-All models include a control for state and pollsters.
-The reference level of the state is Alaska, the reference level of population is likely voter, and the reference level of pollsters is Angus Reid.")
+  fmt = 3,coef_omit="state",notes = "Note. All models include a random variable for state. The reference level of age_of_mother is 15-19.The reference level of monthers_single_race is Asian.")
 ```
+
+The results from the two Bayesian models are summarized in
+@tbl-modelresults. The interaction term between after_injunction and
+abortion_illegal is positive and statistically significant (0.111 in
+Model 1 and 0.280 in Model 2), indicating a change in infant mortality
+rates associated with the combination of abortion bans and the
+post-injunction period.
+
+For the independent effects, states where abortion is illegal show
+higher infant mortality rates, with coefficients of 3.365 in Model 1 and
+1.400 in Model 2. The post-injunction period alone is associated with
+coefficients of -0.525 in Model 1 and -0.940 in Model 2, indicating a
+decrease in infant mortality during this time.
+
+Demographic variables also exhibit notable associations. Maternal age is
+negatively associated with infant mortality, with coefficients for age
+groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, respectively. The
+coefficient for Black mothers in Model 2 is 6.66, reflecting higher
+infant mortality compared to other racial groups.
+
+The models performed well, with low RMSE values and high $R^2$ values,
+indicating that the predictors account for a substantial portion of the
+variance in infant mortality rates.
 
@@
 
-## First discussion point {#sec-first-point}
+## Key Findings and Implications
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The
-discussion is a chance to show off what you know and what you learnt
-from all this.
+The analysis reveals distinct patterns in infant mortality rates across
+various policy and demographic factors. In states where abortion
+remained legal, the post-injunction period (after_injunction1) is linked
+to reductions in infant mortality rates by 0.525 and 0.940 deaths per
+1,000 live births in Models 1 and 2, respectively. However, states with
+abortion bans (abortion_illegal1) exhibit significantly higher infant
+mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
+deaths in Model 2. The DID interaction term (after_injunction1 ×
+abortion_illegal1) reveals that abortion bans during the post-injunction
+period result in an additional increase of 0.111 deaths in Model 1 and
+0.280 deaths in Model 2. Demographic factors also play a critical role,
+with older maternal age (25–34 years) reducing infant mortality rates,
+while younger mothers (15–19 years) face elevated risks. Black mothers
+experience disproportionately higher infant mortality rates, with an
+increase of 6.660 deaths per 1,000 live births compared to Asian
+mothers. These findings collectively emphasize the compounded effects of
+policy and demographic disparities on infant health outcomes.
 
-## Second discussion point
+## Public Health Implications of Restrictive Abortion Laws
 
-Please don't use these as sub-heading labels - change them to be what
-your point actually is.
+The findings highlight the significant public health consequences of
+abortion bans in the United States. The observed increase of 0.285
+infant deaths per 1,000 live births in states enforcing abortion bans
+after the Dobbs decision underscores the detrimental impact of
+restrictive abortion policies. These increases often reflect systemic
+issues such as inadequate prenatal care, gaps in healthcare access, and
+underlying socioeconomic inequities. States with pre-existing healthcare
+challenges may experience amplified effects, suggesting an urgent need
+for comprehensive policy interventions to address these disparities.
+Tackling the root causes of these health disparities—such as improving
+prenatal care, expanding healthcare access, and addressing socioeconomic
+inequalities—will be essential for mitigating the unintended
+consequences of restrictive abortion legislation.
 
-## Third discussion point
+## Racial and Socioeconomic Disparities
 
-## Weaknesses and next steps
+The results reveal stark racial disparities in infant mortality rates,
+with Black mothers experiencing significantly higher rates compared to
+other racial groups. These disparities are exacerbated by restrictive
+abortion policies, which disproportionately affect already vulnerable
+populations. The intersection of systemic healthcare inequities,
+socioeconomic challenges, and restrictive legislation creates compounded
+risks for marginalized communities, leading to worse health outcomes.
+Addressing these inequities requires targeted policies that ensure
+equitable access to maternal and infant healthcare, particularly in
+communities most impacted by these disparities. Such policies should
+prioritize investments in healthcare infrastructure, culturally
+competent care, and programs that support at-risk populations.
 
-Weaknesses and next steps should also be included.
+## Integration of Policy and Healthcare Systems
+
+The significant interaction between abortion legality and the
+post-injunction period illustrates the compounded impact of restrictive
+abortion policies on infant mortality rates. These findings underscore
+the need for integrated healthcare systems that can address the
+consequences of high-risk pregnancies, improve prenatal care, and ensure
+equitable access to medical services. Policymakers must consider the
+broader public health implications of abortion bans and implement
+strategies that safeguard maternal and infant health. This includes
+enhancing healthcare delivery in underserved areas, investing in
+community-based interventions, and ensuring that restrictive policies
+are accompanied by measures to mitigate their public health impact.
+
+## Data and Temporal Limitations
+
+A key limitation of our analysis is the temporal and demographic scope
+of the dataset. The data includes only infant deaths under 1 year of age
+occurring within the United States to U.S. residents during 2021–2022.
+This timeframe does not capture the full year of infant mortality
+outcomes following the Dobbs decision, as infants who died in 2023 are
+not included. Furthermore, the dataset’s focus on infants under 1 year
+of age may exclude cases that reflect longer-term health impacts. These
+limitations could lead to an underestimation of the full effects of
+abortion bans. Future research should incorporate more comprehensive and
+updated datasets to provide a clearer and more complete analysis of the
+long-term trends and implications of restrictive abortion policies.
+
+## Weaknesses and Future Directions
+
+While this study provides valuable findings, several limitations should
+be addressed in future research. The analysis relies on aggregate
+state-level data, which may obscure localized effects and intra-state
+disparities. Unmeasured confounders, such as economic conditions,
+healthcare infrastructure differences, or state-specific policies, could
+also influence the results. Moreover, the short timeframe of the dataset
+limits the ability to assess the long-term effects of abortion policies.
+Future research should integrate more granular data, such as
+county-level analyses, and extend the study period to capture
+longer-term impacts. Qualitative studies focusing on the lived
+experiences of affected populations can complement quantitative
+analyses, offering a deeper understanding of the broader implications of
+restrictive abortion laws. These efforts will support the development of
+evidence-based policies aimed at reducing disparities and improving
+maternal and infant health outcomes.
 
@@
 
-# Additional data details
+# Additional data details {#sec-data-details}
 
@@
 In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
-check. This shows...
+check for the first model. This shows...
 
@@
 #| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+#| fig-subcap: ["First Model Posterior prediction check", "First Model Comparing the posterior with the prior","Second Model Posterior prediction check", "Second Model Comparing the posterior with the prior"]
 
@@
   theme(legend.position = "bottom")
-
+pp_check(second_model) +
+  theme_classic() +
+  theme(legend.position = "bottom")
 posterior_vs_prior(first_model) +
+  theme_minimal() +
+  scale_color_brewer(palette = "Set1") +
+  theme(legend.position = "bottom") +
+  coord_flip()
+posterior_vs_prior(second_model) +
   theme_minimal() +
@@
 ```{r}
-plot(first_model, pars = "(Intercept)", prob = 0.95)
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+prior_summary(first_model)
+prior_summary(second_model)
 ```
@@
 ```{r}
-prior_summary(second_model)
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-area
+#| layout-ncol: 2
+#| fig-cap: "Examining how the model fits, and is affected by, the data"
+#| fig-subcap: ["First Model Posterior prediction check", "First Model Comparing the posterior with the prior","Second Model Posterior prediction check", "Second Model Comparing the posterior with the prior"]
+
+plot(
+  first_model,
+  "areas"
+)
+plot(
+  second_model,
+  "areas"
+)
 ```
@@
 plot(first_model, "rhat")
+
+plot(second_model, "trace")
+
+plot(second_model, "rhat")
 ```
 
-```{r}
-plot(
-  first_model,
-  "areas"
-)
-plot(
-  second_model,
-  "areas"
-)
-```
+# Idealized Methodology for A Survey-Based Qualitative Studies {#sec-survey}
 
-```{r}
-pp_check(second_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-```
+In our study, we examine the impact of abortion bans on infant mortality
+rates in the United States following the Dobbs decision, employing both
+observational data and survey-based research. Observational data provide
+quantitative insights, revealing trends and associations across large
+populations. However, they often fall short in addressing underlying
+mechanisms due to limitations such as confounding factors and challenges
+in establishing causality.
+
+To address these gaps, in this appendix we propose incorporating surveys to gather
+qualitative data, which provide rich, detailed accounts of individual
+experiences and contextual factors. Qualitative methods allow for a
+deeper understanding of how restrictive abortion policies affect
+maternal health, access to healthcare, and infant outcomes, capturing
+nuances that quantitative analysis alone cannot.
+
+By combining quantitative and qualitative approaches, this study
+utilizes the strengths of both methods. Quantitative data offer
+generalizable findings and statistical rigor, while qualitative data
+provide depth and context to interpret those findings. Together, they
+create a more comprehensive framework for understanding the complex
+relationship between restrictive abortion policies, maternal health, and
+infant outcomes, enabling the development of nuanced, evidence-based
+interventions and policy recommendations.
+
+## Introduction
+
+Restrictive abortion laws may exacerbate maternal stress, anxiety, or
+depression—especially in cases of unintended or high-risk
+pregnancies—potentially undermining maternal health and prenatal care.
+Research indicates that maternal mental health is a critical determinant
+of infant health outcomes, as conditions such as chronic stress,
+depression, or anxiety during pregnancy can negatively impact prenatal
+behaviors, access to healthcare, and overall maternal well-being. These
+psychological challenges, compounded by societal pressures and systemic
+barriers, can significantly increase the risk of adverse infant
+outcomes, including infant mortality. The survey will focus on mothers
+unable to access abortion services who subsequently experienced the
+death of their infant, documenting their experiences to better
+understand these interconnected factors.
+
+To provide a comprehensive perspective, the study will also involve
+healthcare providers and support organizations, capturing insights into
+systemic challenges and the broader impacts of abortion bans on family
+healthcare. By combining these perspectives with mothers’ lived
+experiences, the survey aims to inform evidence-based interventions
+addressing maternal psychological and healthcare needs and mitigating
+the negative consequences of restrictive abortion policies.
+
+This dual approach, integrating quantitative and qualitative methods,
+seeks to uncover the complex relationship between restrictive abortion
+laws, maternal mental health, and infant outcomes, forming the basis for
+effective policy recommendations.
+
+## Objective
+
+The objective of this study is to investigate the psychological, social,
+and systemic impacts of restrictive abortion laws on maternal health and
+infant outcomes. By focusing on mothers unable to access abortion
+services who experienced the death of their infant, the study aims to
+document lived experiences, explore the role of maternal stress and
+healthcare barriers in adverse outcomes, and gather perspectives from
+healthcare providers and support organizations. Understanding how
+maternal mental health directly affects prenatal care decisions, access
+to healthcare, and overall maternal behavior is essential to addressing
+the systemic factors contributing to higher infant mortality rates. The
+findings will guide the development of evidence-based strategies to
+improve maternal mental health, enhance healthcare access, and shape
+policy responses to restrictive abortion laws.
+
+## Sampling Approach
+
+In this analysis, we will use non-probability Respondent-driven sampling to finding respondents.
+Respondent-driven sampling (RDS), developed by @heckathorn1997, is ideal for studying hidden populations lacking a sampling frame and facing potential stigma or harm.
+Unlike snowball sampling, RDS offers dual compensation for participation and recruitment, with recruitment occurring peer-to-peer via social networks.
+This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics, such as abortion, by leveraging trust within communities.
+
+We use Respondent-Driven Sampling (RDS) for our survey because it is well-suited for reaching hidden or stigmatized populations, such as mothers who were unable to access abortion services and experienced the death of their infant.
+These individuals may be difficult to identify through conventional methods due to the sensitive nature of the topic and the lack of a clear sampling frame.
+
+RDS utilizes peer networks, allowing participants to recruit others they trust, fostering higher participation rates in populations that may distrust external researchers.
+This approach helps maintain confidentiality and reduces the risk of social stigma.
+Additionally, RDS provides a structured framework for statistical adjustments, mitigating biases inherent in network-based recruitment, and improving the representativeness of the sample.
+These advantages make RDS an effective and ethical choice for gathering insights into the psychological, social, and systemic impacts of restrictive abortion laws.
+
+## Target Population
+
+Our target population is mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision.
+
+## Sample frame
+
+Given the sensitive nature of our study, identifying and reaching the initial participants poses a significant challenge.
+To address this, we aim to collaborate with support groups and advocacy organizations like Abortion Rights Advocacy Groups.
+These groups play a critical role in connecting researchers with individuals who are otherwise difficult to reach, particularly for sensitive topics like the impact of restrictive abortion laws on maternal and infant health.
+
+Therefore, the sample frame of our study will be mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision and are reachable and can be contacted by Abortion Rights Advocacy Groups.
+
+## Sample
+
+We plan to survey 10 respondents for this study.
+Our sample will consist of mothers who fit the defined sample frame—those who sought but were unable to access abortion services due to restrictive laws, experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups.
+These mothers must also be willing to participate in and answer our survey questions, ensuring the reliability and depth of the collected data.
+
+## Ethical Concerns
+
+This study addresses sensitive and emotionally charged topics, necessitating a careful and ethical approach to protect participants’ well-being.
+We acknowledge the potential psychological distress that discussing experiences of unintended pregnancies, restrictive abortion laws, and infant loss may cause.
+To mitigate this, the survey will include trigger warnings, and participants will have the option to skip questions or withdraw at any time without penalty.
+All data collection will adhere to strict confidentiality protocols, ensuring that participants’ identities remain anonymous and their responses are securely stored.
+Recruitment through trusted advocacy groups will ensure that participants feel safe and supported throughout the process.
+Additionally, we will provide resources for mental health support to any participant who may experience discomfort as a result of participating in the study.
+This approach underscores our commitment to conducting ethical, respectful, and responsible research.
+
+## Proposed Survey Design
 

### 2024-11-24 scripts/06-model_data.R

@@ -11,27 +11,39 @@
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(arrow)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+first_model <- stan_glmer(
+  formula = death_rate ~ after_injunction * abortion_illegal + abortion_illegal + (1 | state) ,
+  data = analysis_data,
+  family = gaussian(),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_aux = exponential(rate = 1, autoscale = TRUE),
+  seed = 853
+)
 
 
+second_model <- stan_glmer(
+  formula = death_rate ~ after_injunction * abortion_illegal + year_of_death +age_of_mother+ mothers_single_race + (1 | state),
+  data = analysis_data,
+  family = gaussian(),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_aux = exponential(rate = 1, autoscale = TRUE),
+  seed = 853
+)
 #### Save model ####
 saveRDS(
   first_model,
   file = "models/first_model.rds"
 )
-
+saveRDS(
+  second_model,
+  file = "models/second_model.rds"
+)
 

### 2024-11-24 API/api.R
update

@@ -1,0 +1,50 @@
+id <- do_provision(example = FALSE)
+
+install_r_package(
+  droplet = id,
+  c(
+    "plumber",
+    "remotes",
+    "here"
+  )
+)
+
+debian_apt_get_install(
+  id,
+  "libssl-dev",
+  "libsodium-dev",
+  "libcurl4-openssl-dev"
+)
+
+debian_apt_get_install(
+  id,
+  "libxml2-dev"
+)
+
+install_r_package(
+  id,
+  c(
+    "config",
+    "httr",
+    "urltools",
+    "plumber"
+  )
+)
+
+install_r_package(id, c("xml2"))
+install_r_package(id, c("tidyverse"))
+install_r_package(id, c("tidymodels"))
+
+do_deploy_api(
+  droplet = id,
+  path = "example",
+  localPath = getwd(),
+  
+  
+  port = 8000,
+  docs = TRUE,
+  overwrite = TRUE
+)
+
+
+do_deploy_api(id, "infant-death", "~/Relationship between infant mortality rate and prohibited abortion", 8000,docs = TRUE,overwrite = TRUE)

### 2024-11-24 API/plumber.R

@@ -1,0 +1,46 @@
+library(plumber)
+library(tidyverse)
+
+model <- readRDS(file = here::here("models/first_model.rds"))
+
+version_number <- "0.0.1"
+
+variables <- list(
+  death_rate = "The number of deaths per 1,000 live births, a continuous variable.",
+  after_injunction = "A binary variable indicating whether the infant's death occurred after the implementation of abortion injunctions (1 = Yes, 0 = No).",
+  abortion_illegal = "A binary variable indicating the legality of abortion in the state of the mother's residence (1 = Illegal, 0 = Legal).",
+  state = "The state of the mother's residence, included as a random effect in the model."
+)
+
+#* Predict infant death rate based on inputs
+#* @param after_injunction Whether the death occurred after the injunction (0 = No, 1 = Yes)
+#* @param abortion_illegal Whether abortion is illegal in the state (0 = No, 1 = Yes)
+#* @param state The state of the mother's residence
+#* @get /predict_death_rate
+# Prediction function
+predict_infant_death_rate <- function(after_injunction = 0, abortion_illegal = 0, state = "unknown") {
+  # Create payload based on input
+  payload <- data.frame(
+    after_injunction = as.integer(after_injunction),
+    abortion_illegal = as.integer(abortion_illegal),
+    state = as.factor(state)
+  )
+  
+  # Ensure input state matches model levels
+  if (!state %in% levels(model$data$state)) {
+    stop("State not recognized. Please provide a valid state.")
+  }
+  
+  # Make prediction
+  prediction <- predict(first_model, payload)
+  
+  # Return results
+  result <- list(
+    input = list(payload),
+    response = list("estimated_death_rate" = prediction),
+    status = 200,
+    model_version = "1.0.0"
+  )
+  
+  return(result)
+}

### 2024-11-24 API/server.R

@@ -1,0 +1,4 @@
+library(plumber)
+serve_model <- plumb("plumber.R")
+serve_model$run(port = 8000)
+

### 2024-11-24 API/api.R
update

@@ -1,50 +1,0 @@
-id <- do_provision(example = FALSE)
-
-install_r_package(
-  droplet = id,
-  c(
-    "plumber",
-    "remotes",
-    "here"
-  )
-)
-
-debian_apt_get_install(
-  id,
-  "libssl-dev",
-  "libsodium-dev",
-  "libcurl4-openssl-dev"
-)
-
-debian_apt_get_install(
-  id,
-  "libxml2-dev"
-)
-
-install_r_package(
-  id,
-  c(
-    "config",
-    "httr",
-    "urltools",
-    "plumber"
-  )
-)
-
-install_r_package(id, c("xml2"))
-install_r_package(id, c("tidyverse"))
-install_r_package(id, c("tidymodels"))
-
-do_deploy_api(
-  droplet = id,
-  path = "example",
-  localPath = getwd(),
-  
-  
-  port = 8000,
-  docs = TRUE,
-  overwrite = TRUE
-)
-
-
-do_deploy_api(id, "infant-death", "~/Relationship between infant mortality rate and prohibited abortion", 8000,docs = TRUE,overwrite = TRUE)

### 2024-11-24 API/plumber.R

@@ -1,46 +1,0 @@
-library(plumber)
-library(tidyverse)
-
-model <- readRDS(file = here::here("models/first_model.rds"))
-
-version_number <- "0.0.1"
-
-variables <- list(
-  death_rate = "The number of deaths per 1,000 live births, a continuous variable.",
-  after_injunction = "A binary variable indicating whether the infant's death occurred after the implementation of abortion injunctions (1 = Yes, 0 = No).",
-  abortion_illegal = "A binary variable indicating the legality of abortion in the state of the mother's residence (1 = Illegal, 0 = Legal).",
-  state = "The state of the mother's residence, included as a random effect in the model."
-)
-
-#* Predict infant death rate based on inputs
-#* @param after_injunction Whether the death occurred after the injunction (0 = No, 1 = Yes)
-#* @param abortion_illegal Whether abortion is illegal in the state (0 = No, 1 = Yes)
-#* @param state The state of the mother's residence
-#* @get /predict_death_rate
-# Prediction function
-predict_infant_death_rate <- function(after_injunction = 0, abortion_illegal = 0, state = "unknown") {
-  # Create payload based on input
-  payload <- data.frame(
-    after_injunction = as.integer(after_injunction),
-    abortion_illegal = as.integer(abortion_illegal),
-    state = as.factor(state)
-  )
-  
-  # Ensure input state matches model levels
-  if (!state %in% levels(model$data$state)) {
-    stop("State not recognized. Please provide a valid state.")
-  }
-  
-  # Make prediction
-  prediction <- predict(first_model, payload)
-  
-  # Return results
-  result <- list(
-    input = list(payload),
-    response = list("estimated_death_rate" = prediction),
-    status = 200,
-    model_version = "1.0.0"
-  )
-  
-  return(result)
-}

### 2024-11-24 API/server.R

@@ -1,4 +1,0 @@
-library(plumber)
-serve_model <- plumb("plumber.R")
-serve_model$run(port = 8000)
-

### 2024-11-24 paper/paper.qmd
update paper

@@
 @sec-model explains the modeling approach, including the assumptions and
-specifications of our linear regression framework. In @sec-result, we
+specifications of our linear regression framework. In @sec-results, we
 present our findings, emphasizing the key predictors of Harris's
@@
 [Appendix- @sec-data-details] provides external data detail, [Appendix-
-@sec-model-details] provides model detail, @sec-c provides an datasheet,
-and @sec-d provides an detailed introduction of surveys, sampling and
+@sec-model-details] provides model detail, and [Appendix-@sec-survey] provides an detailed introduction of surveys, sampling and
 observational data in our paper.
@@
 
+# Model {#sec-model}
+
 ## Difference-In-Difference Approach (DID)
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-11-24 scripts/02-download_data.R
delete download data

@@ -8,18 +8,4 @@
 # Any other information needed? [...UPDATE THIS...]
 
 
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
-
-#### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
-
-
-
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
-
-         
+Accroding to CDC Wonder's privacy

### 2024-11-24 scripts/02-download_data.R
delete download file

@@ -1,11 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-Accroding to CDC Wonder's privacy

### 2024-11-24 scripts/00-simulate_data.R
update scripts

@@ -1,52 +1,51 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulates a dataset of infant death rate across US in 2021 and 2022.
+# Author: Diana Shen
+# Date: 24 November 2024 
+# Contact: diana.shen@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 #### Workspace setup ####
 library(tidyverse)
-set.seed(853)
 
+#### Data expected structure ####
+# Simulated Data infant death rate:
+# - Columns: death_rate, after_injunction, abortion_illegal, state, age_of_mother, mothers_single_race, year_of_death, month
+#death_rate = "numeric, between 0 and 50 (representing deaths per 1,000 live births)"
+#after_injunction = "binary, 0 (before injunction) or 1 (after injunction)"
+#abortion_illegal = "binary, 0 (legal abortion) or 1 (illegal abortion)"
+#state = "string, valid US state abbreviations (e.g., TX, CA)"
+#age_of_mother = "categorical, one of: '15–19', '20–24', '25–29', '30–34', '35–39', '40–44', '45+'"
+#mothers_single_race = "categorical, one of: 'White', 'Black', 'Asian', 'Native American', 'Pacific Islander', 'More than #one race', 'Unknown or Not Stated'"
+#year_of_death = "numeric, one of: 2021, 2022"
+#month = "numeric, between 1 (January) and 12 (December)"
 
 #### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
-)
+# Define states, candidates, and other factors
+# Define the list of states where abortion is illegal
+illegal_states <- c("AL", "AR", "ID", "IN", "KY", "LA", "MS", "OK", "SD", "TN", "TX", "WV")
 
-# Political parties
[…]

### 2024-11-24 scripts/01-test_simulated_data.R

@@ -1,89 +1,125 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the infant death rate across US in 2021 and 2022
+# simulated dataset.
+# Author: Diana Shen
+# Date: 24 November 2024 
+# Contact: diana.shen@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# - The `tidyverse` package must be installed and loaded
+# - 00-simulate_data.R must have been run
+# Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 #### Workspace setup ####
 library(tidyverse)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+#### Define the list of states where abortion is illegal ####
+illegal_states <- c("AL", "AR", "ID", "IN", "KY", "LA", "MS", "OK", "SD", "TN", "TX", "WV")
+#### Load data ####
+simulated_data <- read_csv("data/00-simulated_data/simulated_data.csv")
 
 # Test if the data was successfully loaded
-if (exists("analysis_data")) {
+if (exists("simulated_data")) {
   message("Test Passed: The dataset was successfully loaded.")
 } else {
   stop("Test Failed: The dataset could not be loaded.")
 }
 
-
 #### Test data ####
 
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
+# Check if the dataset has 700 rows (expected based on simulation)
+if (nrow(simulated_data) == 700) {
+  message("Test Passed: The dataset has 700 rows.")
 } else {
-  stop("Test Failed: The dataset does not have 151 rows.")
+  stop(paste("Test Failed: The dataset has", nrow(simulated_data), "rows instead of 700."))
 }
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
+# Check if the dataset has the expected 8 columns
+expected_columns <- c("death_rate", "state", "age_of_mother", "mothers_single_race", 
[…]

### 2024-11-24 scripts/02-clean_data.R

@@ -1,11 +1,13 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw dataset of infant death rate across US in 2021 and 2022.
+# Author: Diana Shen
+# Date: 24 November 2024
+# Contact: diana.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites:  
+# - The `tidyverse`, `janitor`,`lubridate`,`arrow` packages must be installed and loaded
+# - 02-download_data.R must have been run
+# Any other information needed?  Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-11-24 scripts/03-test_analysis_data.R

@@ -1,69 +1,162 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests the structure and validity of the cleaned infant death rate across US in 2021 and 2022
+# analysis dataset.
+# Author: Diana Shen
+# Date: 24 November 2024 
+# Contact: diana.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# - 03-clean_data.R must have been run
+# Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 #### Workspace setup ####
 library(tidyverse)
-library(testthat)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
+analysis_data <- read_csv(
+  file = "data/02-analysis_data/analysis_data.csv",
+  col_types = cols(
+    death_rate = col_double(),
+    after_injunction = col_factor(levels = c("0", "1")),
+    abortion_illegal = col_factor(levels = c("0", "1")),
+    state = col_character(),
+    age_of_mother = col_factor(),
+    mothers_single_race = col_character(),
+    year_of_death = col_integer(),
+    month = col_factor(),
+    month_code=col_integer()
+  )
+)
 
+#### Define the list of states where abortion is illegal ####
+illegal_states <- c("Alabama", "Arkansas", "Idaho", "Indiana", "Kentucky", 
+                    "Louisiana", "Mississippi", "Oklahoma", "South Dakota", 
+                    "Tennessee", "Texas", "West Virginia")
+state =c(
+  "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", 
+  "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", 
+  "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", 
[…]

### 2024-11-24 scripts/05-install_packages.R

@@ -1,0 +1,20 @@
+#### Preamble ####
+# Purpose: Install necessary packages.
+# Author: Diana Shen
+# Date: 24 November 2024
+# Contact: diana.shen@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+# Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
+
+# Install necessary packages
+install.packages("tidyverse")
+install.packages("arrow")
+install.packages("knitr")
+install.packages("usmap")
+install.packages("ggplot2")
+install.packages("dplyr")
+install.packages("kableExtra")
+install.packages("rstanarm")
+install.packages("janitor")
+install.packages("lubridate")

### 2024-11-25 paper/paper.qmd
update paper

@@
 library(knitr)
+library(usmap)
+library(ggplot2)
+library(dplyr)
+library(kableExtra)
+library(rstanarm)
 ```
@@
 [Appendix- @sec-data-details] provides external data detail, [Appendix-
-@sec-model-details] provides model detail, and [Appendix-@sec-survey] provides an detailed introduction of surveys, sampling and
-observational data in our paper.
+@sec-model-details] provides model detail, and [Appendix- @sec-survey]
+provides an detailed introduction of surveys, sampling and observational
+data in our paper.
 
@@
 account for potential confounding factors.
+
+In this study, we utilized several R packages to streamline data
+manipulation, modeling, and visualization. The `tidyverse` package
+@tidyverse was central to our workflow, providing tools for data
+wrangling and analysis, while `arrow` @arrow efficiently managed parquet
+files for larger datasets. We used `modelsummary` package @modelsummary
+to generate clean and interpretable summaries of our regression
+models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
+clean tables and reproducible reports, and `usmap` for geographical
+visualizations. We use `here` @here package simplifies the management of
+file paths, ensuring a reproducible workflow by creating a consistent
+point of reference for file locations, regardless of the operating
+system or working directory.Data cleaning was facilitated by `janitor`
+@janitor, while `lubridate` @lubridate handled date-time variables.
+Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
+regression modeling was conducted using `rstanarm` @rstanarm. Together,
+these packages supported an efficient, reproducible workflow for
+analyzing and presenting our findings.
 
@@
 the health implications of varying abortion laws across the U.S. By
-leveraging comprehensive datasets and robust statistical techniques,
+utilizing comprehensive datasets and robust statistical techniques,
 this study offers a nuanced understanding of the intersection between
@@
 #| message: false
-
-library(usmap)
-library(ggplot2)
-library(dplyr)
 
@@
 #| message: false
-library(kableExtra)
 analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
@@
 #| message: false
-library(kableExtra)
+#| 
 analysis_data_state <- analysis_data|>
@@
 
-library(ggnewscale)
 death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
@@
 show the highest rates, while states such as California and New York
-exhibit lower rates. \# Model
+exhibit lower rates.
+
+# Model {#sec-model}
 
@@
 effects.
-
-# Model {#sec-model}
 
@@
 #| include: false
-
-library(rstanarm)
-
 first_model <-
@@
 
-To address these gaps, in this appendix we propose incorporating surveys to gather
-qualitative data, which provide rich, detailed accounts of individual
-experiences and contextual factors. Qualitative methods allow for a
-deeper understanding of how restrictive abortion policies affect
+To address these gaps, in this appendix we propose incorporating surveys
+to gather qualitative data, which provide rich, detailed accounts of
+individual experiences and contextual factors. Qualitative methods allow
+for a deeper understanding of how restrictive abortion policies affect
 maternal health, access to healthcare, and infant outcomes, capturing
@@
 
-In this analysis, we will use non-probability Respondent-driven sampling to finding respondents.
-Respondent-driven sampling (RDS), developed by @heckathorn1997, is ideal for studying hidden populations lacking a sampling frame and facing potential stigma or harm.
-Unlike snowball sampling, RDS offers dual compensation for participation and recruitment, with recruitment occurring peer-to-peer via social networks.
-This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics, such as abortion, by leveraging trust within communities.
-
-We use Respondent-Driven Sampling (RDS) for our survey because it is well-suited for reaching hidden or stigmatized populations, such as mothers who were unable to access abortion services and experienced the death of their infant.
-These individuals may be difficult to identify through conventional methods due to the sensitive nature of the topic and the lack of a clear sampling frame.
-
-RDS utilizes peer networks, allowing participants to recruit others they trust, fostering higher participation rates in populations that may distrust external researchers.
-This approach helps maintain confidentiality and reduces the risk of social stigma.
-Additionally, RDS provides a structured framework for statistical adjustments, mitigating biases inherent in network-based recruitment, and improving the representativeness of the sample.
-These advantages make RDS an effective and ethical choice for gathering insights into the psychological, social, and systemic impacts of restrictive abortion laws.
+In this analysis, we will use non-probability Respondent-driven sampling
+to finding respondents. Respondent-driven sampling (RDS), developed by
+@heckathorn1997, is ideal for studying hidden populations lacking a
+sampling frame and facing potential stigma or harm. Unlike snowball
+sampling, RDS offers dual compensation for participation and
+recruitment, with recruitment occurring peer-to-peer via social
+networks. This method maintains confidentiality, reduces bias through
+statistical adjustments, and is particularly effective for sensitive
+topics, such as abortion, by utilizing trust within communities.
+
+We use Respondent-Driven Sampling (RDS) for our survey because it is
+well-suited for reaching hidden or stigmatized populations, such as
+mothers who were unable to access abortion services and experienced the
+death of their infant. These individuals may be difficult to identify
+through conventional methods due to the sensitive nature of the topic
+and the lack of a clear sampling frame.
+
+RDS utilizes peer networks, allowing participants to recruit others they
+trust, fostering higher participation rates in populations that may
+distrust external researchers. This approach helps maintain
+confidentiality and reduces the risk of social stigma. Additionally, RDS
+provides a structured framework for statistical adjustments, mitigating
+biases inherent in network-based recruitment, and improving the
+representativeness of the sample. These advantages make RDS an effective
+and ethical choice for gathering insights into the psychological,
+social, and systemic impacts of restrictive abortion laws.
 
@@
 
-Our target population is mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision.
+Our target population is mothers who sought but were unable to access
+abortion services due to restrictive laws and experienced the death of
+their infant within the first year of life in states with abortion bans
+following the Dobbs decision.
 
@@
 
-Given the sensitive nature of our study, identifying and reaching the initial participants poses a significant challenge.
-To address this, we aim to collaborate with support groups and advocacy organizations like Abortion Rights Advocacy Groups.
-These groups play a critical role in connecting researchers with individuals who are otherwise difficult to reach, particularly for sensitive topics like the impact of restrictive abortion laws on maternal and infant health.
-
-Therefore, the sample frame of our study will be mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision and are reachable and can be contacted by Abortion Rights Advocacy Groups.
+Given the sensitive nature of our study, identifying and reaching the
+initial participants poses a significant challenge. To address this, we
+aim to collaborate with support groups and advocacy organizations like
+Abortion Rights Advocacy Groups. These groups play a critical role in
+connecting researchers with individuals who are otherwise difficult to
+reach, particularly for sensitive topics like the impact of restrictive
+abortion laws on maternal and infant health.
+
+Therefore, the sample frame of our study will be mothers who live in
+areas where they are unable to access abortion services due to
+restrictive laws and have experienced the death of their infant within
+the first year of life in states with abortion bans following the Dobbs
+decision and are reachable and can be contacted by Abortion Rights
+Advocacy Groups.
 
@@
 
-We plan to survey 10 respondents for this study.
-Our sample will consist of mothers who fit the defined sample frame—those who sought but were unable to access abortion services due to restrictive laws, experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups.
-These mothers must also be willing to participate in and answer our survey questions, ensuring the reliability and depth of the collected data.
+We plan to survey 20 respondents for this study. Our sample will consist
+of mothers who fit the defined sample frame—those who were unable to
+access abortion services due to restrictive laws, experienced the death
+of their infant within the first year of life, and are reachable through
+Abortion Rights Advocacy Groups. These mothers must also be willing to
+participate in and answer our survey questions, ensuring the reliability
+and depth of the collected data.
+
+## Recruitment of Respondents
+
+To recruit participants for this sensitive study, we will use a Respondent-Driven Sampling (RDS) approach in collaboration with  Abortion Rights Advocacy Groups, that have established relationships with our target population.
+These organizations will serve as initial “seeds” to identify and reach mothers affected by restrictive abortion laws who meet our study criteria.
+Outreach materials emphasizing confidentiality and the study’s importance will be shared through trusted channels.
+As part of the RDS methodology, participants may also invite others from their network who share similar experiences, ensuring the sample expands while maintaining anonymity.
+Interested participants will undergo a screening process and provide informed consent, with the option to complete the survey anonymously online, ensuring safety and confidentiality.
+To support participant well-being, we will provide access to mental health resources throughout the process, fostering trust and encouraging meaningful participation.
+
+## Handling Non-response bias
+
+Non-response bias is a concern as it can distort survey results and lead to unrepresentative conclusions about the target population.
+Since our survey addresses a sensitive topic and takes approximately 10–15 minutes to complete, there is a risk of participant dropout or refusal.
+To minimize this, we provide a clear overview of the survey’s purpose, confidentiality measures, and estimated completion time, fostering trust and encouraging participation.
+
+## Respondent Validation
+
+To ensure the credibility of our data, we implement a respondent validation process.
+Participants will undergo an initial screening to confirm they meet the study criteria, such as being unable to access abortion services due to restrictive laws and experiencing the loss of an infant within the first year of life.
+Responses will be reviewed for completeness and consistency, and duplicate entries will be removed to maintain data integrity.
+Collaboration with trusted support groups and advocacy organizations further enhances respondent reliability, ensuring that our findings accurately reflect the experiences of the target population.
 
@@
 
-This study addresses sensitive and emotionally charged topics, necessitating a careful and ethical approach to protect participants’ well-being.
-We acknowledge the potential psychological distress that discussing experiences of unintended pregnancies, restrictive abortion laws, and infant loss may cause.
-To mitigate this, the survey will include trigger warnings, and participants will have the option to skip questions or withdraw at any time without penalty.
-All data collection will adhere to strict confidentiality protocols, ensuring that participants’ identities remain anonymous and their responses are securely stored.
-Recruitment through trusted advocacy groups will ensure that participants feel safe and supported throughout the process.
-Additionally, we will provide resources for mental health support to any participant who may experience discomfort as a result of participating in the study.
-This approach underscores our commitment to conducting ethical, respectful, and responsible research.
+This study addresses sensitive and emotionally charged topics,
+necessitating a careful and ethical approach to protect participants’
+well-being. We acknowledge the potential psychological distress that
+discussing experiences of unintended pregnancies, restrictive abortion
+laws, and infant loss may cause. To mitigate this, the survey will
+include trigger warnings, and participants will have the option to skip
+questions or withdraw at any time without penalty. All data collection
+will adhere to strict confidentiality protocols, ensuring that
+participants’ identities remain anonymous and their responses are
+securely stored. Recruitment through trusted advocacy groups will ensure
+that participants feel safe and supported throughout the process.
+Additionally, we will provide resources for mental health support to any
+participant who may experience discomfort as a result of participating
+in the study. This approach underscores our commitment to conducting
+ethical, respectful, and responsible research.
 
@@
 
+Abortion, or even considering abortion, carries significant social
+stigma. Therefore, our survey requires extremely careful phrasing and
+design to minimize this stigma’s impact on the study of abortion,
+pregnancy, fertility, and related demographic and health outcomes. In
+@lindberg2022, researchers tested innovative methods in a national
+survey to address persistent underreporting of abortion in the U.S.,
+focusing on question design and placement to reduce sensitivity, social
+desirability bias, and perceived intrusiveness. Although none of their
+experimental approaches significantly improved abortion reporting
+compared to the control condition, their study offers valuable insights
+for improving abortion measurement. We aim to adapt our survey design
+based on their methodologies to enhance data accuracy and respondent
+comfort.
+
+This survey explores the impacts of restrictive abortion laws on maternal mental health and infant outcomes, emphasizing neutral, carefully crafted questions to ensure participant comfort and minimize social desirability bias.
+Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., “Prefer not to say”) to address underreporting in sensitive topics.
+Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with nuanced insights, fostering trust and enhancing data accuracy.
+Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each sectionof our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.
+
+## Solution to the response bias in our survey
+
+We draw on recommendations from @stantcheva2023 to minimize response biases.
+Common response biases identified in survey design include moderacybias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB).
+Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB.The detailed definitions of these biases are provided in [Appendix-@sec-definition].
+
+To mitigate bias, we enhance our survey in the following ways, drawing on recommendations
+from @stantcheva2023:
+
+- Addressing Extreme/Moderacy Bias: We use a minimum of five response options for scale questions to provide more nuanced choices, reducing the likelihood of respondents defaulting to extreme or middle answers.
+
+- Mitigating Response Order Bias: For nominal questions, we randomize response options.
+
+- Minimizing Social Desirability Bias (SDB): Our online survey is carefully designed to minimize social desirability bias (SDB) while addressing the sensitivity of the topic.
+The survey begins with a clear introduction explaining its purpose—academic research on the impact of restrictive abortion laws—while emphasizing respondent anonymity and confidentiality.
+Before sensitive questions, participants are reassured that their responses are entirely confidential and will only be used for research purposes.
+The anonymous online format provides a safe space for participants to share their experiences without fear of judgment or stigma.
+Additionally, a feedback section at the end invites respondents to voice any concerns or share additional thoughts, fostering trust and enhancing the quality of the data collected.
+
+### Survey Link
+
+The survey has been implemented using Google Forms. You can access it
+here: [Survey Link](https://forms.gle/YRrb6Z6EyorHPkuF8).
+
+## Copy of Survey on Restrictive Abortion Laws and Maternal Health
+
+Welcome Section
+
+Introduction: Welcome to our study on the impact of restrictive abortion
+laws on maternal and infant health. Your participation in this survey
+will help us understand the psychological, social, and systemic impacts
+of these laws. Rest assured that your responses are anonymous and will
+only be used for academic research purposes.
+
+This survey is conducted by nonpartisan researchers in public health and
+social sciences. It consists of 17 carefully designed questions and
+should take approximately 10–15 minutes to complete.
+
+Please answer the questions honestly. If you experience any discomfort
+while completing the survey, you may stop at any time. For support, we
+provide a list of mental health resources at the end of the survey.
+
+Contact Information: Diana Shen Email: diana.shen\@mail.utoronto.ca
+
+Section 1: Demographics and Background Information
+
+1.  What is your age?
+    -   Under 18
+    -   18–24
+    -   25–34
+    -   35–44
+    -   45–54
+    -   55+
+    -   Prefer not to say
+2.  What is your highest level of education?
+    -   Less than high school
+    -   High school graduate or equivalent
+    -   Some college
+    -   Bachelor’s degree
+    -   Graduate or professional degree
+    -   Prefer not to say
+
+3.What is your marital status? - Single - Married - Divorced - Widowed -
+Prefer not to say
+
+Section 2: Understanding Abortion Experiences
+
+Introduction: This section focuses on understanding abortion
+experiences, including the circumstances and decisions surrounding them.
+Your responses are invaluable in helping researchers and policymakers
+improve health services and support for women and families. Please know
+that your answers are entirely confidential and will only be used for
+research purposes. If you are comfortable, we encourage you to answer as
+honestly as possible. If you prefer not to answer, you are welcome to
+skip this section.
+
+1.  Are you willing to answer this part?
+    -   Yes
+    -   No \[Jump to Section 3\]
+2.  Did you seek abortion services during your pregnancy?
+    -   Yes
+    -   No
+    -   Prefer not to say
+3.  If yes, were you unable to access abortion services due to legal
+    restrictions in your state?
+    -   Yes
+    -   No
+    -   Prefer not to say
+4.  How did the inability to access abortion services impact your mental
+    health during pregnancy?
+    -   No impact
+    -   Mild impact
+    -   Moderate impact
+    -   Severe impact
+    -   Prefer not to say
+5.  Did you receive any support from healthcare providers or community
+    organizations during your pregnancy?
+    -   Yes
+    -   No
+6.  If you were unable to access abortion services, what barriers did
+    you encounter? (Select all that apply)
+    -   Legal restrictions
+    -   Financial constraints
+    -   Lack of healthcare providers
+    -   Distance to clinic
+    -   Fear of stigma or judgment
+    -   Prefer not to say
+7.  If you sought an abortion but were unable to access one, how did
+    this affect your mental health during pregnancy?
+    -   Increased stress
+    -   Anxiety
+    -   Depression
+    -   Anger or frustration
+    -   Feeling of helplessness
+    -   Prefer not to say
+
+Section 3: Experiences and Support for Women Facing Pregnancy Challenges
+
+Introduction: This section seeks to understand the experiences and
+outcomes of women who, due to restrictive abortion laws, were unable to
+access abortion services and subsequently faced the loss of their child.
+We recognize that discussing past pregnancies can be sensitive,
+especially those involving circumstances such as abortion or the loss of
+a child. Your insights are invaluable in identifying areas where
+healthcare and support services can be improved. If you prefer not to
+answer, you are welcome to skip this section.
+
+1.  Are you willing to answer this part?
+    -   Yes
+    -   No \[Jump to Section 4\]
+2.  Did you experience the loss of your child within the first year of
+    life?
+    -   Yes
+    -   No \[Jump to Section 4\]
+    -   Prefer not to say
+3.  What was the primary cause of your child’s death as communicated by
+    healthcare providers?
+    -   Premature birth or related complications
+    -   Congenital abnormalities or genetic conditions
+    -   Sudden Infant Death Syndrome (SIDS)
+    -   Infections (e.g., pneumonia, sepsis)
+    -   Birth trauma or delivery complications
+    -   Lack of access to timely medical care
+    -   Other (please specify)
+4.  To what extent do you believe the instances during your pregnancy
+    where mental health challenges prevented you from seeking or
+    receiving adequate medical care?
+    -   1 (No impact)
+    -   2
+    -   3
+    -   4
+    -   5 (Significantly)
+5.  To what extent do you believe the mental health challenges caused by
+    the inability to access abortion services contributed to health
+    complications for your child?
+    -   Not at all – I don’t believe my mental health challenges had any
+        impact on my child’s health.
+    -   Slightly – I think there may have been a minor impact on my
+        child’s health.
+    -   Moderately – I feel my mental health challenges had a noticeable
+        impact on my child’s health.
+    -   Significantly – I believe my mental health challenges had a
+        considerable impact on my child’s health.
+    -   Completely – I think my mental health challenges were the
+        primary factor in my child’s health complications.
+    -   Prefer not to say
+6.  Looking back, do you believe access to abortion services could have
+    positively affected your mental and physical health during
+    pregnancy?
+    -   Yes
+    -   No
+    -   Maybe
+    -   Prefer not to say
+
+Section 4: Perspectives and Support
+
+If you feel distressed or need support after completing this survey, the
+following resources are available to provide assistance:
+
+1.  Postpartum Support International (PSI)
+    -   Website: www.postpartum.net
+    -   Helpline: 1-800-944-4773 (Text “Help” to 800-944-4773)
+    -   Services: Support for mental health during and after pregnancy,
+        including peer support and counseling.
+2.  Mental Health America (MHA)
+    -   Website: www.mhanational.org
+    -   Services: Online screening tools, support networks, and
+        educational materials.
+3.  SAMHSA National Helpline
+    -   Website: www.samhsa.gov
+    -   Hotline: 1-800-662-HELP (4357)
+    -   Services: Free, confidential referrals for mental health and
+        substance use disorders.
+
+Section 5: Feedback
+
+1.  Do you have any concerns or feedback regarding the survey, surveyor,
+    or entity?
+    -   Your feedback is important to us and will help ensure
+        transparency and trust in the research process.
+
+Section 6: Thank You
+
+Thank you for taking the time to complete this survey. Your honest
+feedback is invaluable and will help us better understand and address
+the experiences of women who have faced similar circumstances. We deeply
+appreciate your participation and the courage it takes to share your
+experiences.
+
+## Response bias definition {#sec-definition}
+
+In the design of the questionnaire, there will be some common biases that
+may occur when running the questionnaire.
+
+@stantcheva2023 defines these biases as:
+
+- **Moderacy response bias**: The tendency to choose mid-scale responses.
+
+- **Extreme response bias**: The inclination to select extreme values on the scale.
+
+- **Response order bias**: Occurs when the position of options influences responses, with the **primacy effect** (favoring early options in written surveys) and **recency effect** (favoring later options in oral surveys) as common forms.
+
+- **Social desirability bias**: Respondents may hide true views to present a socially favorable image, influenced by topic and social context.
+
+- **Acquiescence bias**: The habit of giving positive responses, such as consistently choosing “agree” or “yes.”
 \newpage

### 2024-11-25 paper/paper.qmd
update paper

@@
 #| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| tbl-cap: "Summary of Bayesian Regression Model Examining the Impact of Abortion Bans on Infant Mortality Rates in the United States: A Difference-in-Differences Approach Using 2021-2022 Infant Mortality Data"
 #| warning: false
@@
   list(first_model,second_model),
-  fmt = 3,coef_omit="state",notes = "Note. All models include a random variable for state. The reference level of age_of_mother is 15-19.The reference level of monthers_single_race is Asian.")
+  fmt = 3,coef_omit="state",notes = "Note.
+All models include a random variable for state.
+The reference level of age_of_mother is 15-19.The reference level of monthers_single_race is Asian.
+Source: CDC Wonder")
 ```
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
-check for the first model. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
-prior. This shows...
+In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model.
+It represent posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data.
+The posterior predictive checks for Model 1 (policy-focused) and Model 2 (policy and demographic factors) illustrate model fit.
+In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision.
+@fig-ppcheckandposteriorvsprior-2 shows improved alignment, indicating that adding demographic factors enhances predictive accuracy and better captures the complexities of infant mortality rates.
+
+In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior.
+The posterior vs.
+prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters.
+In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power.
+Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
+The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a crucial role in understanding infant mortality in the context of restrictive abortion laws.
 
@@
 #| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["First Model Posterior prediction check", "First Model Comparing the posterior with the prior","Second Model Posterior prediction check", "Second Model Comparing the posterior with the prior"]
+#| fig-cap: "Examining how the model for infant death rate fits, and is affected by, the data"
+#| fig-subcap: ["First Model Posterior prediction check","Second Model Posterior prediction check", "First Model Comparing the posterior with the prior", "Second Model Comparing the posterior with the prior"]
 
@@
 ```
-
+The prior summaries for the two models, `first_model` and `second_model`, indicate that both use weakly informative priors.
+For the intercept, a normal prior with a mean of 0 and a standard deviation of 2.5 was specified, which adjusts to a scale of 12 after centering the predictors.
+The coefficients for both models are also assigned normal priors with a mean of 0 and a scale of 2.5, which are adjusted based on the predictors, resulting in scales such as [26.84, 28.43, 45.17] for the first model and [26.84, 28.43, 24.78,...] for the second model.
+The auxiliary parameter (sigma) follows an exponential prior with a rate of 1, adjusted to 0.2, while a decov prior regularizes the covariance structure in both models.
+Although the prior structure is similar, the second model incorporates more predictors, leading to a larger number of coefficients with slightly adjusted scales.
+These priors are designed to be weakly informative, allowing the data to primarily guide the posterior estimates while avoiding extreme parameter values.
+
+## Distribution
 ```{r}
@@
 #| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["First Model Posterior prediction check", "First Model Comparing the posterior with the prior","Second Model Posterior prediction check", "Second Model Comparing the posterior with the prior"]
+#| fig-cap: "95% credibility intervals for fitted models"
+#| fig-subcap: ["First Model", "Second Model"]
 
@@
 ```
+Credibility intervals are the Bayesian equivalent of confidence intervals, offering insight into the range of probable values for a parameter @tellingstories.
+Using Bayesian estimation, we derive a posterior distribution for each model coefficient.
+@fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2.
+These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.
 
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+@fig-stanareyouokay-1 is a trace plot for model 1. It shows... This suggests...
 
@@
 #| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
+#| fig-cap: "Checking the convergence of the MCMC algorithm for the infant death rate model"
+#| fig-subcap: ["First Model Trace plot", "First Model Rhat plot","Second Model Trace plot", "Second Model Rhat plot"]
 #| layout-ncol: 2
@@
 ```
+
+The trace plot in @fig-stanareyouokay-1 shows the sampling chains for model 1, indicating good mixing and convergence as the chains overlap and appear stable throughout.
+This suggests that the model has achieved convergence during sampling.
+The Rhat plot in @fig-stanareyouokay-2 demonstrates that all Rhat values are below 1.05, further confirming convergence and indicating that the model’s posterior samples are reliable.
+@fig-stanareyouokay-3 represents the trace plots for Model 2, showing the posterior samples across four chains.
+The plots indicate good mixing, as the chains overlap well and stabilize around the same region, suggesting convergence.
+@fig-stanareyouokay-4 is the Rhat diagnostic plot for Model 2.
+All Rhat values are close to 1.00, confirming that the Markov chains have converged, and the posterior estimates are reliable.
+This supports the validity of Model 2’s Bayesian estimation process.
 

### 2024-11-27 paper/paper.qmd
fix grammar error

@@
 
-Instead of examining the impact of abortion laws over broader time
+Instead of examining the impact of abortion laws over a broader time
 periods like prior research, our study focuses on the effect of abortion
@@
 outcomes. The study utilizes observational data from the Centers for
-Disease Control and Prevention’s WONDER database, capturing infant
+Disease Control and Prevention’s WONDER database, captures infant
 mortality trends across states with varying levels of abortion
@@
 [Appendix- @sec-data-details] provides external data detail, [Appendix-
-@sec-model-details] provides model detail, and [Appendix- @sec-survey]
-provides an detailed introduction of surveys, sampling and observational
+@sec-model-details] provides model details, and [Appendix- @sec-survey]
+provides a detailed introduction to surveys, sampling, and observational
 data in our paper.
@@
 We conduct our infant death rate observational data analysis using the R
-programming language [@citeR]. Our dataset, obtained from CDC
+programming language [@citeR]. Our dataset, obtained from the CDC
 Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a
@@
 point of reference for file locations, regardless of the operating
-system or working directory.Data cleaning was facilitated by `janitor`
+system or working directory. Data cleaning was facilitated by `janitor`
 @janitor, while `lubridate` @lubridate handled date-time variables.
@@
 this study offers a nuanced understanding of the intersection between
-policy, demographics, and infant health outcomes, informing public
+policy, demographics, and infant health outcomes, informing the public
 health strategies and legislative decisions.
@@
 #| label: fig-deathrate
-#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants.The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
+#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants.
+The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
 #| echo: false
@@
 `abortion_illegal` and `after_injunction`. The resulting interaction
-term equals 1 when abortion is illegal and the time is after the
+term equals 1 when abortion is illegal the time is after the
 injunction, and 0 otherwise. This term is included in the analysis to
@@
 
-The Age of Mother (`age_of_mother`) refers to the age group of the
-mother at the time of childbirth.In our observational data, age is
+The Age of the Mother (`age_of_mother`) refers to the age group of the
+mother at the time of childbirth. In our observational data, age is
 categorized into the brackets 15–19 years, 20–24 years, 25–29 years, and
@@
 #| label: tbl-motherage
-#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother.
+#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers.
 It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
@@
 
-The self-identified race of the mother, recorded as a single race
+The self-identified race of the mother was recorded as a single race
 category. In this dataset, the available categories are: Asian, Black or
@@
 #| label: tbl-motherrace
-#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother.
+#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers.
 It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
@@
 States (per 1,000 Live Births), December 2022.
-Infant Mortality Rate is much higher iin Southeastern.
+The infant Mortality Rate is much higher in Southeastern.
 Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
@@
 
-@fig-choromap is a choropleth map visualizes the average infant death
+@fig-choromap is a choropleth map that visualizes the average infant death
 rate per 1,000 live births across U.S. states, with data categorized by
@@
 For our analysis, we use a Bayesian Difference-in-Differences (DID)
-model to quantify the impact of abortion injunctions on the infant death
+model to quantify the impact of abortion injunctions on infant death
 rate across states in the USA. This method compares changes in infant
@@
 shifts in demographic factors such as maternal age or racial makeup
-between pre- and post-intervention periods may confound the observed
+between pre-and post-intervention periods may confound the observed
 changes in infant mortality rates. Including control variables helps
 adjust for these confounders, ensuring that the estimated effect
-reflects the policy intervention rather than changes in population
+reflects the policy intervention rather than changes in the population
 composition.
@@
 The length of the analysis period further complicates the reliability of
-results. While longer periods allow for capturing long-term effects,
+the results. While longer periods allow for capturing long-term effects,
 they also increase the likelihood of unrelated factors, such as new
@@
 -   **Time Indicator `after_injunction`**: A binary variable indicating
-    the time period: 0 for the period before the injunction implemented.
-    1 for the period after the injunction implemented.
+    the time period: 0 for the period before the injunction was implemented.
+    1 for the period after the injunction was implemented.
 
@@
 
--   **`state`**: Mother's residence state when the infant born as a
+-   **`state`**: Mother's residence state when the infant was born as a
     random effect in the model.
@@
 policy intervention. Additional details and diagnostics supporting the
-model’s implementation are provided in [Appendix- @sec-model-details].
+model’s implementation is provided in [Appendix- @sec-model-details].
 
@@
 In our analysis, we employ two models. The first model includes only the
-predictor variables and the random effect for state. The second model
+predictor variables and the random effect for the state. The second model
 builds on the first by incorporating additional control variables to
@@
 estimates of the intervention’s effect. In the other model, we exclude
-these controls to focus on the overall intervention impact, avoid
+these controls focus on the overall intervention impact, and avoid
 overfitting, and prioritize simplicity when variability is adequately
 accounted for by the design or random effects. Thus, our decision to fit
-two models is to compare the outcomes: one adjusts for confounders to
+two models to compare the outcomes: one adjusts for confounders to
 refine the estimates, while the other focuses on the direct effect of
@@
 for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
-priors based on data scale. This enhances convergence and ensures priors
+priors based on the data scale. This enhances convergence and ensures priors
 are calibrated to predictor ranges without dominating the data. The
@@
   fmt = 3,coef_omit="state",notes = "Note.
-All models include a random variable for state.
-The reference level of age_of_mother is 15-19.The reference level of monthers_single_race is Asian.
+All models include a random variable for the state.
+The reference level of age_of_mother is 15-19.
+The reference level of monthers_single_race is Asian.
 Source: CDC Wonder")
@@
 groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, respectively. The
-coefficient for Black mothers in Model 2 is 6.66, reflecting higher
+the coefficient for Black mothers in Model 2 is 6.66, reflecting a higher
 infant mortality compared to other racial groups.
@@
 abortion_illegal1) reveals that abortion bans during the post-injunction
-period result in an additional increase of 0.111 deaths in Model 1 and
+period results in an additional increase of 0.111 deaths in Model 1 and
 0.280 deaths in Model 2. Demographic factors also play a critical role,
@@
 the need for integrated healthcare systems that can address the
-consequences of high-risk pregnancies, improve prenatal care, and ensure
+consequences of high-risk pregnancies, improve prenatal care and ensure
 equitable access to medical services. Policymakers must consider the
@@
 In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model.
-It represent posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data.
+It represents posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data.
 The posterior predictive checks for Model 1 (policy-focused) and Model 2 (policy and demographic factors) illustrate model fit.
@@
 #| fig-cap: "Examining how the model for infant death rate fits, and is affected by, the data"
-#| fig-subcap: ["First Model Posterior prediction check","Second Model Posterior prediction check", "First Model Comparing the posterior with the prior", "Second Model Comparing the posterior with the prior"]
+#| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check", "First Model Comparing the posterior with the prior", "Second Model Comparing the posterior with the prior"]
 
@@
 for a deeper understanding of how restrictive abortion policies affect
-maternal health, access to healthcare, and infant outcomes, capturing
+maternal health, access to healthcare, and infant outcomes capturing
 nuances that quantitative analysis alone cannot.
@@
 psychological challenges, compounded by societal pressures and systemic
-barriers, can significantly increase the risk of adverse infant
+barriers can significantly increase the risk of adverse infant
 outcomes, including infant mortality. The survey will focus on mothers
@@
 findings will guide the development of evidence-based strategies to
-improve maternal mental health, enhance healthcare access, and shape
+improve maternal mental health, enhance healthcare access and shape
 policy responses to restrictive abortion laws.
@@
 In this analysis, we will use non-probability Respondent-driven sampling
-to finding respondents. Respondent-driven sampling (RDS), developed by
+to find respondents. Respondent-driven sampling (RDS), developed by
 @heckathorn1997, is ideal for studying hidden populations lacking a
@@
 RDS utilizes peer networks, allowing participants to recruit others they
-trust, fostering higher participation rates in populations that may
+trust, and fostering higher participation rates in populations that may
 distrust external researchers. This approach helps maintain
@@
 of mothers who fit the defined sample frame—those who were unable to
-access abortion services due to restrictive laws, experienced the death
+access abortion services due to restrictive laws, experience death
 of their infant within the first year of life, and are reachable through
@@
 Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with nuanced insights, fostering trust and enhancing data accuracy.
-Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each sectionof our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.
+Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each section of our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.
 
@@
 We draw on recommendations from @stantcheva2023 to minimize response biases.
-Common response biases identified in survey design include moderacybias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB).
-Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB.The detailed definitions of these biases are provided in [Appendix-@sec-definition].
+Common response biases identified in survey design include moderacy bias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB).
+Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB.
+The detailed definitions of these biases are provided in [Appendix-@sec-definition].
 
@@
 
-3.What is your marital status? - Single - Married - Divorced - Widowed -
+3. What is your marital status? - Single - Married - Divorced - Widowed -
 Prefer not to say

### 2024-11-27 paper/paper.qmd
update data details

@@
 # Additional data details {#sec-data-details}
+
+## Dataset and Graph Sketches
+
+Sketches depicting both the desired dataset and the graphs generated in
+this analysis is available in the GitHub Repository `other/sketches`.
+
+## Data Cleaning
+
+The raw dataset, consisting of U.S.
+infant mortality records from 2021 and 2022, was cleaned and pre-processed to ensure its suitability for analysis.
+The dataset was first imported from a tab-separated file, with irregular lines handled appropriately.
+Variables were standardized using `janitor::clean_names()` to ensure consistent naming conventions.
+Key variables such as state, maternal age, maternal race, year of death, month, and infant death rates were selected for analysis, and missing values were removed.
+
+Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data corresponded to the period after June 2022 (post-Dobbs decision), and `abortion_illegal`, denoting whether the state had restrictive abortion laws.
+Death rate reliability was assessed, and unreliable rates were flagged for transparency.
+The death rate column was cleaned to remove annotations and converted to numeric format for analysis.
+A date variable was created by combining the year and month columns, providing a timeline for analysis.
+Finally, a date variable was created by combining the year and month variables.
+The cleaned dataset was saved in CSV and Parquet formats to ensure compatibility with analytical workflows.
+
+## Attribution Statement
+
+The data utilized in this study was sourced from the CDC WONDER Online Database.
+Access to the data and its use complies with the terms outlined on the [CDC WONDER Data Use Agreement](https://wonder.cdc.gov/datause.html).
+Specifically, the data was used for academic research purposes, with acknowledgment of the Centers for Disease Control and Prevention (CDC) as the original data provider.
+The CDC, however, does not assume responsibility for any analyses, interpretations, or conclusions drawn from the data by the authors of this study.
 

### 2024-11-28 paper/paper.qmd
update paper with prose error

@@
 These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities.
-By utilizing recent, high-quality data and rigorous modeling, this study provides critical insights into the intersection of reproductive rights and infant health outcomes."
+By utilizing recent, high-quality data and rigorous modeling, this study provides essential understanding into the intersection of reproductive rights and infant health outcomes."
 format: pdf
@@
 restrictiveness. Key predictors include the mother’s state of residence,
-race, age, and the year of infant death, enabling a nuanced examination
+race, age, and the year of infant death, enabling a detailed examination
 of the effects of restrictive abortion laws on public health.
@@
 evidence on how restrictive abortion policies disproportionately affect
-vulnerable populations, emphasizing the need for data-driven policies to
+vulnerable populations, emphasizing the need for evidence-based policies to
 mitigate these adverse outcomes.
@@
 impact of abortion policies and demographic factors. The CDC’s Linked
-Birth/Infant Death Records database provides comprehensive data,
+Birth/Infant Death Records database provides detailed data,
 including maternal and infant characteristics, to support this study.
@@
 Once cleaned, the data is stratified by maternal and infant attributes,
-providing insights into trends and disparities across demographic
+providing understanding into trends and disparities across demographic
 groups. Statistical models are employed to estimate associations between
@@
 confounders such as maternal characteristics. This structured approach
-transforms complex datasets into actionable insights, shedding light on
+transforms complex datasets into actionable understanding, highlighting on
 the health implications of varying abortion laws across the U.S. By
-utilizing comprehensive datasets and robust statistical techniques,
-this study offers a nuanced understanding of the intersection between
+utilizing detailed datasets and robust statistical techniques,
+this study offers a detailed understanding of the intersection between
 policy, demographics, and infant health outcomes, informing the public
@@
 #| label: fig-deathrate
-#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants.
-The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
+#| fig-cap: Distribution of Infant Death Rates: The infant death rate (per 1,000 live births) shows a peak between 5 and 10 deaths per 1,000 infants, with a right-skewed distribution.
+This skewness highlights that lower infant death rates are more prevalent in the dataset, while higher rates occur less frequently.
 #| echo: false
@@
 
-Another created variable `after_injunction` indicates whether the date
-of infants' death falls after the implementation of the abortion
-injunctions following the overturn of Roe v. Wade in June 2022. It is
-coded as a binary variable, where 1 represents dates after June 2022,
-signifying the post-injunction period, and 0 represents dates before
-June 2022, indicating the pre-injunction period.
-
-To evaluate the impact of abortion injunctions on infant death rates, we
-utilize the Difference-in-Difference (DID) method. As part of this
-approach, we construct an interaction term defined as the product of
-`abortion_illegal` and `after_injunction`. The resulting interaction
-term equals 1 when abortion is illegal the time is after the
-injunction, and 0 otherwise. This term is included in the analysis to
-evaluate the joint impact of these factors on infant death rates.
+@fig-map illustrates the legality of abortion across U.S. states. Based
+on @reproductiveright, after June 2022, abortion has been illegal in
+Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi,
+Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West
+Virginia, and Wisconsin following the June 2022 decision.
 
@@
 #| label: fig-map
-#| fig-cap: Abortion Legality by State.
-As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
+#| fig-cap: Abortion Legality by State: After June 2022, abortion is prohibited in the following states: Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
 #| echo: false
@@
 
-@fig-map illustrates the legality of abortion across U.S. states. Based
-on @reproductiveright, as of November 2024, abortion has been illegal in
-Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi,
-Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West
-Virginia, and Wisconsin following the June 2022 decision.
-
 ```{r}
 #| label: fig-rateandstate
-#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
+#| fig-cap: Infant Death Rates Over Time: Infant death rates (per 1,000 live births) show a slight increase in states where abortion became illegal following June 2022, whereas rates in states with legal abortion remained relatively stable.
+This trend suggests a potential association between restrictive abortion laws and higher infant mortality rates.
 #| echo: false
@@
   geom_smooth() +
-  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S.
+  labs(x = "Date of Infant Death", y = "Infant Death Rate \n(per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S.
 Supreme Court overturned Roe v.
@@
 #| label: tbl-interact
-#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing.
-States with "Illegal Abortion and After Injunction" (denoted as '1') have a higher average infant death rate compared to states with "Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction" (denoted as '0').
+#| tbl-cap: Average Infant Death Rate by Abortion Legality and Injunction Timing: States categorized as "Illegal Abortion and After Injunction" ('1') exhibit a higher average infant death rate (per 1,000 live births) compared to states in other categories, including "Legal Abortion and Before/After Injunction" or "Illegal Abortion and Before Injunction" ('0').
+This highlights a potential link between restrictive abortion laws implemented after the injunction and elevated infant mortality rates.
 #| echo: false
@@
 ```
+
+Another created variable `after_injunction` indicates whether the date
+of infants' death falls after the implementation of the abortion
+injunctions following the overturn of Roe v. Wade in June 2022. It is
+coded as a binary variable, where 1 represents dates after June 2022,
+signifying the post-injunction period, and 0 represents dates before
+June 2022, indicating the pre-injunction period.
+
+To evaluate the impact of abortion injunctions on infant death rates, we
+utilize the Difference-in-Difference (DID) method. As part of this
+approach, we construct an interaction term defined as the product of
+`abortion_illegal` and `after_injunction`. The resulting interaction
+term equals 1 when abortion is illegal the time is after the
+injunction, and 0 otherwise. This term is included in the analysis to
+evaluate the joint impact of these factors on infant death rates.
 
@@
 and identifying patterns or shifts in infant mortality rates over this
-critical period. By controlling the year of death, we can closely
+essential period. By controlling the year of death, we can closely
 examine the impact of significant policy changes, such as the
@@
 #| label: tbl-motherage
-#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers.
-It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
+#| tbl-cap: Average Infant Death Rate by Maternal Age Group: Infant mortality rates (per 1,000 live births) vary significantly by maternal age, with younger mothers experiencing higher rates.
+Mothers aged 20–24 exhibit the highest infant death rate at 10.0, while the lowest rate of 6.6 is observed among mothers aged 30–34, indicating a potential association between maternal age and infant health outcomes.
 #| echo: false
@@
 #| label: tbl-motherrace
-#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers.
-It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
+#| tbl-cap: Average Infant Death Rate by Mother’s Single Race: Infant mortality rates (per 1,000 live births) vary significantly by race, with Black or African American mothers experiencing the highest rate at 14.5, compared to 5.8 for White mothers and 5.4 for Asian mothers.
+This highlights potential disparities in infant health outcomes linked to racial factors.
 #| echo: false
@@
 #| fig-cap: Infant Death Rates Across U.S.
-States (per 1,000 Live Births), December 2022.
-The infant Mortality Rate is much higher in Southeastern.
-Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
+States (per 1,000 Live Births): The infant mortality rate is significantly higher in the Southeastern states, with Alabama and Mississippi reporting the highest rates.
+In contrast, states like California and New York exhibit notably lower rates.
 #| echo: false
@@
 identifies changes in infant death rates that can be attributed to the
-policy intervention. Additional details and diagnostics supporting the
+policy intervention. Additional details, validation of model assumption, and diagnostics supporting the
 model’s implementation is provided in [Appendix- @sec-model-details].
@@
 
-For our analysis, we use separate Bayesian regression models to estimate
-the infant death rates in states where abortion injunctions were
-implemented and in states where abortion remains legal. We use the
-interaction term between after_injunction and abortion_illegal to
-measure the causal effect of the abortion injunction. Both models share
-the same predictors to ensure consistency and comparability in capturing
-the determinants of infant mortality.
-
-For one of our models, we include control variables (`age_of_mother`,
-`mothers_single_race`, and `year_of_death`) to adjust for confounders
-that could influence infant death rates, addressing the assumption of
-compositional differences in the DID approach and ensuring unbiased
-estimates of the intervention’s effect. In the other model, we exclude
-these controls focus on the overall intervention impact, and avoid
-overfitting, and prioritize simplicity when variability is adequately
-accounted for by the design or random effects. Thus, our decision to fit
-two models to compare the outcomes: one adjusts for confounders to
-refine the estimates, while the other focuses on the direct effect of
-the intervention.
-
-Our model includes `state` as a random effect to account for regional
-variations in infant mortality that could introduce bias. We use a
-random effect rather than a fixed effect to capture variability across
-states without estimating specific effects for each one. This approach
-models unobserved heterogeneity, such as differences in healthcare
-systems and socioeconomic conditions, using a single variance term,
-which avoids overfitting and ensures efficiency and generalizability. By
-adjusting for unmeasured state-specific factors, the model provides
-robust and reliable findings while balancing parsimony and
-interpretability, making it well-suited for our analysis.
-
-We choose the Bayesian framework since it enhances the DID approach by
-incorporating prior information and quantifying uncertainty in a
-probabilistic manner. In our Bayesian models, we use weakly informative
-priors from the `rstanarm` package, including a Normal prior ($\mu=0$
-and $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1)
-for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
-priors based on the data scale. This enhances convergence and ensures priors
-are calibrated to predictor ranges without dominating the data. The
-model assumes plausible parameter centering, random effects capture
-unobserved heterogeneity, and the DID framework relies on parallel
-trends between groups. However, limitations include potential bias from
-misspecified priors, computational challenges for hierarchical models,
-and sensitivity to data sparsity or imbalances. This model may not be
-suitable if group-level effects are negligible, parallel trends are
-violated, or data is insufficient, but when applied appropriately, it
-provides robust and interpretable causal estimates.
+#### Analysis Framework
+
+For our analysis, we employ Bayesian regression models to estimate infant death rates in states where abortion injunctions were implemented and states where abortion remains legal.
+The causal effect of the abortion injunction is measured using the interaction term between after_injunction and abortion_illegal.
+To ensure consistency and comparability, both models share the same predictors, capturing the determinants of infant mortality.
+
+One model includes control variables— `age_of_mother`, `mothers_single_race`, and `year_of_death` to adjust for confounders that may influence infant death rates.
+This adjustment addresses compositional differences assumed in the DID framework and ensures unbiased estimates of the intervention’s effect.
+The second model excludes these controls to focus on the overall intervention impact, avoid overfitting, and prioritize simplicity where variability is adequately accounted for by design or random effects.
+By fitting two models, we compare outcomes to refine estimates and evaluate the direct impact of the intervention.
+
+#### Model Design and Rationale
+
+Our models include state as a random effect to account for regional variations in infant mortality that might introduce bias.
+Using a random effect rather than a fixed effect captures variability across states without estimating individual effects, modeling unobserved heterogeneity—such as differences in healthcare systems and socioeconomic conditions—through a single variance term.
+This approach avoids overfitting, enhances efficiency, and ensures generalizability.
+By adjusting for unmeasured state-specific factors, the models strike a balance between parsimony and interpretability, making them well-suited for our analysis.
+
+#### Why DID Over Alternative Models?
+
+We selected the Difference-in-Differences (DID) framework because randomization is infeasible given the pre-existing differences between states in healthcare systems, socioeconomic conditions, and cultural factors.
+DID utilizes time and group-level variations, controlling for unobserved, time-invariant confounders.
+The parallel trends assumption ensures that untreated states provide a valid counterfactual for treated states.
+Interaction terms, such as after_injunction and abortion_illegal, allow for direct measurement of policy effects, while random effects adjust for regional variability.
+
+Alternative models, such as Fixed Effects, Propensity Score Matching (PSM), and Interrupted Time Series (ITS), were considered but deemed less suitable.
+Fixed Effects cannot estimate the effects of variables that do not vary over time within a unit.
+PSM balances observed covariates but does not account for temporal dynamics.
+ITS analyzes trends within a single group but lacks a comparison group.
+DID’s ability to integrate temporal and group-level differences makes it the most robust and appropriate framework for estimating the impact of abortion bans on infant mortality.
+
+#### Bayesian Framework
+
+We adopt a Bayesian framework to enhance the DID approach by incorporating prior information and quantifying uncertainty probabilistically.
+Using the rstanarm package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale.
+This improves convergence and ensures priors align with predictor ranges without dominating the data.
+The Bayesian framework supports plausible parameter centering, captures unobserved heterogeneity through random effects, and ensures robust estimates within the DID framework, assuming parallel trends between groups.
+
+#### Limitations and Applicability
+
+While Bayesian DID models provide robust and interpretable causal estimates, they have limitations.
+Potential biases from misspecified priors, computational challenges in hierarchical models, and sensitivity to data sparsity or imbalances are notable risks.
+Additionally, these models may not be suitable if group-level effects are negligible, parallel trends are violated, or data are insufficient.
+When applied appropriately, however, they offer a reliable approach to understanding the causal effects of abortion bans on infant mortality.
 
 # Results {#sec-results}
+
+## Model Results
 
@@
 
+The model highlights key demographic and policy-driven impacts on infant mortality rates.
+The interaction between abortion bans and the post-injunction period is linked to increased mortality, while states enforcing abortion bans show consistently higher rates.
+Younger mothers and Black mothers face disproportionately higher risks, with limited access to healthcare and systemic inequities compounding their vulnerabilities, emphasizing the essential need for targeted health and social interventions.
+Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates.These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.
+
 The models performed well, with low RMSE values and high $R^2$ values,
@@
 
-The analysis reveals distinct patterns in infant mortality rates across
+The analysis shows distinct patterns in infant mortality rates across
 various policy and demographic factors. In states where abortion
@@
 deaths in Model 2. The DID interaction term (after_injunction1 ×
-abortion_illegal1) reveals that abortion bans during the post-injunction
+abortion_illegal1) shows that abortion bans during the post-injunction
 period results in an additional increase of 0.111 deaths in Model 1 and
-0.280 deaths in Model 2. Demographic factors also play a critical role,
+0.280 deaths in Model 2. Demographic factors also play a essential role,
 with older maternal age (25–34 years) reducing infant mortality rates,
@@
 challenges may experience amplified effects, suggesting an urgent need
-for comprehensive policy interventions to address these disparities.
+for detailed policy interventions to address these disparities.
 Tackling the root causes of these health disparities—such as improving
@@
 limitations could lead to an underestimation of the full effects of
-abortion bans. Future research should incorporate more comprehensive and
+abortion bans. Future research should incorporate more detailed and
 updated datasets to provide a clearer and more complete analysis of the
@@
 
-While this study provides valuable findings, several limitations should
+While this study provides meaningful findings, several limitations should
 be addressed in future research. The analysis relies on aggregate
@@
 
+## Parall Assumption: An Event Study Analysis
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| include: false
+
+library(fixest)  # For fast DiD modeling
+
+# Ensure the tidyverse package is loaded
+library(tidyverse)
+
+# Define relative_time as numeric (time relative to treatment)
+cleaned_data <- analysis_data %>%filter(year_of_death==2022)|>
+  mutate(relative_time = (month_code-6))
+
+# Round relative_time to account for fractional months if necessary
+cleaned_data <- cleaned_data %>%
+  mutate(relative_time = as.factor(round(relative_time, 0)))|>mutate(abortion_illegal=as.numeric(as.character(abortion_illegal)))
+
+est_did <- feols(
+  death_rate ~ mothers_single_race + age_of_mother + i(relative_time , abortion_illegal, ref = -1) | state + relative_time, 
+  data = cleaned_data
+)
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-eventstudy
+#| tbl-cap: "Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United States: Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May) as the Reference Period"
+#| warning: false
+
+iplot(est_did,main = "Impact of Abortion Bans on Infant Mortality Rates by State",
+  value.lab = "Estimate and __ci__ Conf. Int.",xlab = "Relative Time (Month)")
+```
+
+@fig-eventstudy depicts the estimated effects of abortion bans on infant mortality rates over time, with relative months on the x-axis and coefficient estimates with 95% confidence intervals on the y-axis.
+Before the implementation of abortion bans ($t<0$), the coefficients are centered around zero, indicating no significant differences in infant mortality rates between treated and control states, supporting the parallel trends assumption.
+Post-ban ($t\geq0$), the estimates show a slight upward trend, suggesting a potential increase in infant mortality rates in restricted states; however, the wide confidence intervals, which overlap with zero, indicate that these effects are not statistically significant.
+Overall, the results do not provide conclusive evidence of a causal relationship between abortion bans and changes in infant mortality rates, emphasizing the need for further analysis with more precise data or extended time frames.
+
 ## Posterior predictive check
@@
 Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
-The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a crucial role in understanding infant mortality in the context of restrictive abortion laws.
+The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a essential role in understanding infant mortality in the context of restrictive abortion laws.
 
@@
 ```
-Credibility intervals are the Bayesian equivalent of confidence intervals, offering insight into the range of probable values for a parameter @tellingstories.
+Credibility intervals are the Bayesian equivalent of confidence intervals, offering understanding into the range of probable values for a parameter @tellingstories.
 Using Bayesian estimation, we derive a posterior distribution for each model coefficient.
@@
 
-In our study, we examine the impact of abortion bans on infant mortality
-rates in the United States following the Dobbs decision, employing both
-observational data and survey-based research. Observational data provide
-quantitative insights, revealing trends and associations across large
-populations. However, they often fall short in addressing underlying
-mechanisms due to limitations such as confounding factors and challenges
-in establishing causality.
-
-To address these gaps, in this appendix we propose incorporating surveys
-to gather qualitative data, which provide rich, detailed accounts of
-individual experiences and contextual factors. Qualitative methods allow
-for a deeper understanding of how restrictive abortion policies affect
-maternal health, access to healthcare, and infant outcomes capturing
-nuances that quantitative analysis alone cannot.
-
-By combining quantitative and qualitative approaches, this study
-utilizes the strengths of both methods. Quantitative data offer
-generalizable findings and statistical rigor, while qualitative data
-provide depth and context to interpret those findings. Together, they
-create a more comprehensive framework for understanding the complex
-relationship between restrictive abortion policies, maternal health, and
-infant outcomes, enabling the development of nuanced, evidence-based
-interventions and policy recommendations.
+Our study examines the impact of abortion bans on infant mortality rates in the U.S.
+following the Dobbs decision, combining observational data and surveys.
+Observational data reveal trends and associations, but often struggle with confounding factors and causality.
+Surveys address these gaps by capturing qualitative understanding into individual experiences and contextual factors.
+This mixed-methods approach integrates statistical rigor with detailed personal accounts, offering a detailed framework to understand how restrictive abortion policies affect maternal health and infant outcomes, and informing evidence-based policy recommendations.
 
@@
 
-Restrictive abortion laws may exacerbate maternal stress, anxiety, or
-depression—especially in cases of unintended or high-risk
-pregnancies—potentially undermining maternal health and prenatal care.
-Research indicates that maternal mental health is a critical determinant
-of infant health outcomes, as conditions such as chronic stress,
-depression, or anxiety during pregnancy can negatively impact prenatal
-behaviors, access to healthcare, and overall maternal well-being. These
-psychological challenges, compounded by societal pressures and systemic
-barriers can significantly increase the risk of adverse infant
-outcomes, including infant mortality. The survey will focus on mothers
-unable to access abortion services who subsequently experienced the
-death of their infant, documenting their experiences to better
-understand these interconnected factors.
-
-To provide a comprehensive perspective, the study will also involve
-healthcare providers and support organizations, capturing insights into
-systemic challenges and the broader impacts of abortion bans on family
-healthcare. By combining these perspectives with mothers’ lived
-experiences, the survey aims to inform evidence-based interventions
-addressing maternal psychological and healthcare needs and mitigating
-the negative consequences of restrictive abortion policies.
-
-This dual approach, integrating quantitative and qualitative methods,
-seeks to uncover the complex relationship between restrictive abortion
-laws, maternal mental health, and infant outcomes, forming the basis for
-effective policy recommendations.
+Restrictive abortion laws can heighten maternal stress, anxiety, or depression—particularly in unintended or high-risk pregnancies—compromising maternal health and prenatal care.
+Research @crugnola2022 highlights maternal mental health as a key determinant of infant health outcomes, with conditions like chronic stress and depression adversely affecting prenatal behaviors, healthcare access, and overall well-being.
+These compounded challenges, driven by societal and systemic barriers, can significantly increase risks of adverse infant outcomes, including mortality.
+
+Our survey focuses on mothers unable to access abortion services who later experienced the death of their infant, documenting their experiences to better understand these interconnected factors.
+By integrating quantitative data with qualitative perspectives, this study aims to inform evidence-based interventions, addressing maternal psychological and healthcare needs, and mitigating the adverse effects of restrictive abortion policies.
 
@@
 The objective of this study is to investigate the psychological, social,
-and systemic impacts of restrictive abortion laws on maternal health and
-infant outcomes. By focusing on mothers unable to access abortion
-services who experienced the death of their infant, the study aims to
-document lived experiences, explore the role of maternal stress and
-healthcare barriers in adverse outcomes, and gather perspectives from
-healthcare providers and support organizations. Understanding how
-maternal mental health directly affects prenatal care decisions, access
-to healthcare, and overall maternal behavior is essential to addressing
-the systemic factors contributing to higher infant mortality rates. The
-findings will guide the development of evidence-based strategies to
-improve maternal mental health, enhance healthcare access and shape
-policy responses to restrictive abortion laws.
+and systemic impacts of restrictive abortion laws on maternal health and infant outcomes.
+By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences, explore the role of maternal stress and healthcare barriers in adverse outcomes.
+Understanding how maternal mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the systemic factors contributing to higher infant mortality rates.
+The findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access and shape policy responses to restrictive abortion laws.
 
@@
 
-In this analysis, we will use non-probability Respondent-driven sampling
-to find respondents. Respondent-driven sampling (RDS), developed by
-@heckathorn1997, is ideal for studying hidden populations lacking a
-sampling frame and facing potential stigma or harm. Unlike snowball
-sampling, RDS offers dual compensation for participation and
-recruitment, with recruitment occurring peer-to-peer via social
-networks. This method maintains confidentiality, reduces bias through
-statistical adjustments, and is particularly effective for sensitive
-topics, such as abortion, by utilizing trust within communities.
-
-We use Respondent-Driven Sampling (RDS) for our survey because it is
-well-suited for reaching hidden or stigmatized populations, such as
-mothers who were unable to access abortion services and experienced the
-death of their infant. These individuals may be difficult to identify
-through conventional methods due to the sensitive nature of the topic
-and the lack of a clear sampling frame.
-
-RDS utilizes peer networks, allowing participants to recruit others they
-trust, and fostering higher participation rates in populations that may
-distrust external researchers. This approach helps maintain
-confidentiality and reduces the risk of social stigma. Additionally, RDS
-provides a structured framework for statistical adjustments, mitigating
-biases inherent in network-based recruitment, and improving the
-representativeness of the sample. These advantages make RDS an effective
-and ethical choice for gathering insights into the psychological,
-social, and systemic impacts of restrictive abortion laws.
+In this analysis, we use non-probability Respondent-Driven Sampling (RDS), developed by @heckathorn1997, is a recruitment method designed for hidden populations lacking a sampling frame, where participants recruit others through their social networks, with dual compensation for participation and recruitment, ensuring confidentiality and mitigating biases.
+RDS enables peer-to-peer recruitment with dual compensation for participation and recruitment, fostering trust within communities.
+This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics like abortion.
+
+We use Respondent-Driven Sampling (RDS) for our survey as it is effective for accessing hidden or stigmatized populations, such as mothers unable to access abortion services and who experienced the death of their infant.
+RDS utilizes peer networks for recruitment, fostering trust and participation in sensitive contexts while maintaining confidentiality and reducing social stigma.
+Its statistical adjustments mitigate biases in network-based recruitment, enhancing sample representativeness.
+These strengths make RDS an ethical and effective method for examining the psychological, social, and systemic impacts of restrictive abortion laws.
 
@@
 
-Given the sensitive nature of our study, identifying and reaching the
-initial participants poses a significant challenge. To address this, we
-aim to collaborate with support groups and advocacy organizations like
-Abortion Rights Advocacy Groups. These groups play a critical role in
-connecting researchers with individuals who are otherwise difficult to
-reach, particularly for sensitive topics like the impact of restrictive
-abortion laws on maternal and infant health.
-
-Therefore, the sample frame of our study will be mothers who live in
-areas where they are unable to access abortion services due to
-restrictive laws and have experienced the death of their infant within
-the first year of life in states with abortion bans following the Dobbs
-decision and are reachable and can be contacted by Abortion Rights
-Advocacy Groups.
+Given the sensitive nature of this study, identifying and reaching participants presents a significant challenge.
+To address this, we will collaborate with support groups and advocacy organizations, such as Abortion Rights Advocacy Groups, which can connect researchers with individuals otherwise difficult to reach.
+These organizations are essential for engaging with populations affected by restrictive abortion laws, particularly concerning maternal and infant health.
+
+Thus, the sample frame for our study comprises mothers who reside in areas where abortion services are inaccessible due to restrictive laws, have experienced the death of their infant within the first year of life, and are reachable through advocacy organizations following the Dobbs decision.
 
@@
 
-We plan to survey 20 respondents for this study. Our sample will consist
-of mothers who fit the defined sample frame—those who were unable to
-access abortion services due to restrictive laws, experience death
-of their infant within the first year of life, and are reachable through
-Abortion Rights Advocacy Groups. These mothers must also be willing to
-participate in and answer our survey questions, ensuring the reliability
-and depth of the collected data.
+We aim to survey 20 respondents who meet our defined sample criteria: mothers unable to access abortion services due to restrictive laws, who experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups.
+Participation will be voluntary, and respondents must be willing to answer the survey questions, ensuring the quality and depth of the data collected.
 
@@
 
-To recruit participants for this sensitive study, we will use a Respondent-Driven Sampling (RDS) approach in collaboration with  Abortion Rights Advocacy Groups, that have established relationships with our target population.
-These organizations will serve as initial “seeds” to identify and reach mothers affected by restrictive abortion laws who meet our study criteria.
-Outreach materials emphasizing confidentiality and the study’s importance will be shared through trusted channels.
-As part of the RDS methodology, participants may also invite others from their network who share similar experiences, ensuring the sample expands while maintaining anonymity.
-Interested participants will undergo a screening process and provide informed consent, with the option to complete the survey anonymously online, ensuring safety and confidentiality.
-To support participant well-being, we will provide access to mental health resources throughout the process, fostering trust and encouraging meaningful participation.
+To recruit participants for this sensitive study, we will employ a Respondent-Driven Sampling (RDS) approach in collaboration with Abortion Rights Advocacy Groups, leveraging their established connections with our target population.
+This organization will act as initial "seeds" to identify mothers affected by restrictive abortion laws who meet our study criteria.
+Participants may also invite others from their network with similar experiences, allowing the sample to grow while maintaining confidentiality.
+Outreach materials emphasizing anonymity and the study’s significance will be shared through trusted channels.
+Interested participants will undergo a screening process and provide informed consent, with the option to complete the survey online for safety and privacy.
+To ensure participant well-being, mental health resources will be offered throughout the process, fostering trust and encouraging meaningful engagement.
 
@@
 
-To ensure the credibility of our data, we implement a respondent validation process.
-Participants will undergo an initial screening to confirm they meet the study criteria, such as being unable to access abortion services due to restrictive laws and experiencing the loss of an infant within the first year of life.
-Responses will be reviewed for completeness and consistency, and duplicate entries will be removed to maintain data integrity.
-Collaboration with trusted support groups and advocacy organizations further enhances respondent reliability, ensuring that our findings accurately reflect the experiences of the target population.
+To ensure data credibility, we implement a robust respondent validation process.
+Participants will first undergo screening to confirm eligibility, including inability to access abortion services due to restrictive laws and experiencing infant loss within the first year of life.
+Responses will be assessed for completeness and consistency, and duplicate entries will be removed to maintain integrity.
+Partnering with trusted support groups and advocacy organizations enhances reliability, ensuring the findings accurately represent the target population’s experiences.
 
@@
 
-This study addresses sensitive and emotionally charged topics,
-necessitating a careful and ethical approach to protect participants’
-well-being. We acknowledge the potential psychological distress that
-discussing experiences of unintended pregnancies, restrictive abortion
-laws, and infant loss may cause. To mitigate this, the survey will
-include trigger warnings, and participants will have the option to skip
-questions or withdraw at any time without penalty. All data collection
-will adhere to strict confidentiality protocols, ensuring that
-participants’ identities remain anonymous and their responses are
-securely stored. Recruitment through trusted advocacy groups will ensure
-that participants feel safe and supported throughout the process.
-Additionally, we will provide resources for mental health support to any
-participant who may experience discomfort as a result of participating
-in the study. This approach underscores our commitment to conducting
-ethical, respectful, and responsible research.
+This study addresses sensitive and emotionally charged topics, requiring an ethical approach to safeguard participants’ well-being.
+Recognizing the potential psychological distress associated with discussing unintended pregnancies, restrictive abortion laws, and infant loss, the survey will include trigger warnings and allow participants to skip questions or withdraw at any time without penalty.
+Strict confidentiality protocols will protect participants’ identities, and responses will be securely stored.
+Recruitment through trusted advocacy groups will foster a safe and supportive environment, while mental health resources will be provided to those who experience discomfort during the study.
+This framework reflects our commitment to ethical, respectful, and responsible research practices.
 
@@
 experimental approaches significantly improved abortion reporting
-compared to the control condition, their study offers valuable insights
+compared to the control condition, their study offers meaningful guidelines
 for improving abortion measurement. We aim to adapt our survey design
@@
 This survey explores the impacts of restrictive abortion laws on maternal mental health and infant outcomes, emphasizing neutral, carefully crafted questions to ensure participant comfort and minimize social desirability bias.
-Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., “Prefer not to say”) to address underreporting in sensitive topics.
-Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with nuanced insights, fostering trust and enhancing data accuracy.
-Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each section of our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.
+Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., "Prefer not to say") to address underreporting in sensitive topics.
+Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with detailed understanding, fostering trust and enhancing data accuracy.
+Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a "helping" framework in each section of our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.
 
@@
 
-- Addressing Extreme/Moderacy Bias: We use a minimum of five response options for scale questions to provide more nuanced choices, reducing the likelihood of respondents defaulting to extreme or middle answers.
+- Addressing Extreme/Moderacy Bias: We use a minimum of five response options for scale questions to provide more detailed choices, reducing the likelihood of respondents defaulting to extreme or middle answers.
 
@@
 
-- Minimizing Social Desirability Bias (SDB): Our online survey is carefully designed to minimize social desirability bias (SDB) while addressing the sensitivity of the topic.
-The survey begins with a clear introduction explaining its purpose—academic research on the impact of restrictive abortion laws—while emphasizing respondent anonymity and confidentiality.
-Before sensitive questions, participants are reassured that their responses are entirely confidential and will only be used for research purposes.
-The anonymous online format provides a safe space for participants to share their experiences without fear of judgment or stigma.
-Additionally, a feedback section at the end invites respondents to voice any concerns or share additional thoughts, fostering trust and enhancing the quality of the data collected.
+- Minimizing Social Desirability Bias (SDB): The survey design addresses social desirability bias (SDB) by emphasizing anonymity and confidentiality throughout.
+A clear introduction outlines the survey’s purpose—academic research on the impact of restrictive abortion laws—and reassures participants that their responses will remain confidential and solely used for research.
+The anonymous online format creates a safe environment for participants to share their experiences without fear of judgment or stigma.
+A feedback section at the end encourages participants to express concerns or share additional thoughts, fostering trust and enhancing data quality.
 
@@
 especially those involving circumstances such as abortion or the loss of
-a child. Your insights are invaluable in identifying areas where
+a child. Your understanding are invaluable in identifying areas where
 healthcare and support services can be improved. If you prefer not to
@@
     -   Website: www.postpartum.net
-    -   Helpline: 1-800-944-4773 (Text “Help” to 800-944-4773)
+    -   Helpline: 1-800-944-4773 (Text "Help" to 800-944-4773)
     -   Services: Support for mental health during and after pregnancy,
@@
 
-- **Acquiescence bias**: The habit of giving positive responses, such as consistently choosing “agree” or “yes.”
+- **Acquiescence bias**: The habit of giving positive responses, such as consistently choosing "agree" or "yes."
 \newpage

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper name

--- 2024-11-28 paper/paper.qmd
@@
 We adopt a Bayesian framework to enhance the DID approach by incorporating prior information and quantifying uncertainty probabilistically.
-Using the rstanarm package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale.
+Using the `rstanarm` package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale.
 This improves convergence and ensures priors align with predictor ranges without dominating the data.
@@
 Younger mothers and Black mothers face disproportionately higher risks, with limited access to healthcare and systemic inequities compounding their vulnerabilities, emphasizing the essential need for targeted health and social interventions.
-Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates.These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.
+Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates.
+These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.
 
@@
 period results in an additional increase of 0.111 deaths in Model 1 and
-0.280 deaths in Model 2. Demographic factors also play a essential role,
+0.280 deaths in Model 2. Demographic factors also play an essential role,
 with older maternal age (25–34 years) reducing infant mortality rates,
@@
 
-Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data corresponded to the period after June 2022 (post-Dobbs decision), and `abortion_illegal`, denoting whether the state had restrictive abortion laws.
+Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data correspond to the period after June 2022 (post-Dobbs decision) and `abortion_illegal`, denoting whether the state had restrictive abortion laws.
 Death rate reliability was assessed, and unreliable rates were flagged for transparency.
@@
 Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
-The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a essential role in understanding infant mortality in the context of restrictive abortion laws.
+The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.
 
@@
 ```
-Credibility intervals are the Bayesian equivalent of confidence intervals, offering understanding into the range of probable values for a parameter @tellingstories.
+Credibility intervals are the Bayesian equivalent of confidence intervals, offering an understanding of the range of probable values for a parameter @tellingstories.
 Using Bayesian estimation, we derive a posterior distribution for each model coefficient.
@@
 following the Dobbs decision, combining observational data and surveys.
-Observational data reveal trends and associations, but often struggle with confounding factors and causality.
-Surveys address these gaps by capturing qualitative understanding into individual experiences and contextual factors.
+Observational data reveal trends and associations but often struggle with confounding factors and causality.
+Surveys address these gaps by capturing a qualitative understanding of individual experiences and contextual factors.
 This mixed-methods approach integrates statistical rigor with detailed personal accounts, offering a detailed framework to understand how restrictive abortion policies affect maternal health and infant outcomes, and informing evidence-based policy recommendations.
@@
 and systemic impacts of restrictive abortion laws on maternal health and infant outcomes.
-By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences, explore the role of maternal stress and healthcare barriers in adverse outcomes.
+By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences, and explore the role of maternal stress and healthcare barriers in adverse outcomes.
 Understanding how maternal mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the systemic factors contributing to higher infant mortality rates.
-The findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access and shape policy responses to restrictive abortion laws.
+The findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access, and shape policy responses to restrictive abortion laws.
 
@@
 
-We aim to survey 20 respondents who meet our defined sample criteria: mothers unable to access abortion services due to restrictive laws, who experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups.
+We aim to survey 20 respondents who meet our defined sample criteria: mothers unable to access abortion services due to restrictive laws, who experienced the death of their infant within the first year of life, and who are reachable through Abortion Rights Advocacy Groups.
 Participation will be voluntary, and respondents must be willing to answer the survey questions, ensuring the quality and depth of the data collected.

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper based on peer reviews

@@
 library(rstanarm)
+library(fixest)  
 ```
@@
 Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
-regression modeling was conducted using `rstanarm` @rstanarm. Together,
-these packages supported an efficient, reproducible workflow for
+regression modeling was conducted using `rstanarm` @rstanarm. We use `fixest` @fixest to conduct the event study. Together, these packages supported an efficient, reproducible workflow for
 analyzing and presenting our findings.
@@
 #| label: fig-deathrate
-#| fig-cap: Distribution of Infant Death Rates: The infant death rate (per 1,000 live births) shows a peak between 5 and 10 deaths per 1,000 infants, with a right-skewed distribution.
-This skewness highlights that lower infant death rates are more prevalent in the dataset, while higher rates occur less frequently.
+#| fig-cap: >
+#|   Distribution of Infant Death Rates: The infant death rate (per 1,000 live births) 
+#|   shows a peak between 5 and 10 deaths per 1,000 infants, with a right-skewed 
+#|   distribution. 
 #| echo: false
@@
 #| label: fig-map
-#| fig-cap: Abortion Legality by State: After June 2022, abortion is prohibited in the following states: Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
+#| fig-cap: >
+#|   Abortion Legality by State: After June 2022, abortion is prohibited in the
+#|   following states: Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana,
+#|   Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee,
+#|   Texas, West Virginia, and Wisconsin.
 #| echo: false
@@
 #| label: fig-rateandstate
-#| fig-cap: Infant Death Rates Over Time: Infant death rates (per 1,000 live births) show a slight increase in states where abortion became illegal following June 2022, whereas rates in states with legal abortion remained relatively stable.
-This trend suggests a potential association between restrictive abortion laws and higher infant mortality rates.
+#| fig-cap: >
+#|   Infant Death Rates Over Time: Infant death rates (per 1,000 live births)
+#|   show a slight increase in states where abortion became illegal following
+#|   June 2022, whereas rates in states with legal abortion remained relatively
+#|   stable. 
 #| echo: false
@@
 #| label: tbl-interact
-#| tbl-cap: Average Infant Death Rate by Abortion Legality and Injunction Timing: States categorized as "Illegal Abortion and After Injunction" ('1') exhibit a higher average infant death rate (per 1,000 live births) compared to states in other categories, including "Legal Abortion and Before/After Injunction" or "Illegal Abortion and Before Injunction" ('0').
-This highlights a potential link between restrictive abortion laws implemented after the injunction and elevated infant mortality rates.
+#| tbl-cap: >
+#|   Average Infant Death Rate by Abortion Legality and Injunction Timing: 
+#|   States categorized as "Illegal Abortion and After Injunction" ('1') 
+#|   exhibit a higher average infant death rate (per 1,000 live births) compared 
+#|   to states in other categories, including "Legal Abortion and Before/After 
+#|   Injunction" or "Illegal Abortion and Before Injunction" ('0'). 
 #| echo: false
@@
 #| label: tbl-motherage
-#| tbl-cap: Average Infant Death Rate by Maternal Age Group: Infant mortality rates (per 1,000 live births) vary significantly by maternal age, with younger mothers experiencing higher rates.
-Mothers aged 20–24 exhibit the highest infant death rate at 10.0, while the lowest rate of 6.6 is observed among mothers aged 30–34, indicating a potential association between maternal age and infant health outcomes.
+#| tbl-cap: >
+#|   Average Infant Death Rate by Maternal Age Group: 
+#|   Infant mortality rates (per 1,000 live births) vary significantly by maternal age, 
+#|   with younger mothers experiencing higher rates. Mothers aged 20–24 exhibit the 
+#|   highest infant death rate at 10.0, while the lowest rate of 6.6 is observed among 
+#|   mothers aged 30–34.
 #| echo: false
@@
 #| message: false
-#| 
+
 analysis_data_state <- analysis_data|>
@@
 #| label: tbl-motherrace
-#| tbl-cap: Average Infant Death Rate by Mother’s Single Race: Infant mortality rates (per 1,000 live births) vary significantly by race, with Black or African American mothers experiencing the highest rate at 14.5, compared to 5.8 for White mothers and 5.4 for Asian mothers.
-This highlights potential disparities in infant health outcomes linked to racial factors.
+#| tbl-cap: >
+#|   Average Infant Death Rate by Mother’s Single Race: 
+#|   Infant mortality rates (per 1,000 live births) vary significantly by race, 
+#|   with Black or African American mothers experiencing the highest rate at 14.5, 
+#|   compared to 5.8 for White mothers and 5.4 for Asian mothers. 
 #| echo: false
@@
 #| label: fig-choromap
-#| fig-cap: Infant Death Rates Across U.S.
-States (per 1,000 Live Births): The infant mortality rate is significantly higher in the Southeastern states, with Alabama and Mississippi reporting the highest rates.
-In contrast, states like California and New York exhibit notably lower rates.
+#| fig-cap: >
+#|   Infant Death Rates Across U.S. States (per 1,000 Live Births): 
+#|   The infant mortality rate is significantly higher in the Southeastern states, 
+#|   with Alabama and Mississippi reporting the highest rates. In contrast, 
+#|   states like California and New York exhibit notably lower rates.
 #| echo: false
-
 death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
@@
 
-A key limitation of our analysis is the temporal and demographic scope
-of the dataset. The data includes only infant deaths under 1 year of age
-occurring within the United States to U.S. residents during 2021–2022.
-This timeframe does not capture the full year of infant mortality
-outcomes following the Dobbs decision, as infants who died in 2023 are
-not included. Furthermore, the dataset’s focus on infants under 1 year
-of age may exclude cases that reflect longer-term health impacts. These
-limitations could lead to an underestimation of the full effects of
-abortion bans. Future research should incorporate more detailed and
-updated datasets to provide a clearer and more complete analysis of the
-long-term trends and implications of restrictive abortion policies.
+A key limitation of our analysis is the temporal and demographic scope of the dataset.
+The data includes only infant deaths under 1 year of age occurring within the United States to U.S.
+residents during 2021–2022.
+This timeframe does not capture the full year of infant mortality outcomes following the Dobbs decision, as infants who died in 2023 are not included.
+Furthermore, the dataset’s focus on infants under 1 year of age may exclude cases that reflect longer-term health impacts.
+These limitations could lead to an underestimation of the full effects of abortion bans.
+As no publicly available dataset currently provides complete 2023 data or longer-term health outcomes, future research should incorporate more detailed and updated datasets to provide a clearer and more complete analysis of the long-term trends and implications of restrictive abortion policies.
 
@@
 
-library(fixest)  # For fast DiD modeling
-
-# Ensure the tidyverse package is loaded
-library(tidyverse)
-
-# Define relative_time as numeric (time relative to treatment)
 cleaned_data <- analysis_data %>%filter(year_of_death==2022)|>
@@
 ```{r}
+#| label: fig-eventstudy
+#| fig-cap: >
+#|   Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United   States: 
+#|   Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May) as the Reference Period.
+#| warning: false
 #| echo: false
 #| eval: true
-#| label: fig-eventstudy
-#| tbl-cap: "Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United States: Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May) as the Reference Period"
-#| warning: false
-
-iplot(est_did,main = "Impact of Abortion Bans on Infant Mortality Rates by State",
+
+iplot(est_did,main = "Impact of Abortion Bans on Infant \nMortality Rates by State",
   value.lab = "Estimate and __ci__ Conf. Int.",xlab = "Relative Time (Month)")
@@
 ## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot for model 1. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
 

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper based on peer review

@@
 
-The structure of this paper is organized as follows: @sec-data provides
-details on the data sources and variables used in our analysis.
-@sec-model explains the modeling approach, including the assumptions and
-specifications of our linear regression framework. In @sec-results, we
-present our findings, emphasizing the key predictors of Harris's
-support. Finally, @sec-discussion explores the implications of our
-results and suggests potential directions for future research.
-[Appendix- @sec-data-details] provides external data detail, [Appendix-
-@sec-model-details] provides model details, and [Appendix- @sec-survey]
-provides a detailed introduction to surveys, sampling, and observational
-data in our paper.
+The structure of this paper is as follows: @sec-data details the data sources and variables used in the analysis.
+@sec-model outlines the modeling approach, including assumptions and specifications of the Bayesian DID framework.
+In @sec-results, we present the findings, emphasizing key observations.In @sec-results, we present findings on the impact of abortion bans and the post-injunction period, highlighting their association with increased infant mortality.
+Key disparities by maternal age and race are discussed, along with model validation metrics like RMSE and $R^2$.
+Finally, @sec-discussion explores the implications of the results and suggests potential directions for future research.
+Additional details are provided in the appendices: [Appendix-@sec-data-details] elaborates on external data sources, [Appendix-@sec-model-details] provides an in-depth explanation of the model, and [Appendix-@sec-survey] introduces the idealized methodology for a related qualitative analysis.
 
@@
 
+Our analysis includes three key control variables: maternal age (`age_of_mother`), race (`mothers_single_race`), and year of death (`year_of_death`), to account for demographic and temporal differences.
+
+```{r}
+#| label: tbl-summary
+#| tbl-cap: >
+#|   Demographic Distribution of Infant Mortality Cases by Year, Maternal Age, and Race (2021–2022): 
+#|   The data highlights the proportional representation of cases across different years, 
+#|   maternal age groups, and racial categories. Younger mothers and Black or African American 
+#|   mothers show higher proportions of cases compared to other groups.
+#| echo: false
+#| warning: false
+#| message: false
+analysis_data|>mutate(year_of_death=as.factor(year_of_death))|>
+  select(year_of_death,age_of_mother,mothers_single_race) |> 
+  modelsummary::datasummary_skim()
+```
+
+@tbl-summary shows that 50.7% of infant deaths occurred in 2022, slightly higher than 2021 (49.3%).
+The majority of cases were among mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and 20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
+By race, White mothers accounted for the highest percentage (75.7%), followed by Black or African American mothers (24.0%), and Asian mothers (0.3%).
+
 ### Year of Death
@@
 
-We adopt a Bayesian framework to enhance the DID approach by incorporating prior information and quantifying uncertainty probabilistically.
-Using the `rstanarm` package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale.
-This improves convergence and ensures priors align with predictor ranges without dominating the data.
-The Bayesian framework supports plausible parameter centering, captures unobserved heterogeneity through random effects, and ensures robust estimates within the DID framework, assuming parallel trends between groups.
+We use a Bayesian approach to enhance the Difference-in-Differences (DID) method, which allows us to incorporate prior knowledge and better account for uncertainty.
+With the rstanarm package, we apply weakly informative priors—such as a Normal prior centered at zero with moderate spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for variability parameters.
+These priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data.
+This approach helps us center estimates realistically, account for hidden differences between groups, and produce reliable results while assuming trends between treated and control groups are similar before treatment.
 
@@
 
-While Bayesian DID models provide robust and interpretable causal estimates, they have limitations.
-Potential biases from misspecified priors, computational challenges in hierarchical models, and sensitivity to data sparsity or imbalances are notable risks.
-Additionally, these models may not be suitable if group-level effects are negligible, parallel trends are violated, or data are insufficient.
-When applied appropriately, however, they offer a reliable approach to understanding the causal effects of abortion bans on infant mortality.
+Our Bayesian DID model offers a robust approach to examining the impact of abortion bans on infant mortality but has limitations.
+Misspecified priors, data imbalances, and violations of the parallel trends assumption could introduce biases.
+Despite these challenges, the model effectively integrates prior knowledge and addresses uncertainty, providing valuable insights into policy impacts.
 
@@
 
-## Public Health Implications of Restrictive Abortion Laws
-
-The findings highlight the significant public health consequences of
-abortion bans in the United States. The observed increase of 0.285
-infant deaths per 1,000 live births in states enforcing abortion bans
-after the Dobbs decision underscores the detrimental impact of
-restrictive abortion policies. These increases often reflect systemic
-issues such as inadequate prenatal care, gaps in healthcare access, and
-underlying socioeconomic inequities. States with pre-existing healthcare
-challenges may experience amplified effects, suggesting an urgent need
-for detailed policy interventions to address these disparities.
-Tackling the root causes of these health disparities—such as improving
-prenatal care, expanding healthcare access, and addressing socioeconomic
-inequalities—will be essential for mitigating the unintended
-consequences of restrictive abortion legislation.
-
 ## Racial and Socioeconomic Disparities
@@
 
-## Integration of Policy and Healthcare Systems
-
-The significant interaction between abortion legality and the
-post-injunction period illustrates the compounded impact of restrictive
-abortion policies on infant mortality rates. These findings underscore
-the need for integrated healthcare systems that can address the
-consequences of high-risk pregnancies, improve prenatal care and ensure
-equitable access to medical services. Policymakers must consider the
-broader public health implications of abortion bans and implement
-strategies that safeguard maternal and infant health. This includes
-enhancing healthcare delivery in underserved areas, investing in
-community-based interventions, and ensuring that restrictive policies
-are accompanied by measures to mitigate their public health impact.
+## Addressing Public Health and Long-Term Implications of Restrictive Abortion Laws
+
+Addressing Public Health and Long-Term Implications of Restrictive Abortion Laws
+
+Restrictive abortion policies significantly impact maternal and infant health, as evidenced by an increase of 0.285 infant deaths per 1,000 live births in states enforcing abortion bans during the post-injunction period.
+These policies exacerbate systemic issues such as inadequate prenatal care, gaps in healthcare access, and socioeconomic inequities, disproportionately affecting marginalized communities.
+Over time, the consequences may include higher rates of untreated maternal mental health issues, delayed or insufficient prenatal care, and elevated infant mortality, creating lasting health challenges for families and communities.
+Addressing these disparities requires targeted interventions, such as expanding maternal mental health services, improving prenatal and postnatal care in underserved areas, ensuring culturally competent healthcare, and advocating for equitable policies.
+Investments in community-based support systems, enhanced healthcare delivery, and robust data infrastructure are essential to monitor and mitigate the long-term impacts of these policies, fostering sustainable improvements in maternal and infant health outcomes.
 
@@
 Our study examines the impact of abortion bans on infant mortality rates in the U.S.
-following the Dobbs decision, combining observational data and surveys.
-Observational data reveal trends and associations but often struggle with confounding factors and causality.
-Surveys address these gaps by capturing a qualitative understanding of individual experiences and contextual factors.
-This mixed-methods approach integrates statistical rigor with detailed personal accounts, offering a detailed framework to understand how restrictive abortion policies affect maternal health and infant outcomes, and informing evidence-based policy recommendations.
+following the Dobbs decision, combining observational Difference-in-Differences (DID) analysis with qualitative surveys.
+While the DID analysis quantifies policy impacts, qualitative surveys provide context-specific insights into how restrictive abortion policies affect maternal mental health, healthcare access, and pregnancy decisions.
+These lived experiences help interpret quantitative findings, offering a fuller understanding of the systemic and psychological factors driving observed trends, and supporting evidence-based policy recommendations.
 

### 2024-11-28 scripts/04-model_data.R
update script comments

@@ -1,11 +1,13 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Model Infant Death Rate in the U.S. across states in year 2021-2022 using Bayesian DID Model Approach.
+# Author: Diana Shen
+# Date: 20 November 2024 
+# Contact: diana.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# - Cleaned data file `analysis_data.parquet` should be available in the `data/02-analysis_data/` directory.
+# - Packages `arrow`, `rstanarm`, and `tidyverse` must be installed and loaded
+# Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 
 #### Workspace setup ####

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with prose error

@@
 Misspecified priors, data imbalances, and violations of the parallel trends assumption could introduce biases.
-Despite these challenges, the model effectively integrates prior knowledge and addresses uncertainty, providing valuable insights into policy impacts.
+Despite these challenges, the model effectively integrates prior knowledge and addresses uncertainty, providing valuable understanding into policy impacts.
 
@@
 
-Demographic variables also exhibit notable associations. Maternal age is
+Demographic variables also exhibit some associations. Maternal age is
 negatively associated with infant mortality, with coefficients for age
-groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, respectively. The
-the coefficient for Black mothers in Model 2 is 6.66, reflecting a higher
+groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, relative to group 15-19 respectively. The
+the coefficient for Black mothers in Model 2 is 6.66 relative to Asia monthers, reflecting a higher
 infant mortality compared to other racial groups.
 
-The model highlights key demographic and policy-driven impacts on infant mortality rates.
-The interaction between abortion bans and the post-injunction period is linked to increased mortality, while states enforcing abortion bans show consistently higher rates.
-Younger mothers and Black mothers face disproportionately higher risks, with limited access to healthcare and systemic inequities compounding their vulnerabilities, emphasizing the essential need for targeted health and social interventions.
-Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates.
-These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.
+The model highlights key demographic and policy-driven influences on infant mortality rates.
+The interaction between abortion bans and the post-injunction period is associated with increased mortality, while states with abortion bans consistently exhibit higher rates.
+Younger mothers and Black mothers face significantly elevated risks, highlighting disparities likely driven by limited healthcare access and systemic inequities.
+Conversely, increasing maternal age correlates with lower infant mortality rates, reflecting improved access to resources, better prenatal care, and greater maternal stability.
+These results demonstrate the impact of policy and demographic factors on infant health outcomes.
 
@@
 
-The analysis shows distinct patterns in infant mortality rates across
-various policy and demographic factors. In states where abortion
-remained legal, the post-injunction period (after_injunction1) is linked
-to reductions in infant mortality rates by 0.525 and 0.940 deaths per
-1,000 live births in Models 1 and 2, respectively. However, states with
-abortion bans (abortion_illegal1) exhibit significantly higher infant
-mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
-deaths in Model 2. The DID interaction term (after_injunction1 ×
-abortion_illegal1) shows that abortion bans during the post-injunction
-period results in an additional increase of 0.111 deaths in Model 1 and
-0.280 deaths in Model 2. Demographic factors also play an essential role,
-with older maternal age (25–34 years) reducing infant mortality rates,
-while younger mothers (15–19 years) face elevated risks. Black mothers
-experience disproportionately higher infant mortality rates, with an
-increase of 6.660 deaths per 1,000 live births compared to Asian
-mothers. These findings collectively emphasize the compounded effects of
-policy and demographic disparities on infant health outcomes.
+The analysis reveals notable patterns in infant mortality rates influenced by policy and demographic factors.
+For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in Model 1 and 0.947 in Model 2, compared to the pre-injunction period.
+In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in Model 1 and 1.401 in Model 2, relative to states where abortion remained legal.
+
+The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in Model 1 and 0.280 in Model 2.
+These effects highlight the compounded impact of restrictive abortion policies on infant health outcomes.
+
+Demographic factors also play a significant role.
+Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in Model 2, respectively.
+Regarding maternal race, Black or African American mothers have a significantly higher infant mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian mothers (reference level).
+In contrast, White mothers exhibit a lower infant mortality rate, with a decrease of 2.562 deaths per 1,000 live births compared to Asian mothers.
+
+These findings underscore the interplay of policy and demographic factors, emphasizing the disproportionate burden of restrictive abortion policies and systemic inequities on infant mortality rates.
 
@@
 following the Dobbs decision, combining observational Difference-in-Differences (DID) analysis with qualitative surveys.
-While the DID analysis quantifies policy impacts, qualitative surveys provide context-specific insights into how restrictive abortion policies affect maternal mental health, healthcare access, and pregnancy decisions.
+While the DID analysis quantifies policy impacts, qualitative surveys provide context-specificunderstanding into how restrictive abortion policies affect maternal mental health, healthcare access, and pregnancy decisions.
 These lived experiences help interpret quantitative findings, offering a fuller understanding of the systemic and psychological factors driving observed trends, and supporting evidence-based policy recommendations.

### 2024-11-28 scripts/04-model_data.R
update date error

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Model Infant Death Rate in the U.S. across states in year 2021-2022 using Bayesian DID Model Approach.
 # Author: Diana Shen
-# Date: 20 November 2024 
+# Date: 24 November 2024 
 # Contact: diana.shen@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
add cross-validation

@@
 library(fixest)  
+library(caret)
 ```
@@
 Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
-regression modeling was conducted using `rstanarm` @rstanarm. We use `fixest` @fixest to conduct the event study. Together, these packages supported an efficient, reproducible workflow for
+regression modeling was conducted using `rstanarm` @rstanarm.
+We use `fixest` @fixest to conduct the event study.
+We use `caret` @caret to conduct the cross-validation of model.
+Together, these packages supported an efficient, reproducible workflow for
 analyzing and presenting our findings.
@@
 abortion_illegal is positive and statistically significant (0.111 in
-Model 1 and 0.280 in Model 2), indicating a change in infant mortality
+first model and 0.280 in second model), indicating a change in infant mortality
 rates associated with the combination of abortion bans and the
@@
 For the independent effects, states where abortion is illegal show
-higher infant mortality rates, with coefficients of 3.365 in Model 1 and
-1.400 in Model 2. The post-injunction period alone is associated with
-coefficients of -0.525 in Model 1 and -0.940 in Model 2, indicating a
+higher infant mortality rates, with coefficients of 3.365 in first model and
+1.400 in second model. The post-injunction period alone is associated with
+coefficients of -0.525 in first model and -0.940 in second model, indicating a
 decrease in infant mortality during this time.
@@
 negatively associated with infant mortality, with coefficients for age
-groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, relative to group 15-19 respectively. The
-the coefficient for Black mothers in Model 2 is 6.66 relative to Asia monthers, reflecting a higher
+groups 25–29 and 30–34 of -5.73 and -6.15 in second model, relative to group 15-19 respectively. The
+the coefficient for Black mothers in second model is 6.66 relative to Asia monthers, reflecting a higher
 infant mortality compared to other racial groups.
@@
 The analysis reveals notable patterns in infant mortality rates influenced by policy and demographic factors.
-For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in Model 1 and 0.947 in Model 2, compared to the pre-injunction period.
-In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in Model 1 and 1.401 in Model 2, relative to states where abortion remained legal.
-
-The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in Model 1 and 0.280 in Model 2.
+For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in first model and 0.947 in second model, compared to the pre-injunction period.
+In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in first model and 1.401 in second model, relative to states where abortion remained legal.
+
+The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in first model and 0.280 in second model.
 These effects highlight the compounded impact of restrictive abortion policies on infant health outcomes.
@@
 Demographic factors also play a significant role.
-Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in Model 2, respectively.
+Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in second model, respectively.
 Regarding maternal race, Black or African American mothers have a significantly higher infant mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian mothers (reference level).
@@
 In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model.
-It represents posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data.
-The posterior predictive checks for Model 1 (policy-focused) and Model 2 (policy and demographic factors) illustrate model fit.
+It represents posterior predictive checks for first model and second model, illustrating how well each model captures the observed data.
+The posterior predictive checks for first model (policy-focused) and second model (policy and demographic factors) illustrate model fit.
 In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision.
@@
 The posterior vs.
-prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters.
-In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power.
-Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
-The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.
+prior plots for first model and second model illustrate the role of data in informing the posterior estimates of the model parameters.
+In first model @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power.
+second model @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
+The comparison indicates that second model provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.
 
@@
 Using Bayesian estimation, we derive a posterior distribution for each model coefficient.
-@fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2.
+@fig-area-1 presents the 95% credibility intervals for coefficients in first model, while @fig-area-2 illustrates the same for second model.
 These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.
+
+## Cross Validation
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+set.seed(123)
+
+# Split the data into training and testing sets (80/20 split)
+train_index <- createDataPartition(analysis_data$death_rate, p = 0.8, list = FALSE)
+train_data <- analysis_data[train_index, ]
+test_data <- analysis_data[-train_index, ]
+
+# Fit a Bayesian model on the training set
+model1 <- stan_glmer(
+  death_rate ~ after_injunction * abortion_illegal+ (1 | state),
+  data = train_data,
+  family = gaussian(),
+  prior = normal(0, 2.5),
+  prior_intercept = normal(0, 2.5),
+  prior_aux = exponential(1),
+  iter = 2000, chains = 4, seed = 123
+)
+# Make predictions on the test set
+predictions <- posterior_predict(model1, newdata = test_data)
+predicted_means <- rowMeans(predictions)  # Average over posterior samples
+
+# Calculate RMSE manually
+errors <- test_data$death_rate - predicted_means
+rmse_value <- sqrt(mean(errors^2))
+
+# Output the RMSE
+print(paste("The RMSE is:", round(rmse_value, 2)))
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+# Fit a Bayesian model on the training set
+model2 <- stan_glmer(
+  death_rate ~ after_injunction * abortion_illegal + year_of_death + age_of_mother + mothers_single_race + (1 | state),
+  data = train_data,
+  family = gaussian(),
+  prior = normal(0, 2.5),
+  prior_intercept = normal(0, 2.5),
+  prior_aux = exponential(1),
+  iter = 2000, chains = 4, seed = 123
+)
+# Make predictions on the test set
+predictions <- posterior_predict(model2, newdata = test_data)
+predicted_means <- rowMeans(predictions)  # Average over posterior samples
+
+# Calculate RMSE manually
+errors <- test_data$death_rate - predicted_means
+rmse_value <- sqrt(mean(errors^2))
+
+# Output the RMSE
+print(paste("The RMSE is:", round(rmse_value, 2)))
+```
+The RMSE values from cross-validation indicate that first model and second model perform similarly in predicting infant mortality rates, with RMSE values of 5.07 and 5.06, respectively.
+This similarity suggests that adding demographic factors such as maternal age and race in second model has only a marginal impact on predictive accuracy compared to first model, which primarily relies on policy-related variables like abortion legality and timing of the injunction.
+An average error of approximately 5 deaths per 1,000 live births indicates the magnitude of prediction errors, which may or may not be acceptable depending on the variability in the dataset.
+While second model slightly reduces the error, its added complexity may not justify the trade-off if simplicity is a priority.
+However, if a nuanced understanding of demographic influences is essential, second model provides a modest improvement in accuracy while offering more detailed insights into maternal and infant health disparities.
+The cross-validation process reinforces the reliability of these models by confirming their performance across multiple data splits, reducing the risk of overfitting.
 
@@
 
-The trace plot in @fig-stanareyouokay-1 shows the sampling chains for model 1, indicating good mixing and convergence as the chains overlap and appear stable throughout.
+The trace plot in @fig-stanareyouokay-1 shows the sampling chains for first model, indicating good mixing and convergence as the chains overlap and appear stable throughout.
 This suggests that the model has achieved convergence during sampling.
 The Rhat plot in @fig-stanareyouokay-2 demonstrates that all Rhat values are below 1.05, further confirming convergence and indicating that the model’s posterior samples are reliable.
-@fig-stanareyouokay-3 represents the trace plots for Model 2, showing the posterior samples across four chains.
+@fig-stanareyouokay-3 represents the trace plots for second model, showing the posterior samples across four chains.
 The plots indicate good mixing, as the chains overlap well and stabilize around the same region, suggesting convergence.
-@fig-stanareyouokay-4 is the Rhat diagnostic plot for Model 2.
+@fig-stanareyouokay-4 is the Rhat diagnostic plot for second model.
 All Rhat values are close to 1.00, confirming that the Markov chains have converged, and the posterior estimates are reliable.
-This supports the validity of Model 2’s Bayesian estimation process.
+This supports the validity of second model’s Bayesian estimation process.
 

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update grammar error

@@
 @sec-model outlines the modeling approach, including assumptions and specifications of the Bayesian DID framework.
-In @sec-results, we present the findings, emphasizing key observations.In @sec-results, we present findings on the impact of abortion bans and the post-injunction period, highlighting their association with increased infant mortality.
+In @sec-results, we present the findings, emphasizing key observations.
+In @sec-results, we present findings on the impact of abortion bans and the post-injunction period, highlighting their association with increased infant mortality.
 Key disparities by maternal age and race are discussed, along with model validation metrics like RMSE and $R^2$.
 Finally, @sec-discussion explores the implications of the results and suggests potential directions for future research.
-Additional details are provided in the appendices: [Appendix-@sec-data-details] elaborates on external data sources, [Appendix-@sec-model-details] provides an in-depth explanation of the model, and [Appendix-@sec-survey] introduces the idealized methodology for a related qualitative analysis.
+Additional details are provided in the appendices: [Appendix-@sec-data-details] elaborates on external data sources, [Appendix-@sec-model-details] provides an in-depth explanation of the model, and [Appendix-@sec-survey] introduces the idealized methodology for related qualitative analysis.
 
@@
 We use `fixest` @fixest to conduct the event study.
-We use `caret` @caret to conduct the cross-validation of model.
+We use `caret` @caret to conduct the cross-validation of the model.
 Together, these packages supported an efficient, reproducible workflow for
@@
 Once cleaned, the data is stratified by maternal and infant attributes,
-providing understanding into trends and disparities across demographic
+providing an understanding of trends and disparities across demographic
 groups. Statistical models are employed to estimate associations between
@@
 
-The models are run in R (R Core Team 2023) using the rstanarm package of
+The models are run in R (R Core Team 2023) using the `rstanarm` package of
 @rstanarm. The default priors from `rstanarm` are used for both GLM
@@
 We use a Bayesian approach to enhance the Difference-in-Differences (DID) method, which allows us to incorporate prior knowledge and better account for uncertainty.
-With the rstanarm package, we apply weakly informative priors—such as a Normal prior centered at zero with moderate spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for variability parameters.
+With the `rstanarm` package, we apply weakly informative priors—such as a Normal prior centered at zero with moderate spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for variability parameters.
 These priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data.
@@
 Misspecified priors, data imbalances, and violations of the parallel trends assumption could introduce biases.
-Despite these challenges, the model effectively integrates prior knowledge and addresses uncertainty, providing valuable understanding into policy impacts.
+Despite these challenges, the model effectively integrates prior knowledge and addresses uncertainty, providing a valuable understanding of policy impacts.
 
@@
 For the independent effects, states where abortion is illegal show
-higher infant mortality rates, with coefficients of 3.365 in first model and
-1.400 in second model. The post-injunction period alone is associated with
-coefficients of -0.525 in first model and -0.940 in second model, indicating a
+higher infant mortality rates, with coefficients of 3.365 in the first model and
+1.400 in the second model. The post-injunction period alone is associated with
+coefficients of -0.525 in the first model and -0.940 in the second model, indicating a
 decrease in infant mortality during this time.
@@
 negatively associated with infant mortality, with coefficients for age
-groups 25–29 and 30–34 of -5.73 and -6.15 in second model, relative to group 15-19 respectively. The
-the coefficient for Black mothers in second model is 6.66 relative to Asia monthers, reflecting a higher
+groups 25–29 and 30–34 of -5.73 and -6.15 in the second model, relative to groups 15-19 respectively. The
+the coefficient for Black mothers in the second model is 6.66 relative to Asia mothers, reflecting a higher
 infant mortality compared to other racial groups.
@@
 The analysis reveals notable patterns in infant mortality rates influenced by policy and demographic factors.
-For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in first model and 0.947 in second model, compared to the pre-injunction period.
-In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in first model and 1.401 in second model, relative to states where abortion remained legal.
-
-The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in first model and 0.280 in second model.
+For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in the first model and 0.947 in the second model, compared to the pre-injunction period.
+In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in the first model and 1.401 in the second model, relative to states where abortion remained legal.
+
+The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in the first model and 0.280 in the second model.
 These effects highlight the compounded impact of restrictive abortion policies on infant health outcomes.
@@
 Demographic factors also play a significant role.
-Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in second model, respectively.
+Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in the second model, respectively.
 Regarding maternal race, Black or African American mothers have a significantly higher infant mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian mothers (reference level).
@@
 
-Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data correspond to the period after June 2022 (post-Dobbs decision) and `abortion_illegal`, denoting whether the state had restrictive abortion laws.
+Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data correspond to the period after June 2022 (post-Dobbs decision), and `abortion_illegal`, denoting whether the state had restrictive abortion laws.
 Death rate reliability was assessed, and unreliable rates were flagged for transparency.
@@
 The data utilized in this study was sourced from the CDC WONDER Online Database.
-Access to the data and its use complies with the terms outlined on the [CDC WONDER Data Use Agreement](https://wonder.cdc.gov/datause.html).
+Access to the data and its use complies with the terms outlined in the [CDC WONDER Data Use Agreement](https://wonder.cdc.gov/datause.html).
 Specifically, the data was used for academic research purposes, with acknowledgment of the Centers for Disease Control and Prevention (CDC) as the original data provider.
@@
 prior plots for first model and second model illustrate the role of data in informing the posterior estimates of the model parameters.
-In first model @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power.
+In the first model @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power.
 second model @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
-The comparison indicates that second model provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.
+The comparison indicates that the second model provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.
 
@@
 Using Bayesian estimation, we derive a posterior distribution for each model coefficient.
-@fig-area-1 presents the 95% credibility intervals for coefficients in first model, while @fig-area-2 illustrates the same for second model.
+@fig-area-1 presents the 95% credibility intervals for coefficients in the first model, while @fig-area-2 illustrates the same for the second model.
 These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.
@@
 ```
-The RMSE values from cross-validation indicate that first model and second model perform similarly in predicting infant mortality rates, with RMSE values of 5.07 and 5.06, respectively.
-This similarity suggests that adding demographic factors such as maternal age and race in second model has only a marginal impact on predictive accuracy compared to first model, which primarily relies on policy-related variables like abortion legality and timing of the injunction.
+The RMSE values from cross-validation indicate that the first model and second model perform similarly in predicting infant mortality rates, with RMSE values of 5.07 and 5.06, respectively.
+This similarity suggests that adding demographic factors such as maternal age and race in the second model has only a marginal impact on predictive accuracy compared to the first model, which primarily relies on policy-related variables like abortion legality and timing of the injunction.
 An average error of approximately 5 deaths per 1,000 live births indicates the magnitude of prediction errors, which may or may not be acceptable depending on the variability in the dataset.
-While second model slightly reduces the error, its added complexity may not justify the trade-off if simplicity is a priority.
-However, if a nuanced understanding of demographic influences is essential, second model provides a modest improvement in accuracy while offering more detailed insights into maternal and infant health disparities.
+While the second model slightly reduces the error, its added complexity may not justify the trade-off if simplicity is a priority.
+However, if a nuanced understanding of demographic influences is essential, the second model provides a modest improvement in accuracy while offering more detailed insights into maternal and infant health disparities.
 The cross-validation process reinforces the reliability of these models by confirming their performance across multiple data splits, reducing the risk of overfitting.
@@
 
-The trace plot in @fig-stanareyouokay-1 shows the sampling chains for first model, indicating good mixing and convergence as the chains overlap and appear stable throughout.
+The trace plot in @fig-stanareyouokay-1 shows the sampling chains for the first model, indicating good mixing and convergence as the chains overlap and appear stable throughout.
 This suggests that the model has achieved convergence during sampling.
 The Rhat plot in @fig-stanareyouokay-2 demonstrates that all Rhat values are below 1.05, further confirming convergence and indicating that the model’s posterior samples are reliable.
-@fig-stanareyouokay-3 represents the trace plots for second model, showing the posterior samples across four chains.
+@fig-stanareyouokay-3 represents the trace plots for the second model, showing the posterior samples across four chains.
 The plots indicate good mixing, as the chains overlap well and stabilize around the same region, suggesting convergence.
-@fig-stanareyouokay-4 is the Rhat diagnostic plot for second model.
+@fig-stanareyouokay-4 is the Rhat diagnostic plot for the second model.
 All Rhat values are close to 1.00, confirming that the Markov chains have converged, and the posterior estimates are reliable.
-This supports the validity of second model’s Bayesian estimation process.
+This supports the validity of the second model’s Bayesian estimation process.
 
@@
 following the Dobbs decision, combining observational Difference-in-Differences (DID) analysis with qualitative surveys.
-While the DID analysis quantifies policy impacts, qualitative surveys provide context-specificunderstanding into how restrictive abortion policies affect maternal mental health, healthcare access, and pregnancy decisions.
+While the DID analysis quantifies policy impacts, qualitative surveys provide a context-specific understanding of how restrictive abortion policies affect maternal mental health, healthcare access, and pregnancy decisions.
 These lived experiences help interpret quantitative findings, offering a fuller understanding of the systemic and psychological factors driving observed trends, and supporting evidence-based policy recommendations.
@@
 and systemic impacts of restrictive abortion laws on maternal health and infant outcomes.
-By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences, and explore the role of maternal stress and healthcare barriers in adverse outcomes.
+By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences and explore the role of maternal stress and healthcare barriers in adverse outcomes.
 Understanding how maternal mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the systemic factors contributing to higher infant mortality rates.

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update pdf of paper

@@
 #| echo: false
-#| eval: true
-#| message: false
-#| warning: false
+#| eval: false
+#| message: false
+#| warning: false
+#| include: false
+
 set.seed(123)
-
 # Split the data into training and testing sets (80/20 split)
@@
 #| warning: false
-# Fit a Bayesian model on the training set
+#| include: false
+
+set.seed(123)
+train_index <- createDataPartition(analysis_data$death_rate, p = 0.8, list = FALSE)
+train_data <- analysis_data[train_index, ]
+test_data <- analysis_data[-train_index, ]
 model2 <- stan_glmer(

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with figure title

@@
 manipulation, modeling, and visualization. The `tidyverse` package
-@tidyverse was central to our workflow, providing tools for data
-wrangling and analysis, while `arrow` @arrow efficiently managed parquet
-files for larger datasets. We used `modelsummary` package @modelsummary
+[@tidyverse] was central to our workflow, providing tools for data
+wrangling and analysis, while `arrow` [@arrow] efficiently managed parquet
+files for larger datasets. We used `modelsummary` package [@modelsummary]
 to generate clean and interpretable summaries of our regression
-models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
+models.We used `knitr` [@knitr] and `kableExtra` [@kableExtra] to generate
 clean tables and reproducible reports, and `usmap` for geographical
-visualizations. We use `here` @here package simplifies the management of
+visualizations. We use `here` [@here] package simplifies the management of
 file paths, ensuring a reproducible workflow by creating a consistent
@@
 system or working directory. Data cleaning was facilitated by `janitor`
-@janitor, while `lubridate` @lubridate handled date-time variables.
-Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
-regression modeling was conducted using `rstanarm` @rstanarm.
-We use `fixest` @fixest to conduct the event study.
-We use `caret` @caret to conduct the cross-validation of the model.
+[@janitor], while `lubridate` [@lubridate] handled date-time variables.
+Visualization tasks were enhanced with `ggplot2` [@ggplot2], and Bayesian
+regression modeling was conducted using `rstanarm` [@rstanarm].
+We use `fixest` [@fixest] to conduct the event study.
+We use `caret` [@caret] to conduct the cross-validation of the model.
 Together, these packages supported an efficient, reproducible workflow for
@@
 #|   Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United   States: 
-#|   Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May) as the Reference Period.
+#|   Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May 2022) as the Reference Period.
 #| warning: false

### 2024-11-29 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with peer review

@@
 ---
-title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
-subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
+title: "Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis"
+subtitle: "Infant death rates rose by 0.285 per 1,000 live births in abortion-restricted states compared to protected states."
 author: 
   - Diana Shen
-thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
+thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion](https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
 date: today
 date-format: long
-abstract: "This study examines the impact of the Dobbs decision and subsequent abortion bans on infant mortality rates in the United States, focusing on data from 2021 to 2022.
-Using a Difference-in-Differences (DID) approach, we analyze trends in states where abortion remained legal compared to those where it became illegal after June 2022.
-Our results indicate that states enforcing abortion bans experienced an increase of 0.285 infant deaths per 1,000 live births, with compounded effects observed during the post-injunction period.
-Maternal age and race also emerged as significant predictors, with Black mothers facing disproportionately higher infant mortality rates.
-These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities.
-By utilizing recent, high-quality data and rigorous modeling, this study provides essential understanding into the intersection of reproductive rights and infant health outcomes."
+abstract: "This study investigates the impact of the Dobbs decision and abortion bans on U.S.
+infant mortality rates using 2021–2022 data.
+A Difference-in-Differences (DID) analysis shows a 0.285 increase in infant deaths per 1,000 live births in states enforcing bans, with amplified effects post-injunction.
+Maternal age and race are key factors, with Black mothers experiencing higher mortality rates.
+These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities.
+This study uses latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
 format: pdf
@@
 #| message: false
-
 library(tidyverse)
@@
 Court’s decision in Dobbs v. Jackson Women’s Health Organization, ended
-nearly five decades of constitutional protection for abortion rights.
-This landmark ruling shifted authority over abortion regulations to
-individual states, creating a patchwork of laws nationwide. The decision
-reignited debates about reproductive rights and public health,
-highlighting disparities in access to abortion services and their
-effects on health outcomes, especially in states with restrictive
-policies.
+nearly five decades of constitutional protection for abortion rights in
+the United States. This landmark ruling shifted authority over abortion
+regulations to individual states, creating a patchwork of laws
+nationwide. The decision reignited debates about reproductive rights and
+public health, highlighting disparities in access to abortion services
+and their effects on health outcomes, especially in states with
+restrictive policies.
 
@@
 evidence on how restrictive abortion policies disproportionately affect
-vulnerable populations, emphasizing the need for evidence-based policies to
-mitigate these adverse outcomes.
-
-The structure of this paper is as follows: @sec-data details the data sources and variables used in the analysis.
-@sec-model outlines the modeling approach, including assumptions and specifications of the Bayesian DID framework.
-In @sec-results, we present the findings, emphasizing key observations.
-In @sec-results, we present findings on the impact of abortion bans and the post-injunction period, highlighting their association with increased infant mortality.
-Key disparities by maternal age and race are discussed, along with model validation metrics like RMSE and $R^2$.
-Finally, @sec-discussion explores the implications of the results and suggests potential directions for future research.
-Additional details are provided in the appendices: [Appendix-@sec-data-details] elaborates on external data sources, [Appendix-@sec-model-details] provides an in-depth explanation of the model, and [Appendix-@sec-survey] introduces the idealized methodology for related qualitative analysis.
+vulnerable populations, emphasizing the need for evidence-based policies
+to mitigate these adverse outcomes.
+
+The structure of this paper is as follows: @sec-data details the data
+sources and variables used in the analysis. @sec-model outlines the
+modeling approach, including assumptions and specifications of the
+Bayesian DID framework. In @sec-results, we present the findings,
+emphasizing key observations. In @sec-results, we present findings on
+the impact of abortion bans and the post-injunction period, highlighting
+their association with increased infant mortality. Key disparities by
+maternal age and race are discussed, along with model validation metrics
+like RMSE and $R^2$. Finally, @sec-discussion explores the implications
+of the results and suggests potential directions for future research.
+Additional details are provided in the appendices: [Appendix-
+@sec-data-details] elaborates on external data sources, [Appendix-
+@sec-model-details] provides an in-depth explanation of the model, and
+[Appendix- @sec-survey] introduces the idealized methodology for related
+qualitative analysis.
 
@@
 [@tidyverse] was central to our workflow, providing tools for data
-wrangling and analysis, while `arrow` [@arrow] efficiently managed parquet
-files for larger datasets. We used `modelsummary` package [@modelsummary]
-to generate clean and interpretable summaries of our regression
-models.We used `knitr` [@knitr] and `kableExtra` [@kableExtra] to generate
-clean tables and reproducible reports, and `usmap` for geographical
-visualizations. We use `here` [@here] package simplifies the management of
-file paths, ensuring a reproducible workflow by creating a consistent
-point of reference for file locations, regardless of the operating
-system or working directory. Data cleaning was facilitated by `janitor`
-[@janitor], while `lubridate` [@lubridate] handled date-time variables.
-Visualization tasks were enhanced with `ggplot2` [@ggplot2], and Bayesian
-regression modeling was conducted using `rstanarm` [@rstanarm].
-We use `fixest` [@fixest] to conduct the event study.
-We use `caret` [@caret] to conduct the cross-validation of the model.
-Together, these packages supported an efficient, reproducible workflow for
-analyzing and presenting our findings.
+wrangling and analysis, while `arrow` [@arrow] efficiently managed
+parquet files for larger datasets. We used `modelsummary` package
+[@modelsummary] to generate clean and interpretable summaries of our
+regression models.We used `knitr` [@knitr] and `kableExtra`
+[@kableExtra] to generate clean tables and reproducible reports, and
+`usmap` for geographical visualizations. We use `here` [@here] package
+simplifies the management of file paths, ensuring a reproducible
+workflow by creating a consistent point of reference for file locations,
+regardless of the operating system or working directory. Data cleaning
+was facilitated by `janitor` [@janitor], while `lubridate` [@lubridate]
+handled date-time variables. Visualization tasks were enhanced with
+`ggplot2` [@ggplot2], and Bayesian regression modeling was conducted
+using `rstanarm` [@rstanarm]. We use `fixest` [@fixest] to conduct the
+event study. We use `caret` [@caret] to conduct the cross-validation of
+the model. Together, these packages supported an efficient, reproducible
+workflow for analyzing and presenting our findings.
 
@@
 impact of abortion policies and demographic factors. The CDC’s Linked
-Birth/Infant Death Records database provides detailed data,
-including maternal and infant characteristics, to support this study.
-Variables such as maternal age, race, education, birth weight,
-gestational age, and prenatal care are included to ensure a detailed
-examination of factors influencing infant health outcomes.
+Birth/Infant Death Records database provides detailed data, including
+maternal and infant characteristics, to support this study. Variables
+such as maternal age, race, education, birth weight, gestational age,
+and prenatal care are included to ensure a detailed examination of
+factors influencing infant health outcomes.
 
@@
 confounders such as maternal characteristics. This structured approach
-transforms complex datasets into actionable understanding, highlighting on
-the health implications of varying abortion laws across the U.S. By
-utilizing detailed datasets and robust statistical techniques,
-this study offers a detailed understanding of the intersection between
+transforms complex datasets into actionable understanding, highlighting
+on the health implications of varying abortion laws across the U.S. By
+utilizing detailed datasets and robust statistical techniques, this
+study offers a detailed understanding of the intersection between
 policy, demographics, and infant health outcomes, informing the public
@@
 analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
-  group_by(illegal_after_injunction)|>
+  group_by(illegal_after_injunction)|>rename(`Abortion Restriction Post-Overturn`=illegal_after_injunction)|>
   summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))
@@
 `abortion_illegal` and `after_injunction`. The resulting interaction
-term equals 1 when abortion is illegal the time is after the
-injunction, and 0 otherwise. This term is included in the analysis to
-evaluate the joint impact of these factors on infant death rates.
+term equals 1 when abortion is illegal the time is after the injunction,
+and 0 otherwise. This term is included in the analysis to evaluate the
+joint impact of these factors on infant death rates.
 
@@
 
-Our analysis includes three key control variables: maternal age (`age_of_mother`), race (`mothers_single_race`), and year of death (`year_of_death`), to account for demographic and temporal differences.
+Our analysis includes three key control variables: maternal age
+(`age_of_mother`), race (`mothers_single_race`), and year of death
+(`year_of_death`), to account for demographic and temporal differences.
 
@@
   select(year_of_death,age_of_mother,mothers_single_race) |> 
+  rename(`Year of Death` = year_of_death,
+         `Age of Mother` = age_of_mother,
+         `Mother's Single Race` = mothers_single_race)|>
   modelsummary::datasummary_skim()
@@
 
-@tbl-summary shows that 50.7% of infant deaths occurred in 2022, slightly higher than 2021 (49.3%).
-The majority of cases were among mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and 20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
-By race, White mothers accounted for the highest percentage (75.7%), followed by Black or African American mothers (24.0%), and Asian mothers (0.3%).
+@tbl-summary shows that 50.7% of infant deaths occurred in 2022,
+slightly higher than 2021 (49.3%). The majority of cases were among
+mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
+20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
+By race, White mothers accounted for the highest percentage (75.7%),
+followed by Black or African American mothers (24.0%), and Asian mothers
+(0.3%).
 
@@
 
-@fig-choromap is a choropleth map that visualizes the average infant death
-rate per 1,000 live births across U.S. states, with data categorized by
-color intensity. States with higher infant death rates are shown in
-lighter shades (yellow), while states with lower rates are represented
-in darker shades (purple). States with missing data are indicated in
-grey. The map highlights disparities in infant mortality rates, with
-some states, particularly in the southeastern U.S., exhibiting notably
-higher rates compared to others. States like Alabama and Mississippi
-show the highest rates, while states such as California and New York
-exhibit lower rates.
+@fig-choromap is a choropleth map that visualizes the average infant
+death rate per 1,000 live births across U.S. states, with data
+categorized by color intensity. States with higher infant death rates
+are shown in lighter shades (yellow), while states with lower rates are
+represented in darker shades (purple). States with missing data are
+indicated in grey. The map highlights disparities in infant mortality
+rates, with some states, particularly in the southeastern U.S.,
+exhibiting notably higher rates compared to others. States like Alabama
+and Mississippi show the highest rates, while states such as California
+and New York exhibit lower rates.
 
@@
 -   **Time Indicator `after_injunction`**: A binary variable indicating
-    the time period: 0 for the period before the injunction was implemented.
-    1 for the period after the injunction was implemented.
+    the time period: 0 for the period before the injunction was
+    implemented. 1 for the period after the injunction was implemented.
 
@@
 identifies changes in infant death rates that can be attributed to the
-policy intervention. Additional details, validation of model assumption, and diagnostics supporting the
-model’s implementation is provided in [Appendix- @sec-model-details].
+policy intervention. Additional details, validation of model assumption,
+and diagnostics supporting the model’s implementation is provided in
+[Appendix- @sec-model-details].
 
@@
 In our analysis, we employ two models. The first model includes only the
-predictor variables and the random effect for the state. The second model
-builds on the first by incorporating additional control variables to
-account for potential confounding factors.
+predictor variables and the random effect for the state. The second
+model builds on the first by incorporating additional control variables
+to account for potential confounding factors.
 
@@
 
-The models are run in R (R Core Team 2023) using the `rstanarm` package of
-@rstanarm. The default priors from `rstanarm` are used for both GLM
+The models are run in R (R Core Team 2023) using the `rstanarm` package
+of @rstanarm. The default priors from `rstanarm` are used for both GLM
 Bayesian models. We run the model in R [@citeR] using the `rstanarm`
@@
 
-For our analysis, we employ Bayesian regression models to estimate infant death rates in states where abortion injunctions were implemented and states where abortion remains legal.
-The causal effect of the abortion injunction is measured using the interaction term between after_injunction and abortion_illegal.
-To ensure consistency and comparability, both models share the same predictors, capturing the determinants of infant mortality.
-
-One model includes control variables— `age_of_mother`, `mothers_single_race`, and `year_of_death` to adjust for confounders that may influence infant death rates.
-This adjustment addresses compositional differences assumed in the DID framework and ensures unbiased estimates of the intervention’s effect.
-The second model excludes these controls to focus on the overall intervention impact, avoid overfitting, and prioritize simplicity where variability is adequately accounted for by design or random effects.
-By fitting two models, we compare outcomes to refine estimates and evaluate the direct impact of the intervention.
+For our analysis, we employ Bayesian regression models to estimate
+infant death rates in states where abortion injunctions were implemented
+and states where abortion remains legal. The causal effect of the
+abortion injunction is measured using the interaction term between
+after_injunction and abortion_illegal. To ensure consistency and
+comparability, both models share the same predictors, capturing the
+determinants of infant mortality.
+
+One model includes control variables— `age_of_mother`,
+`mothers_single_race`, and `year_of_death` to adjust for confounders
+that may influence infant death rates. This adjustment addresses
+compositional differences assumed in the DID framework and ensures
+unbiased estimates of the intervention’s effect. The second model
+excludes these controls to focus on the overall intervention impact,
+avoid overfitting, and prioritize simplicity where variability is
+adequately accounted for by design or random effects. By fitting two
+models, we compare outcomes to refine estimates and evaluate the direct
+impact of the intervention.
 
@@
 
-Our models include state as a random effect to account for regional variations in infant mortality that might introduce bias.
-Using a random effect rather than a fixed effect captures variability across states without estimating individual effects, modeling unobserved heterogeneity—such as differences in healthcare systems and socioeconomic conditions—through a single variance term.
-This approach avoids overfitting, enhances efficiency, and ensures generalizability.
-By adjusting for unmeasured state-specific factors, the models strike a balance between parsimony and interpretability, making them well-suited for our analysis.
+Our models include state as a random effect to account for regional
+variations in infant mortality that might introduce bias. Using a random
+effect rather than a fixed effect captures variability across states
+without estimating individual effects, modeling unobserved
+heterogeneity—such as differences in healthcare systems and
+socioeconomic conditions—through a single variance term. This approach
+avoids overfitting, enhances efficiency, and ensures generalizability.
+By adjusting for unmeasured state-specific factors, the models strike a
+balance between parsimony and interpretability, making them well-suited
+for our analysis.
 
@@
 
-We selected the Difference-in-Differences (DID) framework because randomization is infeasible given the pre-existing differences between states in healthcare systems, socioeconomic conditions, and cultural factors.
-DID utilizes time and group-level variations, controlling for unobserved, time-invariant confounders.
-The parallel trends assumption ensures that untreated states provide a valid counterfactual for treated states.
-Interaction terms, such as after_injunction and abortion_illegal, allow for direct measurement of policy effects, while random effects adjust for regional variability.
-
-Alternative models, such as Fixed Effects, Propensity Score Matching (PSM), and Interrupted Time Series (ITS), were considered but deemed less suitable.
-Fixed Effects cannot estimate the effects of variables that do not vary over time within a unit.
-PSM balances observed covariates but does not account for temporal dynamics.
-ITS analyzes trends within a single group but lacks a comparison group.
-DID’s ability to integrate temporal and group-level differences makes it the most robust and appropriate framework for estimating the impact of abortion bans on infant mortality.
+We selected the Difference-in-Differences (DID) framework because
+randomization is infeasible given the pre-existing differences between
+states in healthcare systems, socioeconomic conditions, and cultural
+factors. DID utilizes time and group-level variations, controlling for
+unobserved, time-invariant confounders. The parallel trends assumption
+ensures that untreated states provide a valid counterfactual for treated
+states. Interaction terms, such as after_injunction and
+abortion_illegal, allow for direct measurement of policy effects, while
+random effects adjust for regional variability.
+
+Alternative models, such as Fixed Effects, Propensity Score Matching
+(PSM), and Interrupted Time Series (ITS), were considered but deemed
+less suitable. Fixed Effects cannot estimate the effects of variables
+that do not vary over time within a unit. PSM balances observed
+covariates but does not account for temporal dynamics. ITS analyzes
+trends within a single group but lacks a comparison group. DID’s ability
+to integrate temporal and group-level differences makes it the most
+robust and appropriate framework for estimating the impact of abortion
+bans on infant mortality.
 
@@
 
-We use a Bayesian approach to enhance the Difference-in-Differences (DID) method, which allows us to incorporate prior knowledge and better account for uncertainty.
-With the `rstanarm` package, we apply weakly informative priors—such as a Normal prior centered at zero with moderate spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for variability parameters.
-These priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data.
-This approach helps us center estimates realistically, account for hidden differences between groups, and produce reliable results while assuming trends between treated and control groups are similar before treatment.
+We use a Bayesian approach to enhance the Difference-in-Differences
+(DID) method, which allows us to incorporate prior knowledge and better
+account for uncertainty. With the `rstanarm` package, we apply weakly
+informative priors—such as a Normal prior centered at zero with moderate
+spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential
+prior (rate = 1) for variability parameters. These priors automatically
+adjust to the scale of the data, improving model stability and ensuring
+they complement rather than dominate the observed data. This approach
+helps us center estimates realistically, account for hidden differences
+between groups, and produce reliable results while assuming trends
+between treated and control groups are similar before treatment.
 
@@
 
-Our Bayesian DID model offers a robust approach to examining the impact of abortion bans on infant mortality but has limitations.
-Misspecified priors, data imbalances, and violations of the parallel trends assumption could introduce biases.
-Despite these challenges, the model effectively integrates prior knowledge and addresses uncertainty, providing a valuable understanding of policy impacts.
+Our Bayesian DID model offers a robust approach to examining the impact
+of abortion bans on infant mortality but has limitations. Misspecified
+priors, data imbalances, and violations of the parallel trends
+assumption could introduce biases. Despite these challenges, the model
+effectively integrates prior knowledge and addresses uncertainty,
+providing a valuable understanding of policy impacts.
 
@@
 abortion_illegal is positive and statistically significant (0.111 in
-first model and 0.280 in second model), indicating a change in infant mortality
-rates associated with the combination of abortion bans and the
+first model and 0.280 in second model), indicating a change in infant
+mortality rates associated with the combination of abortion bans and the
 post-injunction period.
@@
 For the independent effects, states where abortion is illegal show
-higher infant mortality rates, with coefficients of 3.365 in the first model and
-1.400 in the second model. The post-injunction period alone is associated with
-coefficients of -0.525 in the first model and -0.940 in the second model, indicating a
-decrease in infant mortality during this time.
+higher infant mortality rates, with coefficients of 3.365 in the first
+model and 1.400 in the second model. The post-injunction period alone is
+associated with coefficients of -0.525 in the first model and -0.940 in
+the second model, indicating a decrease in infant mortality during this
+time.
 
@@
 negatively associated with infant mortality, with coefficients for age
-groups 25–29 and 30–34 of -5.73 and -6.15 in the second model, relative to groups 15-19 respectively. The
-the coefficient for Black mothers in the second model is 6.66 relative to Asia mothers, reflecting a higher
+groups 25–29 and 30–34 of -5.73 and -6.15 in the second model, relative
+to groups 15-19 respectively. The the coefficient for Black mothers in
+the second model is 6.66 relative to Asia mothers, reflecting a higher
 infant mortality compared to other racial groups.
 
-The model highlights key demographic and policy-driven influences on infant mortality rates.
-The interaction between abortion bans and the post-injunction period is associated with increased mortality, while states with abortion bans consistently exhibit higher rates.
-Younger mothers and Black mothers face significantly elevated risks, highlighting disparities likely driven by limited healthcare access and systemic inequities.
-Conversely, increasing maternal age correlates with lower infant mortality rates, reflecting improved access to resources, better prenatal care, and greater maternal stability.
-These results demonstrate the impact of policy and demographic factors on infant health outcomes.
+The model highlights key demographic and policy-driven influences on
+infant mortality rates. The interaction between abortion bans and the
+post-injunction period is associated with increased mortality, while
+states with abortion bans consistently exhibit higher rates. Younger
+mothers and Black mothers face significantly elevated risks,
+highlighting disparities likely driven by limited healthcare access and
+systemic inequities. Conversely, increasing maternal age correlates with
+lower infant mortality rates, reflecting improved access to resources,
+better prenatal care, and greater maternal stability. These results
+demonstrate the impact of policy and demographic factors on infant
+health outcomes.
 
@@
 
-The analysis reveals notable patterns in infant mortality rates influenced by policy and demographic factors.
-For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in the first model and 0.947 in the second model, compared to the pre-injunction period.
-In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in the first model and 1.401 in the second model, relative to states where abortion remained legal.
-
-The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in the first model and 0.280 in the second model.
-These effects highlight the compounded impact of restrictive abortion policies on infant health outcomes.
-
-Demographic factors also play a significant role.
-Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in the second model, respectively.
-Regarding maternal race, Black or African American mothers have a significantly higher infant mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian mothers (reference level).
-In contrast, White mothers exhibit a lower infant mortality rate, with a decrease of 2.562 deaths per 1,000 live births compared to Asian mothers.
-
-These findings underscore the interplay of policy and demographic factors, emphasizing the disproportionate burden of restrictive abortion policies and systemic inequities on infant mortality rates.
+The analysis reveals notable patterns in infant mortality rates
+influenced by policy and demographic factors. For states where abortion
+remained legal, the post-injunction period (after_injunction1) is
+associated with reductions in infant mortality rates, with a decrease of
+0.525 deaths per 1,000 live births in the first model and 0.947 in the
+second model, compared to the pre-injunction period. In contrast, states
+where abortion became illegal (abortion_illegal1) experienced
+significantly higher infant mortality rates, with increases of 3.365
+deaths in the first model and 1.401 in the second model, relative to
+states where abortion remained legal.
+
+The Difference-in-Differences (DID) interaction term (after_injunction1
+× abortion_illegal1) indicates that abortion bans during the
+post-injunction period resulted in an additional increase of 0.111
+deaths per 1,000 live births in the first model and 0.280 in the second
+model. These effects highlight the compounded impact of restrictive
+abortion policies on infant health outcomes.
+
+Demographic factors also play a significant role. Compared to mothers
+aged 15–19 years (reference level), older maternal age groups (20–24,
+25–29, and 30–34 years) are associated with progressively lower infant
+mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per
+1,000 live births in the second model, respectively. Regarding maternal
+race, Black or African American mothers have a significantly higher
+infant mortality rate, with an increase of 6.604 deaths per 1,000 live
+births compared to Asian mothers (reference level). In contrast, White
+mothers exhibit a lower infant mortality rate, with a decrease of 2.562
+deaths per 1,000 live births compared to Asian mothers.
+
+These findings underscore the interplay of policy and demographic
+factors, emphasizing the disproportionate burden of restrictive abortion
+policies and systemic inequities on infant mortality rates.
 
@@
 
-Addressing Public Health and Long-Term Implications of Restrictive Abortion Laws
-
-Restrictive abortion policies significantly impact maternal and infant health, as evidenced by an increase of 0.285 infant deaths per 1,000 live births in states enforcing abortion bans during the post-injunction period.
-These policies exacerbate systemic issues such as inadequate prenatal care, gaps in healthcare access, and socioeconomic inequities, disproportionately affecting marginalized communities.
-Over time, the consequences may include higher rates of untreated maternal mental health issues, delayed or insufficient prenatal care, and elevated infant mortality, creating lasting health challenges for families and communities.
-Addressing these disparities requires targeted interventions, such as expanding maternal mental health services, improving prenatal and postnatal care in underserved areas, ensuring culturally competent healthcare, and advocating for equitable policies.
-Investments in community-based support systems, enhanced healthcare delivery, and robust data infrastructure are essential to monitor and mitigate the long-term impacts of these policies, fostering sustainable improvements in maternal and infant health outcomes.
+Restrictive abortion policies significantly impact maternal and infant
+health, as evidenced by an increase of 0.285 infant deaths per 1,000
+live births in states enforcing abortion bans during the post-injunction
+period. These policies exacerbate systemic issues such as inadequate
+prenatal care, gaps in healthcare access, and socioeconomic inequities,
+disproportionately affecting marginalized communities. Over time, the
+consequences may include higher rates of untreated maternal mental
+health issues, delayed or insufficient prenatal care, and elevated
+infant mortality, creating lasting health challenges for families and
+communities. Addressing these disparities requires targeted
+interventions, such as expanding maternal mental health services,
+improving prenatal and postnatal care in underserved areas, ensuring
+culturally competent healthcare, and advocating for equitable policies.
+Investments in community-based support systems, enhanced healthcare
+delivery, and robust data infrastructure are essential to monitor and
+mitigate the long-term impacts of these policies, fostering sustainable
+improvements in maternal and infant health outcomes.
 
@@
 
-A key limitation of our analysis is the temporal and demographic scope of the dataset.
-The data includes only infant deaths under 1 year of age occurring within the United States to U.S.
-residents during 2021–2022.
-This timeframe does not capture the full year of infant mortality outcomes following the Dobbs decision, as infants who died in 2023 are not included.
-Furthermore, the dataset’s focus on infants under 1 year of age may exclude cases that reflect longer-term health impacts.
-These limitations could lead to an underestimation of the full effects of abortion bans.
-As no publicly available dataset currently provides complete 2023 data or longer-term health outcomes, future research should incorporate more detailed and updated datasets to provide a clearer and more complete analysis of the long-term trends and implications of restrictive abortion policies.
+A key limitation of our analysis is the temporal and demographic scope
+of the dataset. The data includes only infant deaths under 1 year of age
+occurring within the United States to U.S. residents during 2021–2022.
+This timeframe does not capture the full year of infant mortality
+outcomes following the Dobbs decision, as infants who died in 2023 are
+not included. Furthermore, the dataset’s focus on infants under 1 year
+of age may exclude cases that reflect longer-term health impacts. These
+limitations could lead to an underestimation of the full effects of
+abortion bans. As no publicly available dataset currently provides
+complete 2023 data or longer-term health outcomes, future research
+should incorporate more detailed and updated datasets to provide a
+clearer and more complete analysis of the long-term trends and
+implications of restrictive abortion policies.
 
@@
 
-While this study provides meaningful findings, several limitations should
-be addressed in future research. The analysis relies on aggregate
+While this study provides meaningful findings, several limitations
+should be addressed in future research. The analysis relies on aggregate
 state-level data, which may obscure localized effects and intra-state
@@
 
-The raw dataset, consisting of U.S.
-infant mortality records from 2021 and 2022, was cleaned and pre-processed to ensure its suitability for analysis.
-The dataset was first imported from a tab-separated file, with irregular lines handled appropriately.
-Variables were standardized using `janitor::clean_names()` to ensure consistent naming conventions.
-Key variables such as state, maternal age, maternal race, year of death, month, and infant death rates were selected for analysis, and missing values were removed.
-
-Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data correspond to the period after June 2022 (post-Dobbs decision), and `abortion_illegal`, denoting whether the state had restrictive abortion laws.
-Death rate reliability was assessed, and unreliable rates were flagged for transparency.
-The death rate column was cleaned to remove annotations and converted to numeric format for analysis.
-A date variable was created by combining the year and month columns, providing a timeline for analysis.
-Finally, a date variable was created by combining the year and month variables.
-The cleaned dataset was saved in CSV and Parquet formats to ensure compatibility with analytical workflows.
+The raw dataset, consisting of U.S. infant mortality records from 2021
+and 2022, was cleaned and pre-processed to ensure its suitability for
+analysis. The dataset was first imported from a tab-separated file, with
+irregular lines handled appropriately. Variables were standardized using
+`janitor::clean_names()` to ensure consistent naming conventions. Key
+variables such as state, maternal age, maternal race, year of death,
+month, and infant death rates were selected for analysis, and missing
+values were removed.
+
+Additional variables were created to facilitate analysis:
+`after_injunction`, indicating whether the data correspond to the period
+after June 2022 (post-Dobbs decision), and `abortion_illegal`, denoting
+whether the state had restrictive abortion laws. Death rate reliability
+was assessed, and unreliable rates were flagged for transparency. The
+death rate column was cleaned to remove annotations and converted to
+numeric format for analysis. A date variable was created by combining
+the year and month columns, providing a timeline for analysis. Finally,
+a date variable was created by combining the year and month variables.
+The cleaned dataset was saved in CSV and Parquet formats to ensure
+compatibility with analytical workflows.
 
@@
 
-The data utilized in this study was sourced from the CDC WONDER Online Database.
-Access to the data and its use complies with the terms outlined in the [CDC WONDER Data Use Agreement](https://wonder.cdc.gov/datause.html).
-Specifically, the data was used for academic research purposes, with acknowledgment of the Centers for Disease Control and Prevention (CDC) as the original data provider.
-The CDC, however, does not assume responsibility for any analyses, interpretations, or conclusions drawn from the data by the authors of this study.
+The data utilized in this study was sourced from the CDC WONDER Online
+Database. Access to the data and its use complies with the terms
+outlined in the [CDC WONDER Data Use
+Agreement](https://wonder.cdc.gov/datause.html). Specifically, the data
+was used for academic research purposes, with acknowledgment of the
+Centers for Disease Control and Prevention (CDC) as the original data
+provider. The CDC, however, does not assume responsibility for any
+analyses, interpretations, or conclusions drawn from the data by the
+authors of this study.
 
@@
 
-@fig-eventstudy depicts the estimated effects of abortion bans on infant mortality rates over time, with relative months on the x-axis and coefficient estimates with 95% confidence intervals on the y-axis.
-Before the implementation of abortion bans ($t<0$), the coefficients are centered around zero, indicating no significant differences in infant mortality rates between treated and control states, supporting the parallel trends assumption.
-Post-ban ($t\geq0$), the estimates show a slight upward trend, suggesting a potential increase in infant mortality rates in restricted states; however, the wide confidence intervals, which overlap with zero, indicate that these effects are not statistically significant.
-Overall, the results do not provide conclusive evidence of a causal relationship between abortion bans and changes in infant mortality rates, emphasizing the need for further analysis with more precise data or extended time frames.
+@fig-eventstudy depicts the estimated effects of abortion bans on infant
+mortality rates over time, with relative months on the x-axis and
+coefficient estimates with 95% confidence intervals on the y-axis.
+Before the implementation of abortion bans ($t<0$), the coefficients are
+centered around zero, indicating no significant differences in infant
+mortality rates between treated and control states, supporting the
+parallel trends assumption. Post-ban ($t\geq0$), the estimates show a
+slight upward trend, suggesting a potential increase in infant mortality
+rates in restricted states; however, the wide confidence intervals,
+which overlap with zero, indicate that these effects are not
+statistically significant. Overall, the results do not provide
+conclusive evidence of a causal relationship between abortion bans and
+changes in infant mortality rates, emphasizing the need for further
+analysis with more precise data or extended time frames.
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model.
-It represents posterior predictive checks for first model and second model, illustrating how well each model captures the observed data.
-The posterior predictive checks for first model (policy-focused) and second model (policy and demographic factors) illustrate model fit.
-In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision.
-@fig-ppcheckandposteriorvsprior-2 shows improved alignment, indicating that adding demographic factors enhances predictive accuracy and better captures the complexities of infant mortality rates.
-
-In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior.
-The posterior vs.
-prior plots for first model and second model illustrate the role of data in informing the posterior estimates of the model parameters.
-In the first model @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power.
-second model @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables.
-The comparison indicates that the second model provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.
+In @fig-ppcheck-1 and @fig-ppcheck-2 we implement a posterior predictive
+check for the first model and second model. It represents posterior
+predictive checks for first model and second model, illustrating how
+well each model captures the observed data. The posterior predictive
+checks for first model (policy-focused) and second model (policy and
+demographic factors) illustrate model fit. In @fig-ppcheck-1, the
+observed ($y$) and replicated ($y_{rep}$) distributions align
+moderately, reflecting general trends but limited precision.
+@fig-ppcheck-2 shows improved alignment, indicating that adding
+demographic factors enhances predictive accuracy and better captures the
+complexities of infant mortality rates.
+
+In @fig-posteriorvsprior1 and @fig-posteriorvsprior2 we compare the
+posterior with the prior. The posterior vs. prior plots for first model
+and second model illustrate the role of data in informing the posterior
+estimates of the model parameters. In the first model
+@fig-posteriorvsprior1, which focuses on policy-related predictors
+(e.g., abortion restrictions), the data significantly shift the
+posterior distributions for key variables, demonstrating their
+explanatory power. second model @fig-posteriorvsprior2, incorporating
+additional demographic factors such as maternal age and race, shows a
+more refined and concentrated posterior distribution, emphasizing the
+added predictive utility of demographic variables. The comparison
+indicates that the second model provides better-informed parameter
+estimates, supporting the hypothesis that demographic factors play an
+essential role in understanding infant mortality in the context of
+restrictive abortion laws.
 
@@
 #| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model for infant death rate fits, and is affected by, the data"
-#| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check", "First Model Comparing the posterior with the prior", "Second Model Comparing the posterior with the prior"]
+#| label: fig-ppcheck
+#| layout-ncol: 1
+#| fig-cap: "Posterior Predictive Check: Examining how the model for infant death rate fits, and is affected by, the data"
+#| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check" ]
 
@@
   theme(legend.position = "bottom")
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-posteriorvsprior1
+#| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the first model for infant death rate fits, and is affected by, the data"
 posterior_vs_prior(first_model) +
@@
   coord_flip()
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-posteriorvsprior2
+#| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the second model for infant death rate fits, and is affected by, the data"
+#| 
 posterior_vs_prior(second_model) +
@@
 ```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
+#| eval: false
+#| echo: false
+#| message: false
+#| warning: false
+#| include: false
+
 prior_summary(first_model)
@@
 ```
-The prior summaries for the two models, `first_model` and `second_model`, indicate that both use weakly informative priors.
-For the intercept, a normal prior with a mean of 0 and a standard deviation of 2.5 was specified, which adjusts to a scale of 12 after centering the predictors.
-The coefficients for both models are also assigned normal priors with a mean of 0 and a scale of 2.5, which are adjusted based on the predictors, resulting in scales such as [26.84, 28.43, 45.17] for the first model and [26.84, 28.43, 24.78,...] for the second model.
-The auxiliary parameter (sigma) follows an exponential prior with a rate of 1, adjusted to 0.2, while a decov prior regularizes the covariance structure in both models.
-Although the prior structure is similar, the second model incorporates more predictors, leading to a larger number of coefficients with slightly adjusted scales.
-These priors are designed to be weakly informative, allowing the data to primarily guide the posterior estimates while avoiding extreme parameter values.
+
+## Prior Summary for Models
+
+| **Model**        | **Component**     | **Specified Prior**                              | **Adjusted Prior**                               | **Description**                                         |
+|--------------|--------------|--------------|-------------------|------------------|
+| **First Model**  | Intercept         | Normal(0, 2.5)                                   | Normal(0, 12)                                    | Prior for intercept after centering predictors.         |
+|                  | Coefficients      | Normal(\[0,0,0\], \[2.5,2.5,2.5\])               | Normal(\[0,0,0\], \[26.84,28.43,45.17\])         | Prior for coefficients, adjusted for predictor scaling. |
+|                  | Auxiliary (sigma) | Exponential(1)                                   | Exponential(0.2)                                 | Prior for the standard deviation of residuals.          |
+|                  | Covariance        | Decov(reg. = 1, conc. = 1, shape = 1, scale = 1) | Decov(reg. = 1, conc. = 1, shape = 1, scale = 1) | Regularizing prior for covariance.                      |
+| **Second Model** | Intercept         | Normal(0, 2.5)                                   | Normal(0, 12)                                    | Prior for intercept after centering predictors.         |
+|                  | Coefficients      | Normal(\[0,0,0,...\], \[2.5,2.5,2.5,...\])       | Normal(\[0,0,0,...\], \[26.84,28.43,24.78,...\]) | Prior for coefficients, adjusted for predictor scaling. |
+|                  | Auxiliary (sigma) | Exponential(1)                                   | Exponential(0.2)                                 | Prior for the standard deviation of residuals.          |
+|                  | Covariance        | Decov(reg. = 1, conc. = 1, shape = 1, scale = 1) | Decov(reg. = 1, conc. = 1, shape = 1, scale = 1) | Regularizing prior for covariance.                      |
+
+: Prior Specifications for Bayesian Models {#tbl-priorsummary}
+
+@tbl-priorsummary shows the prior summaries for the two models,
+`first_model` and `second_model`, indicate that both use weakly
+informative priors. For the intercept, a normal prior with a mean of 0
+and a standard deviation of 2.5 was specified, which adjusts to a scale
+of 12 after centering the predictors. The coefficients for both models
+are also assigned normal priors with a mean of 0 and a scale of 2.5,
+which are adjusted based on the predictors, resulting in scales such as
+\[26.84, 28.43, 45.17\] for the first model and \[26.84, 28.43,
+24.78,...\] for the second model. The auxiliary parameter (sigma)
+follows an exponential prior with a rate of 1, adjusted to 0.2, while a
+decov prior regularizes the covariance structure in both models.
+Although the prior structure is similar, the second model incorporates
+more predictors, leading to a larger number of coefficients with
+slightly adjusted scales. These priors are designed to be weakly
+informative, allowing the data to primarily guide the posterior
+estimates while avoiding extreme parameter values.
 
 ## Distribution
+
 ```{r}
@@
 #| label: fig-area
-#| layout-ncol: 2
+#| layout-ncol: 1
 #| fig-cap: "95% credibility intervals for fitted models"
@@
 )
+
 plot(
@@
 ```
-Credibility intervals are the Bayesian equivalent of confidence intervals, offering an understanding of the range of probable values for a parameter @tellingstories.
-Using Bayesian estimation, we derive a posterior distribution for each model coefficient.
-@fig-area-1 presents the 95% credibility intervals for coefficients in the first model, while @fig-area-2 illustrates the same for the second model.
-These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.
+
+Credibility intervals are the Bayesian equivalent of confidence
+intervals, offering an understanding of the range of probable values for
+a parameter @tellingstories. Using Bayesian estimation, we derive a
+posterior distribution for each model coefficient. @fig-area-1 presents
+the 95% credibility intervals for coefficients in the first model, while
+@fig-area-2 illustrates the same for the second model. These intervals
+represent the range within which 95% of the posterior probability mass
+falls, offering a robust understanding of parameter uncertainty and the
+influence of predictors under the Bayesian framework @tellingstories.
 
@@
 ```
-The RMSE values from cross-validation indicate that the first model and second model perform similarly in predicting infant mortality rates, with RMSE values of 5.07 and 5.06, respectively.
-This similarity suggests that adding demographic factors such as maternal age and race in the second model has only a marginal impact on predictive accuracy compared to the first model, which primarily relies on policy-related variables like abortion legality and timing of the injunction.
-An average error of approximately 5 deaths per 1,000 live births indicates the magnitude of prediction errors, which may or may not be acceptable depending on the variability in the dataset.
-While the second model slightly reduces the error, its added complexity may not justify the trade-off if simplicity is a priority.
-However, if a nuanced understanding of demographic influences is essential, the second model provides a modest improvement in accuracy while offering more detailed insights into maternal and infant health disparities.
-The cross-validation process reinforces the reliability of these models by confirming their performance across multiple data splits, reducing the risk of overfitting.
+
+The RMSE values from cross-validation indicate that the first model and
+second model perform similarly in predicting infant mortality rates,
+with RMSE values of 5.07 and 5.06, respectively. This similarity
+suggests that adding demographic factors such as maternal age and race
+in the second model has only a marginal impact on predictive accuracy
+compared to the first model, which primarily relies on policy-related
+variables like abortion legality and timing of the injunction. An
+average error of approximately 5 deaths per 1,000 live births indicates
+the magnitude of prediction errors, which may or may not be acceptable
+depending on the variability in the dataset. While the second model
+slightly reduces the error, its added complexity may not justify the
+trade-off if simplicity is a priority. However, if a nuanced
+understanding of demographic influences is essential, the second model
+provides a modest improvement in accuracy while offering more detailed
+insights into maternal and infant health disparities. The
+cross-validation process reinforces the reliability of these models by
+confirming their performance across multiple data splits, reducing the
+risk of overfitting.
 
@@
 #| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm for the infant death rate model"
-#| fig-subcap: ["First Model Trace plot", "First Model Rhat plot","Second Model Trace plot", "Second Model Rhat plot"]
-#| layout-ncol: 2
+#| label: fig-stanareyouokay1
+#| fig-cap: "First Model: Checking the convergence of the MCMC algorithm for the infant death rate model"
+#| fig-subcap: ["First Model Trace plot", "First Model Rhat plot"]
+#| layout-ncol: 1
 
@@
 plot(first_model, "rhat")
-
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-stanareyouokay2
+#| fig-cap: "Second Model: Checking the convergence of the MCMC algorithm for the infant death rate model"
+#| fig-subcap: ["Second Model Trace plot", "Second Model Rhat plot"]
+#| layout-ncol: 1
 plot(second_model, "trace")
@@
 
-The trace plot in @fig-stanareyouokay-1 shows the sampling chains for the first model, indicating good mixing and convergence as the chains overlap and appear stable throughout.
-This suggests that the model has achieved convergence during sampling.
-The Rhat plot in @fig-stanareyouokay-2 demonstrates that all Rhat values are below 1.05, further confirming convergence and indicating that the model’s posterior samples are reliable.
-@fig-stanareyouokay-3 represents the trace plots for the second model, showing the posterior samples across four chains.
-The plots indicate good mixing, as the chains overlap well and stabilize around the same region, suggesting convergence.
-@fig-stanareyouokay-4 is the Rhat diagnostic plot for the second model.
-All Rhat values are close to 1.00, confirming that the Markov chains have converged, and the posterior estimates are reliable.
-This supports the validity of the second model’s Bayesian estimation process.
+The trace plot in @fig-stanareyouokay1-1 shows the sampling chains for
+the first model, indicating good mixing and convergence as the chains
+overlap and appear stable throughout. This suggests that the model has
+achieved convergence during sampling. The Rhat plot in
+@fig-stanareyouokay1-2 demonstrates that all Rhat values are below 1.05,
+further confirming convergence and indicating that the model’s posterior
+samples are reliable. @fig-stanareyouokay2-1 represents the trace plots
+for the second model, showing the posterior samples across four chains.
+The plots indicate good mixing, as the chains overlap well and stabilize
+around the same region, suggesting convergence. @fig-stanareyouokay2-2 is
+the Rhat diagnostic plot for the second model. All Rhat values are close
+to 1.00, confirming that the Markov chains have converged, and the
+posterior estimates are reliable. This supports the validity of the
+second model’s Bayesian estimation process.
 
@@
 
-Our study examines the impact of abortion bans on infant mortality rates in the U.S.
-following the Dobbs decision, combining observational Difference-in-Differences (DID) analysis with qualitative surveys.
-While the DID analysis quantifies policy impacts, qualitative surveys provide a context-specific understanding of how restrictive abortion policies affect maternal mental health, healthcare access, and pregnancy decisions.
-These lived experiences help interpret quantitative findings, offering a fuller understanding of the systemic and psychological factors driving observed trends, and supporting evidence-based policy recommendations.
+Our study examines the impact of abortion bans on infant mortality rates
+in the U.S. following the Dobbs decision, combining observational
+Difference-in-Differences (DID) analysis with qualitative surveys. While
+the DID analysis quantifies policy impacts, qualitative surveys provide
+a context-specific understanding of how restrictive abortion policies
+affect maternal mental health, healthcare access, and pregnancy
+decisions. These lived experiences help interpret quantitative findings,
+offering a fuller understanding of the systemic and psychological
+factors driving observed trends, and supporting evidence-based policy
+recommendations.
 
@@
 
-Restrictive abortion laws can heighten maternal stress, anxiety, or depression—particularly in unintended or high-risk pregnancies—compromising maternal health and prenatal care.
-Research @crugnola2022 highlights maternal mental health as a key determinant of infant health outcomes, with conditions like chronic stress and depression adversely affecting prenatal behaviors, healthcare access, and overall well-being.
-These compounded challenges, driven by societal and systemic barriers, can significantly increase risks of adverse infant outcomes, including mortality.
-
-Our survey focuses on mothers unable to access abortion services who later experienced the death of their infant, documenting their experiences to better understand these interconnected factors.
-By integrating quantitative data with qualitative perspectives, this study aims to inform evidence-based interventions, addressing maternal psychological and healthcare needs, and mitigating the adverse effects of restrictive abortion policies.
+Restrictive abortion laws can heighten maternal stress, anxiety, or
+depression—particularly in unintended or high-risk
+pregnancies—compromising maternal health and prenatal care. Research
+@crugnola2022 highlights maternal mental health as a key determinant of
+infant health outcomes, with conditions like chronic stress and
+depression adversely affecting prenatal behaviors, healthcare access,
+and overall well-being. These compounded challenges, driven by societal
+and systemic barriers, can significantly increase risks of adverse
+infant outcomes, including mortality.
+
+Our survey focuses on mothers unable to access abortion services who
+later experienced the death of their infant, documenting their
+experiences to better understand these interconnected factors. By
+integrating quantitative data with qualitative perspectives, this study
+aims to inform evidence-based interventions, addressing maternal
+psychological and healthcare needs, and mitigating the adverse effects
+of restrictive abortion policies.
 
@@
 The objective of this study is to investigate the psychological, social,
-and systemic impacts of restrictive abortion laws on maternal health and infant outcomes.
-By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences and explore the role of maternal stress and healthcare barriers in adverse outcomes.
-Understanding how maternal mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the systemic factors contributing to higher infant mortality rates.
-The findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access, and shape policy responses to restrictive abortion laws.
+and systemic impacts of restrictive abortion laws on maternal health and
+infant outcomes. By focusing on mothers unable to access abortion
+services who experienced the death of their infant, the study aims to
+document lived experiences and explore the role of maternal stress and
+healthcare barriers in adverse outcomes. Understanding how maternal
+mental health directly affects prenatal care decisions, access to
+healthcare, and overall maternal behavior is essential to addressing the
+systemic factors contributing to higher infant mortality rates. The
+findings will guide the development of evidence-based strategies to
+improve maternal mental health, enhance healthcare access, and shape
+policy responses to restrictive abortion laws.
 
@@
 
-In this analysis, we use non-probability Respondent-Driven Sampling (RDS), developed by @heckathorn1997, is a recruitment method designed for hidden populations lacking a sampling frame, where participants recruit others through their social networks, with dual compensation for participation and recruitment, ensuring confidentiality and mitigating biases.
-RDS enables peer-to-peer recruitment with dual compensation for participation and recruitment, fostering trust within communities.
-This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics like abortion.
-
-We use Respondent-Driven Sampling (RDS) for our survey as it is effective for accessing hidden or stigmatized populations, such as mothers unable to access abortion services and who experienced the death of their infant.
-RDS utilizes peer networks for recruitment, fostering trust and participation in sensitive contexts while maintaining confidentiality and reducing social stigma.
-Its statistical adjustments mitigate biases in network-based recruitment, enhancing sample representativeness.
-These strengths make RDS an ethical and effective method for examining the psychological, social, and systemic impacts of restrictive abortion laws.
+In this analysis, we use non-probability Respondent-Driven Sampling
+(RDS), developed by @heckathorn1997, is a recruitment method designed
+for hidden populations lacking a sampling frame, where participants
+recruit others through their social networks, with dual compensation for
+participation and recruitment, ensuring confidentiality and mitigating
+biases. RDS enables peer-to-peer recruitment with dual compensation for
+participation and recruitment, fostering trust within communities. This
+method maintains confidentiality, reduces bias through statistical
+adjustments, and is particularly effective for sensitive topics like
+abortion.
+
+We use Respondent-Driven Sampling (RDS) for our survey as it is
+effective for accessing hidden or stigmatized populations, such as
+mothers unable to access abortion services and who experienced the death
+of their infant. RDS utilizes peer networks for recruitment, fostering
+trust and participation in sensitive contexts while maintaining
+confidentiality and reducing social stigma. Its statistical adjustments
+mitigate biases in network-based recruitment, enhancing sample
+representativeness. These strengths make RDS an ethical and effective
+method for examining the psychological, social, and systemic impacts of
+restrictive abortion laws.
 
@@
 
-Given the sensitive nature of this study, identifying and reaching participants presents a significant challenge.
-To address this, we will collaborate with support groups and advocacy organizations, such as Abortion Rights Advocacy Groups, which can connect researchers with individuals otherwise difficult to reach.
-These organizations are essential for engaging with populations affected by restrictive abortion laws, particularly concerning maternal and infant health.
-
-Thus, the sample frame for our study comprises mothers who reside in areas where abortion services are inaccessible due to restrictive laws, have experienced the death of their infant within the first year of life, and are reachable through advocacy organizations following the Dobbs decision.
+Given the sensitive nature of this study, identifying and reaching
+participants presents a significant challenge. To address this, we will
+collaborate with support groups and advocacy organizations, such as
+Abortion Rights Advocacy Groups, which can connect researchers with
+individuals otherwise difficult to reach. These organizations are
+essential for engaging with populations affected by restrictive abortion
+laws, particularly concerning maternal and infant health.
+
+Thus, the sample frame for our study comprises mothers who reside in
+areas where abortion services are inaccessible due to restrictive laws,
+have experienced the death of their infant within the first year of
+life, and are reachable through advocacy organizations following the
+Dobbs decision.
 
@@
 
-We aim to survey 20 respondents who meet our defined sample criteria: mothers unable to access abortion services due to restrictive laws, who experienced the death of their infant within the first year of life, and who are reachable through Abortion Rights Advocacy Groups.
-Participation will be voluntary, and respondents must be willing to answer the survey questions, ensuring the quality and depth of the data collected.
+We aim to survey 20 respondents who meet our defined sample criteria:
+mothers unable to access abortion services due to restrictive laws, who
+experienced the death of their infant within the first year of life, and
+who are reachable through Abortion Rights Advocacy Groups. Participation
+will be voluntary, and respondents must be willing to answer the survey
+questions, ensuring the quality and depth of the data collected.
 
@@
 
-To recruit participants for this sensitive study, we will employ a Respondent-Driven Sampling (RDS) approach in collaboration with Abortion Rights Advocacy Groups, leveraging their established connections with our target population.
-This organization will act as initial "seeds" to identify mothers affected by restrictive abortion laws who meet our study criteria.
-Participants may also invite others from their network with similar experiences, allowing the sample to grow while maintaining confidentiality.
-Outreach materials emphasizing anonymity and the study’s significance will be shared through trusted channels.
-Interested participants will undergo a screening process and provide informed consent, with the option to complete the survey online for safety and privacy.
-To ensure participant well-being, mental health resources will be offered throughout the process, fostering trust and encouraging meaningful engagement.
+To recruit participants for this sensitive study, we will employ a
+Respondent-Driven Sampling (RDS) approach in collaboration with Abortion
+Rights Advocacy Groups, leveraging their established connections with
+our target population. This organization will act as initial "seeds" to
+identify mothers affected by restrictive abortion laws who meet our
+study criteria. Participants may also invite others from their network
+with similar experiences, allowing the sample to grow while maintaining
+confidentiality. Outreach materials emphasizing anonymity and the
+study’s significance will be shared through trusted channels. Interested
+participants will undergo a screening process and provide informed
+consent, with the option to complete the survey online for safety and
+privacy. To ensure participant well-being, mental health resources will
+be offered throughout the process, fostering trust and encouraging
+meaningful engagement.
 
@@
 
-Non-response bias is a concern as it can distort survey results and lead to unrepresentative conclusions about the target population.
-Since our survey addresses a sensitive topic and takes approximately 10–15 minutes to complete, there is a risk of participant dropout or refusal.
-To minimize this, we provide a clear overview of the survey’s purpose, confidentiality measures, and estimated completion time, fostering trust and encouraging participation.
+Non-response bias is a concern as it can distort survey results and lead
+to unrepresentative conclusions about the target population. Since our
+survey addresses a sensitive topic and takes approximately 10–15 minutes
+to complete, there is a risk of participant dropout or refusal. To
+minimize this, we provide a clear overview of the survey’s purpose,
+confidentiality measures, and estimated completion time, fostering trust
+and encouraging participation.
 
@@
 
-To ensure data credibility, we implement a robust respondent validation process.
-Participants will first undergo screening to confirm eligibility, including inability to access abortion services due to restrictive laws and experiencing infant loss within the first year of life.
-Responses will be assessed for completeness and consistency, and duplicate entries will be removed to maintain integrity.
-Partnering with trusted support groups and advocacy organizations enhances reliability, ensuring the findings accurately represent the target population’s experiences.
+To ensure data credibility, we implement a robust respondent validation
+process. Participants will first undergo screening to confirm
+eligibility, including inability to access abortion services due to
+restrictive laws and experiencing infant loss within the first year of
+life. Responses will be assessed for completeness and consistency, and
+duplicate entries will be removed to maintain integrity. Partnering with
+trusted support groups and advocacy organizations enhances reliability,
+ensuring the findings accurately represent the target population’s
+experiences.
 
@@
 
-This study addresses sensitive and emotionally charged topics, requiring an ethical approach to safeguard participants’ well-being.
-Recognizing the potential psychological distress associated with discussing unintended pregnancies, restrictive abortion laws, and infant loss, the survey will include trigger warnings and allow participants to skip questions or withdraw at any time without penalty.
-Strict confidentiality protocols will protect participants’ identities, and responses will be securely stored.
-Recruitment through trusted advocacy groups will foster a safe and supportive environment, while mental health resources will be provided to those who experience discomfort during the study.
-This framework reflects our commitment to ethical, respectful, and responsible research practices.
+This study addresses sensitive and emotionally charged topics, requiring
+an ethical approach to safeguard participants’ well-being. Recognizing
+the potential psychological distress associated with discussing
+unintended pregnancies, restrictive abortion laws, and infant loss, the
+survey will include trigger warnings and allow participants to skip
+questions or withdraw at any time without penalty. Strict
+confidentiality protocols will protect participants’ identities, and
+responses will be securely stored. Recruitment through trusted advocacy
+groups will foster a safe and supportive environment, while mental
+health resources will be provided to those who experience discomfort
+during the study. This framework reflects our commitment to ethical,
+respectful, and responsible research practices.
 
@@
 experimental approaches significantly improved abortion reporting
-compared to the control condition, their study offers meaningful guidelines
-for improving abortion measurement. We aim to adapt our survey design
-based on their methodologies to enhance data accuracy and respondent
-comfort.
-
-This survey explores the impacts of restrictive abortion laws on maternal mental health and infant outcomes, emphasizing neutral, carefully crafted questions to ensure participant comfort and minimize social desirability bias.
-Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., "Prefer not to say") to address underreporting in sensitive topics.
-Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with detailed understanding, fostering trust and enhancing data accuracy.
-Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a "helping" framework in each section of our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.
+compared to the control condition, their study offers meaningful
+guidelines for improving abortion measurement. We aim to adapt our
+survey design based on their methodologies to enhance data accuracy and
+respondent comfort.
+
+This survey explores the impacts of restrictive abortion laws on
+maternal mental health and infant outcomes, emphasizing neutral,
+carefully crafted questions to ensure participant comfort and minimize
+social desirability bias. Inspired by @lindberg2022, it employs
+strategies like thoughtful question placement, phrasing, and opt-out
+options (e.g., "Prefer not to say") to address underreporting in
+sensitive topics. Combining multiple-choice and open-ended formats, the
+survey aims to balance structured data collection with detailed
+understanding, fostering trust and enhancing data accuracy. Also, based
+on @lindberg2022 findings that respondents preferred introductions
+emphasizing how sharing their abortion experiences could help improve
+health services for other women, we adopted a "helping" framework in
+each section of our survey introductions to encourage accurate reporting
+and increase participants’ motivation to contribute.
 
@@
 
-We draw on recommendations from @stantcheva2023 to minimize response biases.
-Common response biases identified in survey design include moderacy bias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB).
-Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB.
-The detailed definitions of these biases are provided in [Appendix-@sec-definition].
-
-To mitigate bias, we enhance our survey in the following ways, drawing on recommendations
-from @stantcheva2023:
-
-- Addressing Extreme/Moderacy Bias: We use a minimum of five response options for scale questions to provide more detailed choices, reducing the likelihood of respondents defaulting to extreme or middle answers.
-
-- Mitigating Response Order Bias: For nominal questions, we randomize response options.
-
-- Minimizing Social Desirability Bias (SDB): The survey design addresses social desirability bias (SDB) by emphasizing anonymity and confidentiality throughout.
-A clear introduction outlines the survey’s purpose—academic research on the impact of restrictive abortion laws—and reassures participants that their responses will remain confidential and solely used for research.
-The anonymous online format creates a safe environment for participants to share their experiences without fear of judgment or stigma.
-A feedback section at the end encourages participants to express concerns or share additional thoughts, fostering trust and enhancing data quality.
+We draw on recommendations from @stantcheva2023 to minimize response
+biases. Common response biases identified in survey design include
+moderacy bias, extreme response bias, ordering bias, acquiescence bias,
+experimenter demand effect(EDE), and social desirability bias (SDB). Our
+survey primarily focuses on strategies to reduce moderacy bias, extreme
+response bias, ordering bias, and SDB. The detailed definitions of these
+biases are provided in [Appendix- @sec-definition].
+
+To mitigate bias, we enhance our survey in the following ways, drawing
+on recommendations from @stantcheva2023:
+
+-   Addressing Extreme/Moderacy Bias: We use a minimum of five response
+    options for scale questions to provide more detailed choices,
+    reducing the likelihood of respondents defaulting to extreme or
+    middle answers.
+
+-   Mitigating Response Order Bias: For nominal questions, we randomize
+    response options.
+
+-   Minimizing Social Desirability Bias (SDB): The survey design
+    addresses social desirability bias (SDB) by emphasizing anonymity
+    and confidentiality throughout. A clear introduction outlines the
+    survey’s purpose—academic research on the impact of restrictive
+    abortion laws—and reassures participants that their responses will
+    remain confidential and solely used for research. The anonymous
+    online format creates a safe environment for participants to share
+    their experiences without fear of judgment or stigma. A feedback
+    section at the end encourages participants to express concerns or
+    share additional thoughts, fostering trust and enhancing data
+    quality.
 
@@
     -   Prefer not to say
-
-3. What is your marital status? - Single - Married - Divorced - Widowed -
-Prefer not to say
+3.  What is your marital status? 
+    - Single 
+    - Married 
+    - Divorced 
+    - Widowed 
+    - Prefer not to say
 
@@
 
-1.  Do you have any concerns or feedback regarding the survey, surveyor,
-    or entity?
+1.  Do you have any concerns or feedback regarding the survey or entity?
     -   Your feedback is important to us and will help ensure
@@
 
-In the design of the questionnaire, there will be some common biases that
-may occur when running the questionnaire.
+In the design of the questionnaire, there will be some common biases
+that may occur when running the questionnaire.
 
@@
 
-- **Moderacy response bias**: The tendency to choose mid-scale responses.
-
-- **Extreme response bias**: The inclination to select extreme values on the scale.
-
-- **Response order bias**: Occurs when the position of options influences responses, with the **primacy effect** (favoring early options in written surveys) and **recency effect** (favoring later options in oral surveys) as common forms.
-
-- **Social desirability bias**: Respondents may hide true views to present a socially favorable image, influenced by topic and social context.
-
-- **Acquiescence bias**: The habit of giving positive responses, such as consistently choosing "agree" or "yes."
-\newpage
+-   **Moderacy response bias**: The tendency to choose mid-scale
+    responses.
+
+-   **Extreme response bias**: The inclination to select extreme values
+    on the scale.
+
+-   **Response order bias**: Occurs when the position of options
+    influences responses, with the **primacy effect** (favoring early
+    options in written surveys) and **recency effect** (favoring later
+    options in oral surveys) as common forms.
+
+-   **Social desirability bias**: Respondents may hide true views to
+    present a socially favorable image, influenced by topic and social
+    context.
+
+-   **Acquiescence bias**: The habit of giving positive responses, such
+    as consistently choosing "agree" or "yes." \newpage
 

### 2024-11-29 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
fix grammar error

@@
 These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities.
-This study uses latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
+This study uses the latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
 format: pdf
@@
 [@modelsummary] to generate clean and interpretable summaries of our
-regression models.We used `knitr` [@knitr] and `kableExtra`
+regression models. We used `knitr` [@knitr] and `kableExtra`
 [@kableExtra] to generate clean tables and reproducible reports, and
@@
 we analyze infant mortality rates in the United States to understand the
-impact of abortion policies and demographic factors. The CDC’s Linked
-Birth/Infant Death Records database provides detailed data, including
+impact of abortion policies and demographic factors. The CDC’s Linked Birth/Infant Death Records database provides detailed data, including
 maternal and infant characteristics, to support this study. Variables
@@
 figures do not meet the National Center for Health Statistics (NCHS)
-standards for reliability or precision. After cleaning the dataset, the
-"Unreliable" label is converted into a separate variable named
+standards for reliability or precision. After cleaning the dataset, the "Unreliable" label is converted into a separate variable named
 `reliable`.
@@
 
-@tbl-summary shows that 50.7% of infant deaths occurred in 2022,
-slightly higher than 2021 (49.3%). The majority of cases were among
+@tbl-summary shows that 50.7% of infant deaths occurred in 2022, slightly higher than in 2021 (49.3%). The majority of cases were among
 mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
@@
 spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential
-prior (rate = 1) for variability parameters. These priors automatically
-adjust to the scale of the data, improving model stability and ensuring
-they complement rather than dominate the observed data. This approach
+prior (rate = 1) for variability parameters.
+These priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data.
+This approach
 helps us center estimates realistically, account for hidden differences
@@
 abortion_illegal is positive and statistically significant (0.111 in
-first model and 0.280 in second model), indicating a change in infant
+first model and 0.280 in the second model), indicating a change in infant
 mortality rates associated with the combination of abortion bans and the
@@
 groups 25–29 and 30–34 of -5.73 and -6.15 in the second model, relative
-to groups 15-19 respectively. The the coefficient for Black mothers in
+to groups 15-19 respectively. The coefficient for Black mothers in
 the second model is 6.66 relative to Asia mothers, reflecting a higher
@@
 parallel trends assumption. Post-ban ($t\geq0$), the estimates show a
-slight upward trend, suggesting a potential increase in infant mortality
-rates in restricted states; however, the wide confidence intervals,
-which overlap with zero, indicate that these effects are not
-statistically significant. Overall, the results do not provide
+slight upward trend, suggesting a potential increase in infant mortality rates in restricted states; however, the wide confidence intervals, which overlap with zero, indicate that these effects are not statistically significant.
+Overall, the results do not provide
 conclusive evidence of a causal relationship between abortion bans and
@@
 check for the first model and second model. It represents posterior
-predictive checks for first model and second model, illustrating how
+predictive checks for the first model and second model, illustrating how
 well each model captures the observed data. The posterior predictive
-checks for first model (policy-focused) and second model (policy and
+checks for the first model (policy-focused) and the second model (policy and
 demographic factors) illustrate model fit. In @fig-ppcheck-1, the
@@
 In @fig-posteriorvsprior1 and @fig-posteriorvsprior2 we compare the
-posterior with the prior. The posterior vs. prior plots for first model
-and second model illustrate the role of data in informing the posterior
-estimates of the model parameters. In the first model
-@fig-posteriorvsprior1, which focuses on policy-related predictors
-(e.g., abortion restrictions), the data significantly shift the
+posterior with the prior.
+The posterior vs.
+prior plots for the first model and second model illustrate the role of data in informing the posterior estimates of the model parameters.
+In the first model
+@fig-posteriorvsprior1, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shifts the
 posterior distributions for key variables, demonstrating their
-explanatory power. second model @fig-posteriorvsprior2, incorporating
-additional demographic factors such as maternal age and race, shows a
-more refined and concentrated posterior distribution, emphasizing the
+explanatory power.
+The second model @fig-posteriorvsprior2, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the
 added predictive utility of demographic variables. The comparison
@@
 
-@tbl-priorsummary shows the prior summaries for the two models,
-`first_model` and `second_model`, indicate that both use weakly
+@tbl-priorsummary shows the prior summaries for the two models, `first_model` and `second_model`, indicating that both use weakly
 informative priors. For the intercept, a normal prior with a mean of 0
@@
 
-The RMSE values from cross-validation indicate that the first model and
-second model perform similarly in predicting infant mortality rates,
+The RMSE values from cross-validation indicate that the first model and the second model perform similarly in predicting infant mortality rates,
 with RMSE values of 5.07 and 5.06, respectively. This similarity
@@
 compared to the first model, which primarily relies on policy-related
-variables like abortion legality and timing of the injunction. An
-average error of approximately 5 deaths per 1,000 live births indicates
+variables like abortion legality and timing of the injunction. An average error of approximately 5 deaths per 1,000 live births indicates
 the magnitude of prediction errors, which may or may not be acceptable
@@
 in the U.S. following the Dobbs decision, combining observational
-Difference-in-Differences (DID) analysis with qualitative surveys. While
-the DID analysis quantifies policy impacts, qualitative surveys provide
+Difference-in-Differences (DID) analysis with qualitative surveys. While the DID analysis quantifies policy impacts, qualitative surveys provide
 a context-specific understanding of how restrictive abortion policies
@@
 @crugnola2022 highlights maternal mental health as a key determinant of
-infant health outcomes, with conditions like chronic stress and
-depression adversely affecting prenatal behaviors, healthcare access,
-and overall well-being. These compounded challenges, driven by societal
-and systemic barriers, can significantly increase risks of adverse
+infant health outcomes, with conditions like chronic stress and depression adversely affecting prenatal behaviors, healthcare access,
+and overall well-being. These compounded challenges, driven by societal and systemic barriers, can significantly increase risks of adverse
 infant outcomes, including mortality.
@@
 healthcare barriers in adverse outcomes. Understanding how maternal
-mental health directly affects prenatal care decisions, access to
-healthcare, and overall maternal behavior is essential to addressing the
+mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the
 systemic factors contributing to higher infant mortality rates. The
-findings will guide the development of evidence-based strategies to
-improve maternal mental health, enhance healthcare access, and shape
+findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access, and shape
 policy responses to restrictive abortion laws.
@@
 biases. RDS enables peer-to-peer recruitment with dual compensation for
-participation and recruitment, fostering trust within communities. This
-method maintains confidentiality, reduces bias through statistical
-adjustments, and is particularly effective for sensitive topics like
-abortion.
+participation and recruitment, fostering trust within communities.
+This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics like abortion.
 
@@
 participants presents a significant challenge. To address this, we will
-collaborate with support groups and advocacy organizations, such as
-Abortion Rights Advocacy Groups, which can connect researchers with
-individuals otherwise difficult to reach. These organizations are
+collaborate with support groups and advocacy organizations, such as Abortion Rights Advocacy Groups, which can connect researchers with individuals otherwise difficult to reach.
+These organizations are
 essential for engaging with populations affected by restrictive abortion
@@
 sensitive topics. Combining multiple-choice and open-ended formats, the
-survey aims to balance structured data collection with detailed
-understanding, fostering trust and enhancing data accuracy. Also, based
+survey aims to balance structured data collection with detailed understanding, fostering trust and enhancing data accuracy. Also, based
 on @lindberg2022 findings that respondents preferred introductions
@@
     their experiences without fear of judgment or stigma. A feedback
-    section at the end encourages participants to express concerns or
-    share additional thoughts, fostering trust and enhancing data
+    section at the end encourages participants to express concerns or share additional thoughts, fostering trust and enhancing data
     quality.

### 2024-11-29 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with format error

@@
 
--   **Treatment Group Indicator `abortion_illegal`**: A binary variable
+-   **Treatment Group Indicator (`abortion_illegal`)**: A binary variable
     indicating the group: 1 if abortion is illegal in the state. 0 if
@@
 
--   **Time Indicator `after_injunction`**: A binary variable indicating
+-   **Time Indicator (`after_injunction`)**: A binary variable indicating
     the time period: 0 for the period before the injunction was
@@
 
--   **Interaction Term`after_injunction` × `abortion_illegal`**: The
+-   **Interaction Term (`after_injunction` × `abortion_illegal`)**: The
     interaction term captures the additional change in the outcome
@@
 
--   **`age_of_mother`**: The age of the mother at the time of birth,
+-   **Age of Mother (`age_of_mother`)**: The age of the mother at the time of birth,
     represented as a continuous variable.
 
--   **`mothers_single_race`**: Self-identified race of the mother,
+-   **Mother's Single Race (`mothers_single_race`)**: Self-identified race of the mother,
     represented as a categorical variable.
 
--   **`year_of_death`**: The year in which the infant death occurred,
+-   **Year of Death (`year_of_death`)**: The year in which the infant death occurred,
     represented as a continuous variable.
@@
 
--   **`state`**: Mother's residence state when the infant was born as a
+-   **State (`state`)**: Mother's residence state when the infant was born as a
     random effect in the model.
@@
 
-One model includes control variables— `age_of_mother`,
-`mothers_single_race`, and `year_of_death` to adjust for confounders
+One model includes control variables — Age of Mother,
+Mother's Single Race, and Year of Death to adjust for confounders
 that may influence infant death rates. This adjustment addresses

### 2024-11-29 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update model card in the paper

@@
 
-Sketches depicting both the desired dataset and the graphs generated in
+Sketches depicting both the desired dataset and the tables generated in
 this analysis is available in the GitHub Repository `other/sketches`.
@@
 # Model details {#sec-model-details}
+
+## Model Card
+
+The model card, referenced from @ozoani2022modelcard, is included in the README under the section titled `Model Card for Infant Mortality and Abortion Policy Analysis Model` for detailed documentation.
 

### 2024-12-01 scripts/01-test_simulated_data.R
update test scripts

@@ -6,12 +6,13 @@
 # Contact: diana.shen@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-# - The `tidyverse` package must be installed and loaded
+# - The `tidyverse`,`testthat` package must be installed and loaded
 # - 00-simulate_data.R must have been run
 # Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 #### Workspace setup ####
 library(tidyverse)
+library(testthat)
 
 #### Define the list of states where abortion is illegal ####
 illegal_states <- c("AL", "AR", "ID", "IN", "KY", "LA", "MS", "OK", "SD", "TN", "TX", "WV")
@@ -26,6 +27,26 @@ if (exists("simulated_data")) {
 }
 
 #### Test data ####
+
+## test simulated data has correct structure
+test_that("Simulated data has correct structure", {
+  expect_true(is.numeric(simulated_data$death_rate)) # death_rate should be numeric
+  expect_true(is.factor(simulated_data$state)) # state should be a factor
+  expect_true(all(levels(simulated_data$state) %in% state.abb)) # state levels should match US state abbreviations
+  expect_true(is.character(simulated_data$age_of_mother) || is.factor(simulated_data$age_of_mother)) # age_of_mother can be character or factor
+  expect_true(all(simulated_data$age_of_mother %in% c("15–19", "20–24", "25–29", "30–34", "35–39", "40–44", "45+"))) # age_of_mother must match predefined categories
+  expect_true(is.character(simulated_data$mothers_single_race) || is.factor(simulated_data$mothers_single_race)) # mothers_single_race can be character or factor
+  expect_true(all(simulated_data$mothers_single_race %in% c("White", "Black", "Asian", "Native American", "Pacific Islander", 
+                                                            "More than one race", "Unknown or Not Stated"))) # mothers_single_race must match predefined categories
+  expect_true(is.numeric(simulated_data$year_of_death)) # year_of_death should be numeric
+  expect_true(all(simulated_data$year_of_death %in% c(2021, 2022))) # year_of_death should be 2021 or 2022
+  expect_true(is.numeric(simulated_data$month)) # month should be numeric
+  expect_true(all(simulated_data$month >= 1 & simulated_data$month <= 12)) # month must be between 1 and 12
[…]

### 2024-12-01 scripts/03-test_analysis_data.R

@@ -6,12 +6,13 @@
 # Contact: diana.shen@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-# - The `tidyverse` package must be installed and loaded
+# - The `tidyverse`,`testthat` package must be installed and loaded
 # - 03-clean_data.R must have been run
 # Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 #### Workspace setup ####
 library(tidyverse)
+library(testthat)
 
 analysis_data <- read_csv(
   file = "data/02-analysis_data/analysis_data.csv",
@@ -43,6 +44,32 @@ state =c(
   "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"
 )
 
+## test analysis data has correct structure
+test_that("Cleaned analysis data has correct structure and values", {
+  
+  # Check that the dataset is a data frame
+  expect_true(is.data.frame(cleaned_data))
+  
+  # Check that all required columns exist
+  required_columns <- c("state", "age_of_mother", "mothers_single_race", "year_of_death", 
+                        "month", "month_code", "death_rate", "after_injunction", 
+                        "abortion_illegal", "reliable", "date")
+  expect_true(all(required_columns %in% colnames(cleaned_data)))
+  
+  # Check column types
+  expect_true(is.character(cleaned_data$state)) # State should be character
+  expect_true(is.factor(cleaned_data$age_of_mother)) # Age of mother should be a factor
+  expect_true(is.character(cleaned_data$mothers_single_race)) # Mother's race should be character
+  expect_true(is.numeric(cleaned_data$year_of_death)) # Year of death should be numeric
+  expect_true(is.numeric(cleaned_data$month_code)) # Month code should be numeric
+  expect_true(is.factor(cleaned_data$month)) # Month should be a factor
+  expect_true(is.numeric(cleaned_data$death_rate)) # Death rate should be numeric
+  expect_true(is.factor(cleaned_data$after_injunction)) # After injunction should be a factor
+  expect_true(is.factor(cleaned_data$abortion_illegal)) # Abortion illegal should be a factor
+  expect_true(is.character(cleaned_data$reliable)) # Reliable column should be character
+  expect_true(inherits(cleaned_data$date, "Date")) # Date should be a valid Date class
+})
+  
 # Test if the data was successfully loaded
 if (exists("analysis_data")) {
   message("Test Passed: The dataset was successfully loaded.")

### 2024-12-01 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with check

@@
 A Difference-in-Differences (DID) analysis shows a 0.285 increase in infant deaths per 1,000 live births in states enforcing bans, with amplified effects post-injunction.
-Maternal age and race are key factors, with Black mothers experiencing higher mortality rates.
+Maternal age and race are key factors, with Black mothers experiencing higher infant mortality rates.
 These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities.
@@
 outcomes. The study utilizes observational data from the Centers for
-Disease Control and Prevention’s WONDER database, captures infant
+Disease Control and Prevention’s WONDER database [@cdc_wonder_2024], captures infant
 mortality trends across states with varying levels of abortion
@@
 programming language [@citeR]. Our dataset, obtained from the CDC
-Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a
-detailed overview of the Infant Death Rate across the USA. CDC WONDER
+Wonder[@cdc_wonder_2024], based on data as of December 2022, provides detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions.
+CDC WONDER
 (Wide-ranging ONline Data for Epidemiologic Research) is an online
@@
 to a wide range of the Centers for Disease Control and Prevention (CDC)
-epidemiologic data and resources. Following the guidelines outlined in
+epidemiologic data and resources.
+This dataset offers detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions.
+Following the guidelines outlined in
 @tellingstories, we analyze the impact of abortion injunctions on infant
-death rates using the Difference-in-Difference method. This approach
-utilizes the interaction between the legality of abortion in the
-mother’s state of residence and the timing of the infant’s death as the
-predictor variable. Additionally, we include control variables such as
-infant year of death, the mother’s state of residence, race, and age to
-account for potential confounding factors.
+death rates using the Difference-in-Difference method. This approach, shown in @sec-model, utilizes the main effects and the interaction between the legality of abortion in the
+mother’s state of residence and the timing of the infant’s death as the predictor variable.
+Additionally, we include control variables, infant year of death, race, and age to account for potential confounding factors, and the mother’s state of residence as a random variable.
+
+We selected the CDC WONDER dataset for its combination of granularity, consistency, and accessibility, making it suitable for a nationwide analysis.
+Alternative datasets lacked either the breadth or the standardization required for a study of this scale.
+Additionally, CDC WONDER’s public availability enhances transparency and allows for independent verification, ensuring the reliability and reproducibility of our research.
 
@@
 
-The `death rate` refers to the number of deaths of infants under one
+The death rate (`death rate`) refers to the number of deaths of infants under one
 year of age per 1,000 live births in a specified population during a
@@
 figures do not meet the National Center for Health Statistics (NCHS)
-standards for reliability or precision. After cleaning the dataset, the "Unreliable" label is converted into a separate variable named
-`reliable`.
+standards for reliability or precision.
+After cleaning the dataset, the "Unreliable" label is converted into a separate variable named `reliable` (This variable is available in the `analysis data` for reference but was not included in the model analysis.).
 
@@
 
-### Interaction Term: Abortion Legality and After Injunction
-
-Abortion legality (named `abortion_illegal` in the dataset) is a newly
+### Abortion Legality
+
+Abortion legality (`abortion_illegal`) is a newly
 created variable in our dataset, which refers to the legal status of
@@
 #| echo: false
+#| fig-pos: "H"
 ggplot() +
@@
     labels = c("Legal", "Illegal"),
-    name = "Abortion Legal Or Not"
+    name = "State is Illegal for Abortion or Not"
   )+
@@
 
+### After Injunction
+
+Another created variable After Injunction (`after_injunction`) indicates whether the date
+of infants' death falls after the implementation of the abortion
+injunctions following the overturn of Roe v. Wade in June 2022. It is
+coded as a binary variable, where 1 represents dates after June 2022,
+signifying the post-injunction period, and 0 represents dates before
+June 2022, indicating the pre-injunction period.
+
+@fig-rateandstate illustrates trends in infant death rates (per 1,000
+live births) over time in states where abortion is illegal versus legal.
+In states with abortion bans, infant mortality shows a slight increase
+after June 2022, marked by the dashed line representing the overturn of
+Roe v. Wade. In contrast, states where abortion remains legal exhibit
+relatively stable trends. This suggests a potential link between
+abortion bans and rising infant mortality rates, warranting further
+investigation.
+
 ```{r}
@@
 #| message: false
+#| fig-pos: "H"
 analysis_data|>
@@
   geom_smooth() +
-  labs(x = "Date of Infant Death", y = "Infant Death Rate \n(per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S.
+  labs(x = "Date of Infant Death", y = "Infant Death Rate \n(per 1,000 live births)",color="State is Illegal for Abortion or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S.
 Supreme Court overturned Roe v.
@@
 
-@fig-rateandstate illustrates trends in infant death rates (per 1,000
-live births) over time in states where abortion is illegal versus legal.
-In states with abortion bans, infant mortality shows a slight increase
-after June 2022, marked by the dashed line representing the overturn of
-Roe v. Wade. In contrast, states where abortion remains legal exhibit
-relatively stable trends. This suggests a potential link between
-abortion bans and rising infant mortality rates, warranting further
-investigation.
+### Interaction Term: Abortion Legality and After Injunction
+
+To evaluate the impact of abortion injunctions on infant death rates, we
+utilize the Difference-in-Difference (DID) method. As part of this
+approach, we construct an interaction term defined as the product of
+`abortion_illegal` and `after_injunction`. The resulting interaction
+term equals 1 when abortion is illegal the time is after the injunction,
+and 0 otherwise. This term is included in the analysis to evaluate the
+joint impact of these factors on infant death rates.
 
@@
 #| message: false
+
 analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
@@
 ```
-
-Another created variable `after_injunction` indicates whether the date
-of infants' death falls after the implementation of the abortion
-injunctions following the overturn of Roe v. Wade in June 2022. It is
-coded as a binary variable, where 1 represents dates after June 2022,
-signifying the post-injunction period, and 0 represents dates before
-June 2022, indicating the pre-injunction period.
-
-To evaluate the impact of abortion injunctions on infant death rates, we
-utilize the Difference-in-Difference (DID) method. As part of this
-approach, we construct an interaction term defined as the product of
-`abortion_illegal` and `after_injunction`. The resulting interaction
-term equals 1 when abortion is illegal the time is after the injunction,
-and 0 otherwise. This term is included in the analysis to evaluate the
-joint impact of these factors on infant death rates.
 
@@
 
-@tbl-summary shows that 50.7% of infant deaths occurred in 2022, slightly higher than in 2021 (49.3%). The majority of cases were among
+@tbl-summary shows the summary statistics of control variables. 50.7% of infant deaths occurred in 2022, slightly higher than in 2021 (49.3%). The majority of cases were among
 mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
@@
 
-The Year of Death variable represents the calendar year in which the
+The Year of Death (`year_of_death`) represents the calendar year in which the
 infant death occurred, specifically focusing on the years 2021 and 2022
@@
 
-The self-identified race of the mother was recorded as a single race
-category. In this dataset, the available categories are: Asian, Black or
-African American, and White. This classification facilitates the
-analysis of maternal and infant health outcomes across these racial
-groups.
+Mother's Single Race (`mothers_single_race`) refers to the self-identified race of the mother, which was recorded as a single race category.
+In this dataset, the available categories are Asian, Black or African American, and White.
+This classification facilitates the analysis of maternal and infant health outcomes across these racial groups.
 
@@
 
+@tbl-motherrace represents the average infant death rates (per 1,000 live births) by maternal race.
+Black or African American mothers have the highest rate at 14.5, compared to 5.8 for White mothers and 5.4 for Asian mothers.
+These differences indicate significant disparities in infant mortality outcomes across racial groups.
+
 ## Random Variable
@@
 
-The state of the mother’s legal residence at the time of birth. This
-data is categorized by the mother’s state of residence and excludes
-territories.
+State (`state`) referes to the state of the mother’s legal residence at the time of birth. This
+data is categorized by the mother’s state of residence and excludes territories.
 
@@
 #| echo: false
+#| fig-pos: "H"
 death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
@@
 Define $y_i$ as as the infant death rate per 1,000 live births, and
-$\gamma_j$ as the random state effect. $$
+$\gamma_j$ as the random state effect for state j. 
+$$
 \begin{aligned}
@@
 \sigma &\sim \mbox{Exponential}(1)
-\end{aligned}$$
-
-The models are run in R (R Core Team 2023) using the `rstanarm` package
-of @rstanarm. The default priors from `rstanarm` are used for both GLM
-Bayesian models. We run the model in R [@citeR] using the `rstanarm`
-package of @rstanarm. We use the default priors from `rstanarm` used for
-both Gaussian Bayesian models.
+\end{aligned}
+$$
+
+We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm` used for both Gaussian Bayesian models.
 
@@
 prior (rate = 1) for variability parameters.
-These priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data.
+By using `autoscale=TRUE`, these priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data.
 This approach
@@
 @tbl-modelresults. The interaction term between after_injunction and
-abortion_illegal is positive and statistically significant (0.111 in
+abortion_illegal is positive and significant (0.111 in
 first model and 0.280 in the second model), indicating a change in infant
@@
 
-The models performed well, with low RMSE values and high $R^2$ values,
-indicating that the predictors account for a substantial portion of the
-variance in infant mortality rates.
+Both models exhibit low RMSE values and high $R^2$, indicating a strong overall fit and that the predictors explain much of the variance in infant mortality rates.
+The second model, with its additional covariates, provides a more detailed and adjusted view of the intervention’s effects, while the first model offers a simpler and more direct perspective.
+The choice between models depends on the balance between interpretability and precision.
 
@@
 #| eval: true
+#| fig-pos: "H"
 
@@
 #| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check" ]
+#| fig-pos: "H"
 
@@
 #| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the first model for infant death rate fits, and is affected by, the data"
+#| fig-pos: "H"
 posterior_vs_prior(first_model) +
@@
 #| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the second model for infant death rate fits, and is affected by, the data"
-#| 
+#| fig-pos: "H"
 posterior_vs_prior(second_model) +
@@
 
-@tbl-priorsummary shows the prior summaries for the two models, `first_model` and `second_model`, indicating that both use weakly
+@tbl-priorsummary shows the prior summaries for the two models, indicating that both use weakly
 informative priors. For the intercept, a normal prior with a mean of 0
@@
 #| label: fig-area
-#| layout-ncol: 1
+#| layout-ncol: 2
 #| fig-cap: "95% credibility intervals for fitted models"
 #| fig-subcap: ["First Model", "Second Model"]
+#| fig-pos: "H"
 
@@
 #| layout-ncol: 1
+#| fig-pos: "H"
 
@@
 #| layout-ncol: 1
+#| fig-pos: "H"
 plot(second_model, "trace")

### 2024-12-02 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with citation error

@@
 @tellingstories, we analyze the impact of abortion injunctions on infant
-death rates using the Difference-in-Difference method. This approach, shown in @sec-model, utilizes the main effects and the interaction between the legality of abortion in the
-mother’s state of residence and the timing of the infant’s death as the predictor variable.
-Additionally, we include control variables, infant year of death, race, and age to account for potential confounding factors, and the mother’s state of residence as a random variable.
+death rates using the Difference-in-Difference method.
+This approach, shown in @sec-model, utilizes the main effects and the interaction between the legality of abortion in the mother’s state of residence and the timing of the infant’s death as the predictor variable.
+Additionally, we include the infants' year of death, race, and age as control variables to account for potential confounding factors, and the mother’s state of residence as a random variable.
 
@@
 Define $y_i$ as as the infant death rate per 1,000 live births, and
-$\gamma_j$ as the random state effect for state j. 
+$\gamma_j$ as the random state effect for state $j$. 
 $$
@@
 intervals, offering an understanding of the range of probable values for
-a parameter @tellingstories. Using Bayesian estimation, we derive a
+a parameter [@tellingstories]. Using Bayesian estimation, we derive a
 posterior distribution for each model coefficient. @fig-area-1 presents
@@
 falls, offering a robust understanding of parameter uncertainty and the
-influence of predictors under the Bayesian framework @tellingstories.
+influence of predictors under the Bayesian framework [@tellingstories].
 
@@
 depression—particularly in unintended or high-risk
-pregnancies—compromising maternal health and prenatal care. Research
+pregnancies—compromising maternal health and prenatal care. Research from
 @crugnola2022 highlights maternal mental health as a key determinant of

### 2024-12-02 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper, pointing how to download raw data
update paper with new reference package

@@
 [@kableExtra] to generate clean tables and reproducible reports, and
-`usmap` for geographical visualizations. We use `here` [@here] package
-simplifies the management of file paths, ensuring a reproducible
+`usmap` for geographical visualizations.
+We use `dplyr` [@dplyr] to manipulate data.
+We use `testthat` [@testthat] to test the simulated and analysis dataset.
+We use `usmap` [@usmap] to plot and analyze U.S.
+state and county-level data on maps by providing preformatted map geometries and easy integration with data.
+We use `here` [@here] package simplifies the management of file paths, ensuring a reproducible
 workflow by creating a consistent point of reference for file locations,
@@
 # Additional data details {#sec-data-details}
+
+## Raw Dataset Download
+
+The complete process of downloading the raw dataset is shown in the `README.md` file.
 

### 2024-12-02 scripts/05-install_packages.R
update packages installed

@@ -8,13 +8,17 @@
 # Any other information needed? Make sure you are in the `Relationship between_infant_mortality_rate_and_prohibited_abortion` rproj
 
 # Install necessary packages
-install.packages("tidyverse")
-install.packages("arrow")
-install.packages("knitr")
-install.packages("usmap")
-install.packages("ggplot2")
-install.packages("dplyr")
-install.packages("kableExtra")
-install.packages("rstanarm")
-install.packages("janitor")
-install.packages("lubridate")
+# List of packages used in the study
+required_packages <- c(
+  "tidyverse", "arrow", "modelsummary", "knitr", "kableExtra", 
+  "usmap", "dplyr", "testthat", "here", "janitor", 
+  "lubridate", "ggplot2", "rstanarm", "fixest", "caret"
+)
+
+# Install any packages not already installed
+installed_packages <- installed.packages()
+for (pkg in required_packages) {
+  if (!(pkg %in% installed_packages[, "Package"])) {
+    install.packages(pkg)
+  }
+}

### 2024-12-04 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update discussion section

@@
 policies and systemic inequities on infant mortality rates.
+
+## Impact of Control Variables on Model Performance and Results
+
+The inclusion of control variables in the second model significantly improves its explanatory power compared to the first model, as evidenced by the higher adjusted $R^2$ (0.778) in the second model versus 0.213 in the first.
+These control variables—such as maternal age, race, and year of death—capture critical demographic and temporal variations influencing infant mortality rates, leading to more precise and reliable estimates of the impact of abortion bans.
+
+For instance, the coefficient for the interaction term (`after_injunction` × `abortion_illegal`) increases from 0.111 in the first model to 0.285 in the second model, suggesting a stronger observed effect of abortion injunctions after adjusting for confounding factors.
+Additionally, the second model highlights disparities, such as the substantial positive effect of 6.604 for Black mothers on infant mortality, underscoring the influence of racial and demographic factors.
+
+The improvement in model fit is further demonstrated by a lower RMSE (2.36 in the second model versus 4.21 in the first) and better predictive performance metrics, including LOOIC and WAIC.
+These results emphasize the importance of controlling for demographic and socioeconomic heterogeneity across states to ensure that the estimated effects of abortion bans are robust and not confounded by underlying differences.
 

### 2024-12-06 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update prose issue

@@
 effectively integrates prior knowledge and addresses uncertainty,
-providing a valuable understanding of policy impacts.
+providing an important understanding of policy impacts.
 
@@
 
-The analysis reveals notable patterns in infant mortality rates
+The analysis explains notable patterns in infant mortality rates
 influenced by policy and demographic factors. For states where abortion
@@
 The inclusion of control variables in the second model significantly improves its explanatory power compared to the first model, as evidenced by the higher adjusted $R^2$ (0.778) in the second model versus 0.213 in the first.
-These control variables—such as maternal age, race, and year of death—capture critical demographic and temporal variations influencing infant mortality rates, leading to more precise and reliable estimates of the impact of abortion bans.
+These control variables—such as maternal age, race, and year of death—capture detailed demographic and temporal variations influencing infant mortality rates, leading to more precise and reliable estimates of the impact of abortion bans.
 
@@
 
-The results reveal stark racial disparities in infant mortality rates,
+The results explain stark racial disparities in infant mortality rates,
 with Black mothers experiencing significantly higher rates compared to
@@
 slightly reduces the error, its added complexity may not justify the
-trade-off if simplicity is a priority. However, if a nuanced
-understanding of demographic influences is essential, the second model
-provides a modest improvement in accuracy while offering more detailed
-insights into maternal and infant health disparities. The
+trade-off if simplicity is a priority. However, if a detailed understanding of demographic influences is essential, the second model
+provides a modest improvement in accuracy while offering a more detailed understanding of maternal and infant health disparities. The
 cross-validation process reinforces the reliability of these models by
@@
 especially those involving circumstances such as abortion or the loss of
-a child. Your understanding are invaluable in identifying areas where
+a child. Your understanding is invaluable in identifying areas where
 healthcare and support services can be improved. If you prefer not to

### 2024-12-06 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
fix gramma error

@@
 state and county-level data on maps by providing preformatted map geometries and easy integration with data.
-We use `here` [@here] package simplifies the management of file paths, ensuring a reproducible
+We use `here` [@here] package to simplify the management of file paths, ensuring a reproducible
 workflow by creating a consistent point of reference for file locations,
@@
 
-State (`state`) referes to the state of the mother’s legal residence at the time of birth. This
+State (`state`) refers to the state of the mother’s legal residence at the time of birth. This
 data is categorized by the mother’s state of residence and excludes territories.
@@
 
-Define $y_i$ as as the infant death rate per 1,000 live births, and
+Define $y_i$ as the infant death rate per 1,000 live births, and
 $\gamma_j$ as the random state effect for state $j$. 

### 2024-12-06 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
fix typo error

@@
 
-One model includes control variables — Age of Mother,
+The first model
+excludes these controls to focus on the overall intervention impact,
+avoid overfitting, and prioritize simplicity where variability is
+adequately accounted for by design or random effects. The second model includes control variables — Age of Mother,
 Mother's Single Race, and Year of Death to adjust for confounders
@@
 compositional differences assumed in the DID framework and ensures
-unbiased estimates of the intervention’s effect. The second model
-excludes these controls to focus on the overall intervention impact,
-avoid overfitting, and prioritize simplicity where variability is
-adequately accounted for by design or random effects. By fitting two
+unbiased estimates of the intervention’s effect. By fitting two
 models, we compare outcomes to refine estimates and evaluate the direct

### 2024-12-07 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update render of paper

@@
 #| label: fig-ppcheck
-#| layout-ncol: 1
+#| layout-ncol: 2
 #| fig-cap: "Posterior Predictive Check: Examining how the model for infant death rate fits, and is affected by, the data"
